# Brazil AISWEB Ingestion

```{r}
#| label: setup
#| include: false

# Libraries, project paths and analysis parameters. `include: false` keeps the
# rendered document free of the package startup banners.
source(here::here("_chapter-setup.R"))
source(here::here("AISWEB", "download_aisweb.R"))
```

This note documents the ingestion of **AISWEB**, the aeronautical information service the
ICA/DECEA publishes: the aerodromes of the ROTAER, the location indicators (GEILOC),
waypoints, routes, NOTAM and charts. It follows the same shape as the other ingestion notes
— get the data, check what arrived, inventory it — with one difference that changes how the
download works.

Every other source in this project is a **feed of movements**, walked day by day. AISWEB is
**reference data**: it describes the things movements happen *at* — which aerodromes exist,
where they are, which runways they have. It changes with the AIRAC amendment, not with the
clock. So there is no day to resume at: an area is fetched whole, its file is replaced, and
the date it was fetched on is written into the file (`AISWEB_FETCHED`), because that is the
only thing that says how old the picture is.

It is here because the movement tables lean on it. `adpartida`, `addestino` and `pista` in
`dstaxi`, and the aerodrome codes of TOTALBR, are only as good as the list they are checked
against, and today that list is `data-raw/airports.csv`, a public database. The ROTAER is
the authority's own.

::: {.callout-important}
## Prerequisites to run this document
The AISWEB files are **not** shipped with the repository, and the API needs a credential — a
key and a pass, sent by e-mail to whoever is granted access. Put both in `.Renviron` with
the chunk below, restart R, then run the download chunk. Until data exists, the tables
below render with a short notice instead of failing.
:::

## The credential, and where it lives

`.Renviron` is the one file a clone cannot bring you: it holds the secrets, so it is
git-ignored and has to be created on each machine. AISWEB adds two lines to it — and unlike
the CGNA, which takes one token, **both** are sent on every call:

| Variable | What it is |
| --- | --- |
| `AISWEB_API_KEY` | the API key |
| `AISWEB_API_PASS` | the API pass — the second half of the same credential |
| `AISWEB_URL` | optional; the endpoint, if it ever moves from `https://api.decea.mil.br/aisweb/` |

```{r}
#| label: setup-bra-aisweb-env
#| eval: false

source(here::here("setup_renviron.R"))

setup_renviron()                 # adds the AISWEB lines; leaves what is there alone
set_token("AISWEB_API_KEY")      # prompts for the key and writes it there
set_token("AISWEB_API_PASS")     # the same, for the pass
env_status()                     # what is set right now; secrets shown only as a length
```

`set_token()` prompts at the console on purpose: a credential typed at a prompt does not
land in `.Rhistory`, in a script, or in a commit. Never write either value into this
document. And R reads `.Renviron` **once, at startup** — restart R after writing them, or
the session will still see them empty.

```{r}
#| label: check-bra-aisweb-env

# is this session able to download? (never prints a credential)
shown <- function(v) {
  if (nzchar(Sys.getenv(v))) paste0("set (", nchar(Sys.getenv(v)), " chars)")
  else "not set in this session"
}
tibble::tibble(
  `.Renviron`       = if (file.exists(here::here(".Renviron"))) "present" else "missing",
  `AISWEB_API_KEY`  = shown("AISWEB_API_KEY"),
  `AISWEB_API_PASS` = shown("AISWEB_API_PASS"),
  `endpoint`        = AISWEB_URL
) |>
  knitr::kable()
```

## Getting the Source Data

There is **one endpoint**, and the area is a parameter of it. Every call is a `GET` on

```
https://api.decea.mil.br/aisweb/?apiKey=…&apiPass=…&area=<area>&…
```

and what comes back is **XML**, not the JSON the ODIN and CGNA APIs return. The areas, as
the [documentation](https://documenter.getpostman.com/view/7201070/SzKQyg3H) lists them:

| `area` | What it returns | Downloaded here |
| --- | --- | --- |
| `rotaer` | the aerodromes of the ROTAER — a paged list, or one aerodrome in full with `icaoCode` | yes |
| `geiloc` | location indicators: aerodromes, heliports, helidecks, TMAs, FIRs, with their status | yes |
| `waypoints` | waypoints, with coordinates and effective date | yes |
| `routesp` | routes | not yet |
| `notam`, `infotemp` | NOTAM and temporary aerodrome information | not yet |
| `met`, `sol` | METAR/TAF, and sunrise/sunset tables | not yet |
| `cartas`, `checklist`, `pub`, `suplementos` | charts and AIP documents | not yet |

### Check the endpoint first

One request, printed in full: the query with the credential redacted, the status **and how
long it took**, the start of the raw body, the columns found and the first row. Run it
before a download, and again whenever one comes back empty.

```{r}
#| label: check-bra-aisweb-endpoint
#| eval: false

aisweb_check("geiloc", name = "cong", type = "ad", feature = "airport")              # one location indicator
# `name` searches the NAME of the place, not its code: "SBGR" there matches nothing
aisweb_check("rotaer", rowstart = 0, rowend = 5)   # the first five aerodromes of the list
aisweb_check("rotaer", icaoCode = "SBGR")          # one aerodrome, in full
aisweb_check("waypoints")
```

It matters more here than elsewhere, because **a refusal is an HTTP 200**. Asked without a
valid credential, the endpoint answers `200` with an HTML page reading *Erro nos parâmetros
obrigatórios* — the status code says nothing. An answer counts only if it parses as XML
rooted at `<aisweb>`; anything else is reported with the text the server sent, and nothing
is written.

### Download

```{r}
#| label: download-bra-aisweb-data
#| eval: false

download_aisweb_rotaer()        # -> data-raw/aisweb/aisweb_rotaer.csv      (every page)
download_aisweb_waypoints()     # -> data-raw/aisweb/aisweb_waypoints.csv   (every page)
download_aisweb_geiloc()        # -> data-raw/aisweb/aisweb_geiloc.csv      (AD, HP and HD)

# filters go through as the API documents them
download_aisweb_rotaer(uf = "SP", type = "AD")
```

The lists are **paged** by `rowstart` and `rowend`, which the documentation does not
explain. Measured against the service: `rowend` is the page *size*, and the wrapper's
`total` is the size of the **whole** result, repeated on every page — the ROTAER answered
`total=6123` thirteen times and delivered 12 × 500 + 123. A download is therefore checked
against the count the API itself reports, and a mismatch is said out loud.

An area can also be paged **without saying so**. `waypoints` is documented as a call with no
parameters, and asked that way it returns exactly 100 rows — a round number, which is what
a default page size looks like and not what a table of waypoints looks like. So the first
page is never taken for the table: every list is walked the same way, advancing by the rows
received, and stopping only on an empty page, on a page that brings nothing new, or when
`total` is reached. A page that fails writes **nothing** — a table short of its last pages
would read as a complete one.

Measured on the first full download: the ROTAER holds **6,123** aerodromes and `waypoints`
**6,785**, both matching the API's own `total`.

`geiloc` is the exception twice over. It is **not paged** — asked for 500 rows it returns
everything, 6,607 aerodromes in one answer, and carries no `total`. And asked with no filter it returns nothing,
so it is asked once per `type` — `ad` aerodrome, `hp` heliport, `hd` helideck, the three the
ROTAER carries — and the answers are written into one file: 8,620 indicators (6,607 `AD`,
1,998 `HP`, 15 `HD`).

### `geiloc` is one row per *version*, not per indicator

Those 8,620 rows are not 8,620 places. Set against the ROTAER's 6,123: every ROTAER code is
in `geiloc`, only 52 `geiloc` codes are missing from the ROTAER — so `geiloc` holds some
6,175 distinct codes, and about 2,400 of its rows repeat a code already there. They are
**versions of the same indicator**: `SBAE` appears eight times, `SBAC` five, each with an
`id` of its own, and across `SBAC`'s five the name changes from *Regional de Aracati* to
*Aeroporto Regional de Canoa Quebrada Dragão do Mar*.

Two things were checked and neither picks the current one out. Every row is `Ativo`, so
`status` does not separate them. And although the ROTAER's `id` is documented as a version
identifier, it is not the same identifier: none of `SBAC`'s five `geiloc` ids exists in the
ROTAER. `geiloc` carries no date either.

So the file is kept as the API returns it, one row per version, and it is **not**
deduplicated by code — that would keep one version at random and present it as the current
one. For "which aerodromes exist, and what are they called today", read the **ROTAER**;
`geiloc` is the record of the names an indicator has had.

### The aerodrome in full, and its runways

The list carries the identity of an aerodrome. The detail — coordinates, elevation, FIR,
operator, and the **runways** — is one call per aerodrome, and it comes out as two files
because it is two grains:

```{r}
#| label: download-bra-aisweb-aerodromes
#| eval: false

# the study airports; any vector of ICAO codes works
download_aisweb_aerodromes(c("SBGR", "SBSP", "SBGL", "SBRJ", "SBBR", "SBCF",
                             "SBKP", "SBPA", "SBCT", "SBSV", "SBRF", "SBFZ"))
# -> data-raw/aisweb/aisweb_aerodromes.csv   one row per aerodrome
# -> data-raw/aisweb/aisweb_runways.csv      one row per runway, keyed on AeroCode
```

## Checking What Arrived

```{r}
#| label: check-bra-aisweb-data

aisweb_files <- list.files(AISWEB_OUT_DIR, pattern = "^aisweb_.*\\.csv$", full.names = TRUE)

if (length(aisweb_files) == 0) {
  cat("No AISWEB file in data-raw/aisweb/ yet. Set the credential and run the",
      "download chunk above.")
} else {
  purrr::map_dfr(aisweb_files, function(f) {
    d <- cgna_read_csv(f)
    tibble::tibble(
      FILE    = basename(f),
      ROWS    = nrow(d),
      COLUMNS = ncol(d),
      FETCHED = if ("AISWEB_FETCHED" %in% names(d)) max(d$AISWEB_FETCHED, na.rm = TRUE) else NA,
      SIZE_MB = round(file.size(f) / 1024^2, 2)
    )
  }) |>
    knitr::kable()
}
```

## What the Columns Mean

The names are the API's own, kept as they arrive.

**`aisweb_rotaer.csv`** — one row per aerodrome of the list.

| Column | Meaning |
| --- | --- |
| `ciad_id` | numeric identifier of the aerodrome in the AISWEB database |
| `id` | identifier of this *version* of the aerodrome — each update creates a new one |
| `type` | `AD` aerodrome, or heliport / helideck |
| `AeroCode` | the ICAO location indicator |
| `ciad` | the aerodrome's business identifier (`SP0831`) |
| `name`, `city`, `uf` | name, municipality, state |
| `lat`, `lng` | coordinates, decimal degrees — returned by the service, though not documented |
| `dt` | date the record was published or updated |

**`aisweb_geiloc.csv`** — `id`, `type`, `IcaoCode`, `name`, `status`, `city`, `uf`. One row
per *version* of an indicator, so `IcaoCode` repeats; see above.

**`aisweb_waypoints.csv`** — `id`, `ident`, `latitude`, `longitude` (decimal), `dt` (date the
waypoint came into effect).

**`aisweb_aerodromes.csv`** — the list's fields plus `status`, `lat`, `lng` (decimal),
`latRotaer`, `lngRotaer` (as printed in the ROTAER), `altM`, `altFt`, `fir`, `jur`, `utc`,
`typeUtil` (`PUB`/`PRIV`), `cat`, the operator (`org_name`, `org_type`, `org_military`) and
`runways_count`.

**`aisweb_runways.csv`** — `AeroCode`, `ident` (`17R/35L`), `type`, `surface`, `length` and
`width` in metres, `surface_c` (the PCN string) and `thr`, the two thresholds.

Every file also carries `AISWEB_FETCHED`, the date of the download.
