# Brazil ASMA (KPI08) Ingestion

```{r}
#| label: setup
#| message: false
#| warning: false

# Libraries, project paths and analysis parameters — including every ASMA
# setting (asma_data_years, asma_ref_key, asma_max_asma, the directories and the
# output file names). Change the study period or the reference there, not here.
source(here::here("_chapter-setup.R"))
```

This note documents the preparation of the Brazilian **ASMA (KPI08 — additional time in
terminal airspace)** files used for the 2025 Brazil-Europe operational efficiency update.
It is the arrival-side companion to `Taxi-BRA-ingestion.qmd` and follows the same shape:
a source inventory, an APDF-like harmonisation, a reproducible preparation chunk, and a
coverage check.

The source `kpi08` files are ARRIVAL ASMA extracts. They already carry a supplied
`kpi08`/`transito`/`desimp`; the pipeline recomputes the GANP 20th-percentile reference
in-repo (so the private `PBWG` package is not required) and keeps the supplied values
alongside for validation.

::: {.callout-important}
## Prerequisites to run this document
The raw `kpi08` CSV files are **not** shipped with the repository. Get them first with
the download chunk below (ODIN API), or place your own `kpi08_*.csv` in
`data-raw/kpi08/` (or set `BRA_ASMA_ZIP` to a zip archive containing them). Then run the
`prepare-bra-asma-data` chunk once to generate the analytic outputs. Until then, the
tables below render with a short notice instead of failing. See `README.md` for setup.
:::

## Getting the Source Data

The `kpi08` source comes from the ICEA/DECEA **ODIN** API, one file per year in
`data-raw/kpi08/`. The chunk below is `eval: false` — the download is slow, so it is not
executed on render; run it manually when you need data.

It is **incremental**: a year already on disk is resumed from the latest record it holds,
so re-running it daily or monthly only fetches what is new. A year with no file yet is
downloaded in full.

```{r}
#| label: download-bra-kpi08-data
#| eval: false

# sourcing only defines download_kpi08(); nothing is fetched until it is called
source(here::here("KPI08", "download_kpi08.R"))

# the full study period; drop to the current year for the routine refresh
download_kpi08(asma_data_years, out_dir = bra_asma_raw_dir)
```

```{r}
#| label: check-bra-kpi08-data

# what is on disk right now, and what to do if nothing is
have_files <- list.files(bra_asma_raw_dir, pattern = "^kpi08_.*\\.csv$")

if (length(have_files) == 0) {
  knitr::kable(tibble::tibble(
    status = "no source data yet",
    `what to do` = paste0(
      "Run the `download-bra-kpi08-data` chunk above (or, from a terminal, ",
      "`Rscript KPI08/download_kpi08.R ",
      min(asma_data_years), " … ", max(asma_data_years), "`)."
    )
  ))
} else {
  knitr::kable(tibble::tibble(
    status = paste0(length(have_files), " source file(s) in ", bra_asma_raw_dir),
    years  = paste(sort(unique(stringr::str_extract(have_files, "20[0-9]{2}"))),
                   collapse = ", ")
  ))
}
```

## Source Inventory

```{r}
#| label: tbl-bra-asma-source-inventory
#| tbl-cap: "Inventory of the raw ASMA source files (kpi08)"

# Accepts annual files (kpi08_YYYY.csv) and monthly files (kpi08_mmmYYYY.csv).
files <- list.files(
  bra_asma_raw_dir,
  pattern = "^kpi08_.*\\.csv$",
  full.names = TRUE
)

inventory <- purrr::map(files, function(file) {
  member <- basename(file)
  year   <- stringr::str_extract(member, "20[0-9]{2}")
  month  <- stringr::str_match(member, "kpi08_([a-z]{3})20[0-9]{2}\\.csv$")[, 2]

  tibble::tibble(
    File      = member,
    Year      = year,
    Month     = ifelse(is.na(month), "annual", month),
    `Size (MB)` = round(file.info(file)$size / 1024^2, 1)
  )
}) |>
  purrr::list_rbind()

if (nrow(inventory) == 0) {
  knitr::kable(tibble::tibble(
    note = paste0("No source files in ", bra_asma_raw_dir,
                  " yet. Run the `download-bra-kpi08-data` chunk above to fetch them ",
                  "from the ODIN API (or place your own kpi08_*.csv there, or set ",
                  "BRA_ASMA_ZIP).")
  ))
} else {
  inventory |>
    dplyr::arrange(Year, Month, File) |>
    knitr::kable()
}
```

The annual archive contains one file per year (`kpi08_2023.csv`, `kpi08_2024.csv`,
`kpi08_2025.csv`, semicolon-delimited). The monthly archive contains 2025 monthly files
(comma-delimited, with a leading row index). The monthly set has been seen to contain
**both** `kpi08_feb2025.csv` and `kpi08_fev2025.csv` — a likely English/Portuguese
month-name duplicate that must be resolved before combining months.

## Harmonisation

The `kpi08` payload columns are mapped to the APDF-like ARR fields required by the ASMA
metric. For arrivals, the study airport is the destination aerodrome (`ades`).

| Source field | Harmonised field | Note |
| --- | --- | --- |
| `fltid` | `FLTID` | Flight identifier |
| `adep` | `ADEP` | Departure aerodrome |
| `ades` | `ADES` / `ICAO` | Destination aerodrome (the study airport for ARR) |
| `reg` | `REG` | Aircraft registration (often blank) |
| `type` | `ARCTYP` | Aircraft type |
| `CLASS` | `CLASS` | Brazil/WTC-style weight class, kept raw as supplied |
| `drwy_validado` / `drwy` | `RWY` | Validated runway when present, else raw runway |
| `stnd` | `STND` | Stand/gate |
| `aldt` | `ALDT` / `MVT_TIME` | Landing time (the arrival movement time) |
| `aibt` | `AIBT` | In-block time (blank where unmatched) |
| `sibt` | `SIBT` | Scheduled in-block time (often blank) |
| `c` | `RANGE` | **ASMA ring in NM (40 / 100)** → `C40` / `C100`; one row per ring |
| `c_time` | `C_CROSS_TIME` | Crossing time at that ring |
| `bear` | `BEARING` | Bearing at the crossing point |
| `setor` | `SECTOR` | Entry sector |
| `transito` | `ASMA_TRANSIT` | Supplied ASMA transit time (duration) |
| `desimp` | `ASMA_UNIMPEDED` | Supplied unimpeded time (duration) |
| `kpi08` | `ASMA_ADD_SUPPLIED` | Supplied additional time (duration) — kept for validation |
| `id` | `IM_SAMAD_ID` | Source movement identifier |

Two source conventions differ between the archives and are normalised on read: the
monthly files carry a leading `row_index` column and use a comma delimiter, while the
annual files are semicolon-delimited.

**Each arrival appears once per ASMA ring**, identified by `c` (40 or 100 NM), with
`c_time`/`bear` giving that ring's crossing. The two are roughly half the rows each and
about 10 minutes apart, so they are carried as separate `RANGE` values (`C40`, `C100`) and
never pooled: the reference is computed per ring and the dashboard switches between them.

## Preparation Code

The following chunk is the reproducible preparation script. It is not evaluated during
normal rendering because it reads the large source archive.

The ASMA metric is computed in-chunk with base tidyverse (no PBWG dependency). Every
machine-specific location is read from an environment variable with a project-relative
fallback:

- `BRA_ASMA_ZIP` — path to the source zip archive (optional; when unset, the `kpi08_*.csv`
  files in `data-raw/kpi08/` are used).
- `BRA_REPORT_DATA` — data directory of the sibling report project that receives the
  combined analytic CSVs (defaults to this project's `data/`).

```{r}
#| label: prepare-bra-asma-data
#| eval: false

source(here::here("_chapter-setup.R"))

# ---- source/report locations (override with environment variables) ----------
zip_path        <- Sys.getenv("BRA_ASMA_ZIP", unset = "")
report_data_dir <- Sys.getenv("BRA_REPORT_DATA", unset = bra_report_data)

dir_create(bra_asma_apdf_dir)

# ---- small helpers ----------------------------------------------------------
normalise_text <- function(x) {
  x <- as.character(x); x <- trimws(x); x[x == ""] <- NA_character_; x
}

# ASMA timestamps (landing/in-block etc.) are date-times; parse tolerantly.
parse_asma_datetime <- function(x) {
  x <- normalise_text(x)
  parse_date_time(x, orders = c("ymd HMS", "ymd HMS OS", "ymd HM", "dmy HMS"),
                  tz = "UTC", quiet = TRUE)
}

# Durations come as "H:MM:SS(.s)"; return minutes (supports a leading '-').
parse_duration_minutes <- function(x) {
  x   <- normalise_text(x)
  sgn <- ifelse(!is.na(x) & grepl("^-", x), -1, 1)   # capture sign before stripping
  x   <- sub("^-", "", x)
  p   <- stringr::str_match(x, "^([0-9]+):([0-9]{2}):([0-9]{2}(?:\\.[0-9]+)?)$")
  suppressWarnings(
    sgn * (as.numeric(p[, 2]) * 60 + as.numeric(p[, 3]) + as.numeric(p[, 4]) / 60)
  )
}

# Entry sector as the official report codes it: SIX 60-degree sectors labelled
# S000, S060, ... S300 (NOT an 8-point compass — that binning creates smaller
# reference groups, pushes them under min_n and inflates the "no reference"
# count). `setor` is used when it already carries those codes; otherwise the
# sector is binned from the bearing.
sector_group <- function(setor, bear) {
  setor <- normalise_text(setor)
  deg   <- suppressWarnings(as.numeric(normalise_text(bear))) %% 360
  binned <- ifelse(is.na(deg), NA_character_,
                   sprintf("S%03d", (floor(deg / 60) * 60) %% 360))
  ifelse(!is.na(setor) & grepl("^S[0-9]{3}$", setor), setor, binned)
}

# ---- read one kpi08 file (annual ; or monthly ,) ----------------------------
read_bra_asma_file <- function(csv_path, member = basename(csv_path)) {
  is_monthly <- grepl("kpi08_[a-z]{3}20[0-9]{2}\\.csv$", member)
  sep <- if (is_monthly) "," else ";"

  if (file.exists(csv_path)) {
    message("Reading ", csv_path)
    raw <- fread(csv_path, sep = sep, colClasses = "character", showProgress = FALSE)
  } else if (nzchar(zip_path) && file.exists(zip_path)) {
    message("Reading ", member, " from ", basename(zip_path))
    raw <- fread(cmd = sprintf("unzip -p %s %s", shQuote(zip_path), shQuote(member)),
                 sep = sep, colClasses = "character", showProgress = FALSE)
  } else {
    stop("No source found for ", member, ": expected ", csv_path,
         " or a zip archive via BRA_ASMA_ZIP.")
  }

  raw <- as_tibble(raw)
  # drop the monthly leading row index if present
  if ("V1" %in% names(raw)) raw$V1 <- NULL
  names(raw) <- gsub('^"|"$', "", names(raw))
  raw
}

# ---- harmonise to APDF-like ARR ASMA ---------------------------------------
harmonise_bra_asma <- function(raw) {
  # safe column accessor: NA vector of the right length when a column is absent
  gcol <- function(name) if (name %in% names(raw)) raw[[name]] else rep(NA_character_, nrow(raw))

  rwy    <- dplyr::coalesce(normalise_text(gcol("drwy_validado")),
                            normalise_text(gcol("drwy")))
  sector <- sector_group(gcol("setor"), gcol("bear"))

  tibble::tibble(
    IM_SAMAD_ID = normalise_text(gcol("id")),
    FLTID  = normalise_text(gcol("fltid")),
    ADEP   = normalise_text(gcol("adep")),
    ADES   = normalise_text(gcol("ades")),
    REG    = normalise_text(gcol("reg")),
    ARCTYP = normalise_text(gcol("type")),
    CLASS  = normalise_text(gcol("CLASS")),
    RWY    = rwy,
    STND   = normalise_text(gcol("stnd")),
    ALDT   = parse_asma_datetime(gcol("aldt")),
    AIBT   = parse_asma_datetime(gcol("aibt")),
    SIBT   = parse_asma_datetime(gcol("sibt")),
    C_CROSS_TIME = parse_asma_datetime(gcol("c_time")),
    BEARING      = suppressWarnings(as.numeric(gcol("bear"))),
    SECTOR       = normalise_text(gcol("setor")),
    ASMA_TRANSIT      = parse_duration_minutes(gcol("transito")),
    ASMA_UNIMPEDED    = parse_duration_minutes(gcol("desimp")),
    ASMA_ADD_SUPPLIED = parse_duration_minutes(gcol("kpi08")),
    # ARR-oriented data: phase is arrival, study airport is the destination
    PHASE = "ARR",
    ICAO  = normalise_text(gcol("ades")),
    # The ring the row was measured at (40 or 100 NM). Each arrival appears once
    # per ring and the two are ~10 minutes apart, so this must come from the data
    # — labelling every row with one ring mixes the two distributions and makes
    # the p20 reference meaningless.
    RANGE = paste0("C", normalise_text(gcol(asma_ring_col))),
    SECTOR_GROUP = sector
  ) |>
    dplyr::filter(ICAO %in% bra_apts, RANGE %in% asma_rings)
}

# ---- observed ASMA transit + validity --------------------------------------
# Prefer the transit derived from the timestamps (landing - crossing); fall back
# to the supplied `transito` when a timestamp is missing.
prepare_asma <- function(df) {
  df |>
    mutate(
      ASMA_TRANSIT_CALC = as.numeric(difftime(ALDT, C_CROSS_TIME, units = "mins")),
      ASMA = dplyr::coalesce(ASMA_TRANSIT_CALC, ASMA_TRANSIT),
      DATE = as_date(ALDT),
      # the plausibility cap is per ring: 100 NM legitimately takes much longer
      MAX_ASMA = unname(asma_max_asma[RANGE]),
      VALID_ASMA = !is.na(ASMA) & ASMA > 0 & !is.na(MAX_ASMA) & ASMA <= MAX_ASMA
    )
}

# ---- 2024 GANP p20 reference (unimpeded) -----------------------------------
build_asma_reference <- function(mov_ref) {
  ref_start <- ymd_hms(paste0(asma_ref_year, "-01-01 00:00:00"), tz = "UTC")
  ref_end   <- ymd_hms(paste0(asma_ref_year, "-12-31 23:59:59"), tz = "UTC")

  mov_ref |>
    filter(VALID_ASMA,
           if_all(all_of(asma_ref_key), ~ !is.na(.x)),
           ALDT >= ref_start, ALDT <= ref_end) |>
    group_by(across(all_of(asma_ref_key))) |>
    summarise(N = n(),
              ASMA_REF = quantile(ASMA, asma_p_ref, type = 7, names = FALSE),
              .groups = "drop") |>
    filter(N >= asma_min_n) |>
    select(all_of(asma_ref_key), ASMA_REF)
}

# ---- daily analytic table ---------------------------------------------------
summarise_asma_daily <- function(mov, reference) {
  mov |>
    filter(!is.na(DATE)) |>
    left_join(reference, by = asma_ref_key) |>
    mutate(VALID = VALID_ASMA & !is.na(ASMA_REF)) |>
    # Grouped exactly as the official report: the reference breakdown is kept in
    # the daily table, so results can be validated group by group against golden/.
    group_by(ICAO, PHASE, DATE, RANGE, CLASS, RWY, SECTOR_GROUP) |>
    summarise(
      MVTS_VALID    = sum(VALID),
      MVTS_NA       = sum(!VALID),
      TOT_ASMA      = sum(ASMA[VALID]),
      TOT_REF       = sum(ASMA_REF[VALID]),
      TOT_ADD_TIME  = sum(ASMA[VALID] - ASMA_REF[VALID]),
      # the Brazil-supplied unimpeded and KPI08 totals, kept for validation
      TOT_BRA_REF   = sum(ASMA_UNIMPEDED[VALID], na.rm = TRUE),
      TOT_BRA_KPI08 = sum(ASMA_ADD_SUPPLIED[VALID], na.rm = TRUE),
      .groups = "drop"
    ) |>
    arrange(ICAO, PHASE, DATE, RANGE, CLASS, RWY, SECTOR_GROUP)
}

# ---- run the pipeline -------------------------------------------------------
# Resolve the source files: annual files are preferred; monthly files are used
# only to fill years the annual set does not cover.
raw_files <- list.files(bra_asma_raw_dir, pattern = "^kpi08_.*\\.csv$", full.names = TRUE)
if (length(raw_files) == 0 && !nzchar(zip_path))
  stop("No ASMA source found. Put kpi08_*.csv in ", bra_asma_raw_dir,
       " or set BRA_ASMA_ZIP.")

harmonised <- purrr::map(raw_files, function(f) harmonise_bra_asma(read_bra_asma_file(f))) |>
  purrr::list_rbind() |>
  mutate(YEAR = lubridate::year(ALDT)) |>
  filter(YEAR %in% asma_data_years) |>
  # De-duplicate on id AND ring: every arrival appears once per ASMA ring, and
  # if the source shares one id between those two rows, keying on the id alone
  # silently deletes a whole ring.
  distinct(IM_SAMAD_ID, RANGE, .keep_all = TRUE)

# write one harmonised parquet extract per year
walk(asma_data_years, function(y) {
  ext <- filter(harmonised, YEAR == y)
  if (nrow(ext) > 0)
    write_parquet(ext, path(bra_asma_apdf_dir, paste0("PBWG-BRA-asma-apdf-", y, ".parquet")))
})

movements <- prepare_asma(harmonised)
reference_data <- build_asma_reference(filter(movements, YEAR == asma_ref_year))

daily <- summarise_asma_daily(movements, reference_data)
write_csv(daily, path(report_data_dir, bra_asma_analytic_file))

# ---- coverage summaries -----------------------------------------------------
coverage_summary <- movements |>
  group_by(ICAO, PHASE, YEAR) |>
  summarise(
    N_TOTAL          = n(),
    N_VALID_ASMA     = sum(VALID_ASMA),
    N_MISSING_RANGE  = sum(is.na(RANGE)),
    N_MISSING_CLASS  = sum(is.na(CLASS)),
    N_MISSING_RWY    = sum(is.na(RWY)),
    N_MISSING_SECTOR = sum(is.na(SECTOR_GROUP)),
    .groups = "drop"
  ) |>
  mutate(PCT_VALID_ASMA = N_VALID_ASMA / N_TOTAL) |>
  arrange(ICAO, PHASE, YEAR)
write_csv(coverage_summary, path(report_data_dir, bra_asma_coverage_file))

reference_coverage <- reference_data |>
  group_by(RANGE) |>
  summarise(
    N_REFERENCE_GROUPS       = n(),
    N_VALID_REFERENCE_GROUPS = sum(!is.na(ASMA_REF)),
    .groups = "drop"
  )
write_csv(reference_coverage, path(report_data_dir, bra_asma_reference_coverage_file))
```

## Retained Artefacts

```{r}
#| label: tbl-bra-asma-retained-artefacts

parquet_files <- list.files(bra_asma_apdf_dir, pattern = "\\.parquet$", full.names = TRUE)

if (length(parquet_files) == 0) {
  knitr::kable(tibble::tibble(
    note = "No ASMA parquet extracts yet. Run the `prepare-bra-asma-data` chunk."
  ))
} else {
  tibble::tibble(path = parquet_files) |>
    dplyr::mutate(file = basename(path),
                  size_mb = round(file.info(path)$size / 1024^2, 1)) |>
    dplyr::select(file, size_mb) |>
    knitr::kable()
}
```

## Comparability Check — is the reference transferable?

The reference is built from `asma_ref_year` and applied to every other year, **within each
ring**. Two things can break that, and both have bitten this pipeline:

1. **Mixing the rings.** The source has one row per ring per arrival, ~50/50, and they are
   about 10 minutes apart (measured 2026: C40 median 13.8 min, C100 23.7 min). Labelling
   every row with a single ring pooled the two distributions, so the "reference" tracked
   the ring mix rather than the operation — a year with proportionally more C100 looked
   slower, and 2026 came out with a *negative* average additional time while the supplied
   `kpi08` stayed positive. `RANGE` now comes from the source column, never a constant.
2. **Mixing sources.** Years pulled with different download code can still cover different
   flights.

The table below is the guard. It reports the transit level per year **and ring**; every
year should sit close to the reference year within its own ring.

```{r}
#| label: tbl-bra-asma-comparability

parq <- list.files(bra_asma_apdf_dir, pattern = "\\.parquet$", full.names = TRUE)

if (length(parq) == 0) {
  knitr::kable(tibble::tibble(
    note = "No ASMA extracts yet. Run the `prepare-bra-asma-data` chunk."
  ))
} else {
  # per YEAR *and* RANGE: the two rings are ~10 minutes apart, so a table that
  # pooled them would compare the ring mix rather than the operation
  lvl <- purrr::map(parq, function(f) {
    arrow::read_parquet(f) |>
      dplyr::mutate(
        YEAR = lubridate::year(ALDT),
        TRANSIT = as.numeric(difftime(ALDT, C_CROSS_TIME, units = "mins"))
      ) |>
      dplyr::group_by(YEAR, RANGE) |>
      dplyr::summarise(
        N = dplyr::n(),
        MED_TRANSIT = round(median(TRANSIT, na.rm = TRUE), 2),
        MED_SUPPLIED_ADD = round(median(ASMA_ADD_SUPPLIED, na.rm = TRUE), 2),
        NO_CROSSING_PCT = round(100 * mean(is.na(C_CROSS_TIME)), 1),
        .groups = "drop"
      )
  }) |> purrr::list_rbind() |> dplyr::arrange(RANGE, YEAR)

  # 1 minute is well inside year-to-year operational drift; beyond that the years
  # are not on the same ruler and the shared reference cannot be trusted
  tol <- 1.0
  ref <- lvl |>
    dplyr::filter(YEAR == asma_ref_year) |>
    dplyr::select(RANGE, REF_LEVEL = MED_TRANSIT)

  lvl |>
    dplyr::left_join(ref, by = "RANGE") |>
    dplyr::mutate(
      VS_REF = round(MED_TRANSIT - REF_LEVEL, 2),
      COMPARABLE = dplyr::case_when(
        is.na(VS_REF)      ~ paste0("reference year ", asma_ref_year, " missing"),
        abs(VS_REF) <= tol ~ "yes",
        TRUE               ~ "NO — check the source of this year"
      )
    ) |>
    dplyr::select(-REF_LEVEL) |>
    knitr::kable()
}
```

If any year reads **NO**, do not trust its additional time. Re-download that year so every
year comes from the same source, then re-run the preparation:

```r
source(here::here("KPI08", "download_kpi08.R"))
download_kpi08(<the offending years>, out_dir = bra_asma_raw_dir, force = TRUE)
```

## Coverage Check

The pipeline records, per year and phase/range, the number of valid ASMA movements
(`MVTS_VALID`), the movements that could not be used (`MVTS_NA`), and the observed ASMA
total (`TOT_ASMA`), reference total (`TOT_REF`) and additional-time total
(`TOT_ADD_TIME`). The supplied `transito`/`desimp`/`kpi08` totals are kept alongside for
validation against the in-repo reference.

```{r}
#| label: tbl-bra-asma-coverage

asma_daily_path <- here::here("data", bra_asma_analytic_file)

if (file.exists(asma_daily_path)) {
  readr::read_csv(asma_daily_path, show_col_types = FALSE) |>
    dplyr::mutate(YEAR = lubridate::year(DATE)) |>
    dplyr::group_by(YEAR, PHASE, RANGE) |>
    dplyr::summarise(
      MVTS_VALID   = sum(MVTS_VALID),
      MVTS_NA      = sum(MVTS_NA),
      AVG_ASMA     = sum(TOT_ASMA) / sum(MVTS_VALID),
      AVG_ADD_TIME = sum(TOT_ADD_TIME) / sum(MVTS_VALID),
      AVG_ADD_SUPPLIED = sum(TOT_ADD_SUPPLIED) / sum(MVTS_VALID),
      .groups = "drop"
    ) |>
    dplyr::mutate(dplyr::across(c(AVG_ASMA, AVG_ADD_TIME, AVG_ADD_SUPPLIED),
                                ~ round(.x, 3))) |>
    knitr::kable()
} else {
  knitr::kable(tibble::tibble(
    note = paste0("ASMA analytic output not found. Run the `prepare-bra-asma-data` ",
                  "chunk to generate ", bra_asma_analytic_file, " in data/.")
  ))
}
```

The `AVG_ADD_TIME` (in-repo GANP p20 reference) and `AVG_ADD_SUPPLIED` (Brazil-supplied
`kpi08`) columns are directly comparable: a close match validates that the in-repo
reference reproduces the supplied KPI08 for the populated rows.

## Open Points to Confirm

Before treating the output as final, confirm with DECEA/ICEA:

1. ~~What `c_time` / `bear` represent~~ — **resolved.** The `c` column holds the ring in
   nautical miles (40 or 100) and each arrival appears once per ring; `c_time` is that
   ring's crossing time. `RANGE` is derived from it and the two rings are kept apart
   throughout.
2. Whether `transito`, `desimp` and `kpi08` definitions match the KPI08 expectations used
   for the reference.
3. Whether to use the annual 2025 file or reconstruct 2025 from the monthly files, after
   resolving the `feb`/`fev` duplicate.
4. Whether any DEP/DSMA (departure) evidence exists; these files are arrival-only, so a
   separate DSMA source is still needed for the departure side.

## Dashboard

`kpi08.html` is the interactive companion to this document: headline tiles, additional
time by year, an airport ranking, the monthly trend, and a validation panel putting the
reference recomputed here next to the `kpi08` value supplied by Brazil.

The page carries its data embedded, so it opens offline by double-clicking. Run the chunk
below whenever the analytic CSV changes to refresh those embedded numbers. It is
`eval: false` so rendering this document never rewrites the page by surprise.

```{r}
#| label: build-bra-kpi08-dashboard
#| eval: false

# sourcing only defines build_kpi08_dashboard(); nothing is written until it is called
source(here::here("KPI08", "build_kpi08_dashboard.R"))

build_kpi08_dashboard(ring = asma_ring)
```

```{r}
#| label: check-bra-kpi08-dashboard

# The page is tracked in git; its numbers live in the generated, git-ignored
# kpi08-data.js next to it. Report whether that data file has been built.
dash_file <- here::here("kpi08.html")
data_file <- here::here("kpi08-data.js")
dash_state <- if (!file.exists(dash_file)) {
  "kpi08.html not found"
} else if (!file.exists(data_file)) {
  "no data yet — run the `build-bra-kpi08-dashboard` chunk above"
} else {
  paste0("built ", format(file.info(data_file)$mtime, "%Y-%m-%d %H:%M"),
         " — open kpi08.html to view the dashboard")
}
knitr::kable(tibble::tibble(dashboard = basename(dash_file), status = dash_state))
```
