# Brazil TATIC Data Ingestion

```{r}
#| label: setup
#| include: false

# Libraries, project paths and analysis parameters. `include: false` keeps the
# rendered document free of the package startup banners and "built under R
# version" warnings that loading the tidyverse emits.
source(here::here("_chapter-setup.R"))
```

This note documents the ingestion of **TATIC**, the CGNA/DECEA movement feed, for the
Brazil–Europe operational-efficiency study.

TATIC is not another copy of the taxi source. It carries the **milestones of a movement** —
EOBT, push-back request and clearance, taxi request and start, holding, runway entry,
departure and arrival — where `dstaxi` carries only the movement time and the block time.
What `dstaxi` gives as a single taxi duration, TATIC can decompose into waiting for push,
taxiing, and holding short of the runway.

::: {.callout-important}
## Prerequisites to run this document
The raw TATIC files are **not** shipped with the repository, and the API needs a token.
Create `.Renviron` with the chunk below, put the token in it, restart R, then run the
`download-bra-tatic-data` chunk. Until data exists, the tables below render with a short
notice instead of failing.
:::

## The token, and where it lives

`.Renviron` is the one file a clone cannot bring you: it holds the token, so it is
git-ignored and has to be created on each machine. This writes it, with every variable this
repository reads and a comment explaining each:

```{r}
#| label: setup-bra-tatic-env
#| eval: false

source(here::here("setup_renviron.R"))

setup_renviron()   # creates/updates .Renviron - only adds what is missing
set_token()        # prompts for the TATIC token and writes it there
env_status()       # what is set right now; tokens shown only as a character count
```

`set_token()` prompts at the console on purpose: a token typed at a prompt does not land in
`.Rhistory`, in a script, or in a commit — which is exactly what happens when it is assigned
in code. Never write the token into this document.

Two things that catch people out. On macOS and Linux a file starting with a dot is **hidden**
in Finder and in the RStudio Files pane, so `.Renviron` is there without being visible —
check with `file.exists(renviron_path())` rather than by looking. And R reads `.Renviron`
**once, at startup**: restart R after writing the token, or `Sys.getenv("TATIC_TOKEN")` will
still be empty.

```{r}
#| label: check-bra-tatic-env

# is this session able to download? (never prints the token itself)
tibble::tibble(
  `.Renviron`    = if (file.exists(here::here(".Renviron"))) "present" else "missing",
  `TATIC_TOKEN`  = if (nzchar(Sys.getenv("TATIC_TOKEN")))
                     paste0("set (", nchar(Sys.getenv("TATIC_TOKEN")), " chars)")
                   else "not set in this session",
  `proxy`        = if (nzchar(Sys.getenv("BRA_PROXY", unset = Sys.getenv("ODIN_PROXY"))))
                     "configured" else "none (fine outside a corporate network)"
) |>
  knitr::kable()
```

## Getting the Source Data

The source is the CGNA API at `https://portal.cgna.decea.mil.br/apiv1/tatic`, and the files
land in the same shape as every other dataset here:

```
data-raw/tatic/parts/tatic_2025-01.csv   one file per MONTH
data-raw/tatic/tatic_2025.csv            the months merged into one YEAR
```

### One day per call

This API is not the ODIN API, and the difference decides the whole design: **a wide window
is ignored**. Asking for `datai=20250101&dataf=20250130` returns *only the first day*, with
no error and no warning — the sort of thing that produces a year file with 12 days in it and
no sign anything went wrong.

So the period is walked one day at a time (`datai=d`, `dataf=d+1`), and the days are
accumulated into the month file. **The day is the unit of fetching; the month is the unit of
storage.**

### What resumability means here

Each day is written into its month file as it arrives, and a re-run reads the days already
stored and asks only for what is missing. An interruption costs at most one day, never a
month — which matters, because a single day is 6–7 MB and the API is slow.

The day is recorded in an added column, **`TATIC_DAY`**: the day *we asked for*. It is not
inferred from the record's own date fields, because those describe the **event**, and a
movement can be reported on a day other than the one it is filed under. Resuming on a field
that means something else is how gaps appear.

A day the source has **no records** for is stored as a placeholder row carrying only
`TATIC_DAY`. That is deliberate: without it, an empty day would be requested again on every
single run, forever. The placeholders are dropped when the months are merged into the year,
so they never reach the analysis.

### Requests are serial

The API queues concurrent requests until they time out — with three at once, one day
completed and the other two returned a few bytes before the timeout. `TATIC_CONCURRENCY`
exists, but raising it usually costs time rather than saving it.

```{r}
#| label: download-bra-tatic-data
#| eval: false

source(here::here("API_TATIC", "download_tatic.R"))

# a whole year, resuming whatever is already on disk
download_tatic(tatic_years)

# or a slice, for a first test — one day is enough to see what comes back
download_tatic(2025, from = "20250101", to = "20250101")
```

```{r}
#| label: check-bra-tatic-data

# what is on disk right now, and what to do if nothing is
year_files <- list.files(bra_tatic_raw_dir, pattern = "^tatic_\\d{4}\\.csv$",
                         full.names = TRUE)
part_files <- list.files(file.path(bra_tatic_raw_dir, "parts"),
                         pattern = "^tatic_\\d{4}-\\d{2}\\.csv$", full.names = TRUE)

status <- if (length(year_files) > 0 || length(part_files) > 0) {
  tibble::tibble(
    source = bra_tatic_raw_dir,
    years  = paste(sort(stringr::str_extract(basename(year_files), "\\d{4}")),
                   collapse = ", "),
    months = length(part_files)
  )
} else {
  tibble::tibble(
    source = "no source data yet",
    years  = "set TATIC_TOKEN in .Renviron and run the download chunk above",
    months = 0L
  )
}
knitr::kable(status)
```

## Source Inventory

```{r}
#| label: tbl-bra-tatic-inventory
#| tbl-cap: "Inventory of the TATIC year files"

inventory <- purrr::map(year_files, function(file) {
  year <- as.integer(stringr::str_extract(basename(file), "\\d{4}"))
  days <- data.table::fread(file = file, sep = ";", select = "TATIC_DAY",
                            colClasses = "character", showProgress = FALSE,
                            fill = Inf, header = TRUE)[[1]]
  tibble::tibble(
    Year          = year,
    File          = basename(file),
    `First day`   = min(days, na.rm = TRUE),
    `Last day`    = max(days, na.rm = TRUE),
    `Days`        = dplyr::n_distinct(days),
    Records       = scales::comma(length(days)),
    `Size (MB)`   = round(file.info(file)$size / 1024^2, 1)
  )
}) |>
  purrr::list_rbind()

if (length(year_files) == 0) {
  knitr::kable(tibble::tibble(
    note = paste0("No TATIC year files in ", bra_tatic_raw_dir, " yet.")))
} else {
  inventory |> dplyr::arrange(Year) |> knitr::kable()
}
```

## Coverage: days present, days missing

A day with no movements does not exist in a national movement feed, so a day missing from a
year file is a **gap** — either never downloaded, or downloaded and empty at the source.
Those two are different problems with different fixes, and they are counted separately.

This check exists because of what it caught elsewhere: the ODIN `dstaxi` table returns zero
rows for 2025-10-31 while another export of the same data has ~9,300 movements for it. The
downloader tolerated one missing day per month silently, and the hole stayed invisible for
as long as nobody counted.

```{r}
#| label: tbl-bra-tatic-coverage
#| tbl-cap: "TATIC daily coverage by year"

coverage <- purrr::map(year_files, function(file) {
  year <- as.integer(stringr::str_extract(basename(file), "\\d{4}"))
  days <- unique(data.table::fread(file = file, sep = ";", select = "TATIC_DAY",
                                   colClasses = "character", showProgress = FALSE,
                                   fill = Inf, header = TRUE)[[1]])
  last <- min(as.Date(sprintf("%d-12-31", year)), Sys.Date() - 1)
  want <- format(seq(as.Date(sprintf("%d-01-01", year)), last, by = "day"))
  miss <- setdiff(want, days)

  tibble::tibble(
    Year              = year,
    `Days expected`   = length(want),
    `Days present`    = length(intersect(want, days)),
    `Days missing`    = length(miss),
    `Missing (first)` = if (length(miss) == 0) "-"
                        else paste(utils::head(miss, 8), collapse = ", ")
  )
}) |>
  purrr::list_rbind()

if (length(year_files) == 0) {
  knitr::kable(tibble::tibble(note = "Nothing to check yet."))
} else {
  knitr::kable(coverage)
}
```

## What the Records Contain

The fields the pipeline expects are declared in `API_TATIC/ingest_tatic.R` and checked on
every ingestion, so a change in the source layout surfaces immediately instead of three
steps later.

| Group | Fields |
| --- | --- |
| Identity and context | `Callsign`, `EventType`, `AcftType`, `FlightType`, `RV`, `Adep`, `Ades`, `Locality`, `Runway`, `Transponder`, `Equipment` |
| Event date parts | `NR_ANO_EVENTO`, `NR_MES_EVENTO`, `NR_DIA_EVENTO` |
| Milestones (UTC) | `EOBT`, `wPush`, `cPush`, `wTaxi`, `Taxi`, `Hold`, `cRwy`, `cDep`, `Dep`, `Arr`, `ETA`, `cPos` |
| Added by the download | `TATIC_DAY` — the day requested from the API |

Each milestone has a `*_BSB` twin in Brasília local time. **Use the UTC ones.** Mixing the
two is not a hypothetical risk: in the TOTALBR work, a timestamp column carrying a spurious
`Europe/Paris` label produced a 50-minute offset that took several rounds to unwind, and a
one-hour local-time difference is far easier to mistake for real data than an obvious error.

```{r}
#| label: tbl-bra-tatic-fields
#| tbl-cap: "Fields actually present in the downloaded data"

if (length(year_files) == 0) {
  knitr::kable(tibble::tibble(note = "Nothing downloaded yet."))
} else {
  hdr <- names(data.table::fread(file = year_files[1], sep = ";", nrows = 1L,
                                 showProgress = FALSE, fill = Inf, header = TRUE))
  source(here::here("API_TATIC", "ingest_tatic.R"))
  knitr::kable(tibble::tibble(
    Field   = TATIC_KEY_FIELDS,
    Present = ifelse(TATIC_KEY_FIELDS %in% hdr, "yes", "MISSING")
  ))
}
```

## Milestones, and What They Allow

`dstaxi` gives one duration per movement: the gap between the block time and the movement
time. TATIC gives the sequence, so the same duration can be split into its parts — waiting
for a push-back clearance, the push itself, taxiing, and holding short of the runway.

That decomposition is the reason this dataset is here. It is **not** built in this document:
what the milestones mean operationally has to be agreed with CGNA before any metric rests on
them, and a plausible-looking difference of two timestamps is exactly the kind of thing that
survives review while being wrong. This document ends where the raw data is complete and
verified.

```{r}
#| label: tbl-bra-tatic-milestones
#| tbl-cap: "How often each milestone is populated"

if (length(year_files) == 0) {
  knitr::kable(tibble::tibble(note = "Nothing downloaded yet."))
} else {
  d <- data.table::fread(file = year_files[1], sep = ";", colClasses = "character",
                         showProgress = FALSE, fill = Inf, header = TRUE)
  ms <- intersect(TATIC_MILESTONES, names(d))
  knitr::kable(tibble::tibble(
    Milestone = ms,
    Populated = vapply(ms, function(x) sum(!is.na(d[[x]]) & nzchar(d[[x]])), integer(1)),
    Records   = nrow(d)
  ) |>
    dplyr::mutate(`%` = round(100 * Populated / pmax(Records, 1), 1)) |>
    dplyr::arrange(dplyr::desc(Populated)))
}
```

## The Pipeline End to End

1. **Token** — `TATIC_TOKEN` in `.Renviron`, never in a script, never in the repository.
2. **Download** — `download_tatic(tatic_years)`. Day by day, month by month, resumable;
   re-running it only fetches what is missing.
3. **Verify** — the coverage table above. Days missing are reported, not tolerated.
4. **Read** — `data-raw/tatic/tatic_<year>.csv`, one row per movement, `TATIC_DAY` for the
   day it was requested under.

`API_TATIC/ingest_tatic.R` remains for a different input: TATIC exported as `*.json` files
and dropped into `data-raw/tatic/` by hand. It flattens them into a single
`tatic-movements.csv`. The API route above does not need it — it writes CSV directly.

## Open Points

- **The milestone semantics** are not yet confirmed with CGNA: which clock each one is
  stamped on, and what a missing value means (not applicable, or not recorded).
- **The relationship to `dstaxi`** is unmeasured. Both describe the same movements at the
  same aerodromes; whether they agree, and where they do not, is the same comparison already
  built for taxi (`TAXI/compare_taxi_sources.R`) and is worth repeating here.
- **The study-airport filter** is not applied. The download keeps everything, exactly as
  TOTALBR does; narrowing to the twelve airports is a decision for the analysis, not the
  ingestion.
