# Brazil Taxi Data Ingestion

```{r}
#| label: setup
#| include: false

# Libraries, project paths and analysis parameters. `include: false` keeps the
# rendered document free of the package startup banners and "built under R
# version" warnings that loading the tidyverse emits.
source(here::here("_chapter-setup.R"))
```

This note documents the preparation of the Brazilian taxi-time files used for the 2025 Brazil-Europe operational efficiency update.
The raw taxi files were downloaded from the SharePoint (`Datalake-Rafael/decea-2025/TAXI`) to the local Downloads folder as a transient source archive.
To save local storage, the raw CSV files and the raw zip archive are not copied into the project archive.

The retained archive artefacts are filtered, study-airport extracts in an APDF-like parquet format, stored in `data/apdf`.
The project paths and analysis parameters are defined once in `_chapter-setup.R`.

::: {.callout-important}
## Prerequisites to run this document
The raw `dsTaxiYYYY.csv` files are **not** shipped with the repository. Get them with the `download-bra-taxi-data` chunk below (ODIN API), or place your own in `data-raw/`, or set `BRA_TAXI_ZIP` to an archive holding them. Then run the `prepare-bra-taxi-data` chunk once to generate the analytic outputs. Until then, the tables below render with a short notice instead of failing. See `README.md` for the full setup.
:::

## Getting the Source Data

The taxi source comes from the ICEA/DECEA **ODIN** API, one file per year in `data-raw/`,
named `dsTaxiYYYY.csv` — the same name the rest of this document already reads. Nothing
downstream changes: the inventory, the preparation, the validation and the dashboard take
their files from `data-raw/` and do not care how they got there.

Years you already hold as `dsTaxiYYYY.csv` stay where they are; the downloader inspects what
each covers and fetches **only the missing months**.

### What the table looks like

The table is **`dstaxi`** — the same ODIN API as KPI08 and TOTALBR with the table name
swapped, confirmed by DECEA. One request shows what it holds:

```{r}
#| label: discover-bra-taxi-api
#| eval: false

source(here::here("TAXI", "download_taxi.R"))

odin_probe("dstaxi")     # its columns, with the date-like ones flagged
```

It returns 16 columns — `mov`, `matricula`, `indicativo`, `adpartida`, `addestino`,
`tipovoo`, `vratipolinha`, `companhia`, `tipoaeronave`, `dhbimtra`, `dhvra`, `box`, `pista`,
`matchvra`, `class`, `referencia` — and three of them decide how the download is configured.
All three are already set in `TAXI/download_taxi.R`; nothing needs to be exported.

**The month window filters on `dhbimtra`.** The preparation below maps `dh_bimtra` to
`MVT_TIME`, the movement itself — landing for an arrival, take-off for a departure — and
`dh_vra` to `BLOCK_TIME`. Anchoring on the movement puts both phases in the month they are
reported in; anchoring on the block stamp would move a flight blocked off at 23:50 on the
31st into the wrong month, and the taxi time is the gap between the two. A wrong table
announces itself with an HTTP error, but a wrong date column returns the **wrong months with
no error at all**, so this one is stated rather than guessed.

**There is no `id` and no `pk`.** Offset pagination needs a total order, so the order is
built from the columns that identify a movement: `dhbimtra`, `indicativo`, `mov`, `dhvra`.
For the same reason nothing is de-duplicated on merge — with no unique key, an invented one
would delete real movements, such as the two legs a callsign flies in a day.

**The API names are not the file names.** The `dsTaxiYYYY.csv` files already in `data-raw/`
carry `dh_bimtra`, `dh_vra`, `match_vra` and `vra_tipo_linha`; the API returns the same
columns without the underscores (`dhbimtra`, `dhvra`, `matchvra`, `vratipolinha`). They are
renamed the moment a month arrives, so a downloaded year is read by exactly the same
preparation code as a year that came from the zip. The API's two extra columns (`class`,
`referencia`) travel along and are `NA` for the years that predate them.

Those files also have a **15th field with no name in the header**, read as `V15` — the
trailing payload described further below. It is the reason every read here passes
`header = TRUE`: left to itself, `fread` sees 14 names above 15-field rows, concludes row 1
cannot be a header, and demotes the names to data — making `dh_bimtra` "missing" from a file
that plainly has it.

### Download

`eval: false` — the download is slow, so it is not executed on render; run it when you need
data. It is **incremental and resumable**, sharing its engine with the KPI08 and TOTALBR
downloaders, which means it carries the lessons those cost:

- **one month per request window**, saved the moment it arrives, so an interrupted run keeps
  everything already fetched;
- a month is judged downloaded by the **days it contains**, never by a file existing — a
  month fetched while it was still the current one is partial and is fetched again once it
  closes;
- pagination advances by the **rows actually returned**, and orders on a **total** key, so a
  server-side page cap cannot truncate the download and ties cannot make a row arrive twice
  or not at all;
- existing year files are inspected with `fill = Inf`. `dsTaxi2024.csv` carries rows with
  one field more than its header (the trailing payload field described below), and without
  it the read stops at that line **with only a warning** — every later month would look
  missing and be downloaded again. `fill = TRUE` is not enough: it pads rows that are
  short and still stops on a row that is long;
- when a month **is** re-fetched, the downloaded month replaces that month in the year file
  instead of being appended to it. Without a unique key there is nothing to de-duplicate on
  afterwards, so the old copy is dropped before the merge.

```{r}
#| label: download-bra-taxi-data
#| eval: false

source(here::here("TAXI", "download_taxi.R"))

# the study period; drop to the current year for the routine refresh
download_taxi(dsTaxi_years, out_dir = bra_raw_data_dir)
```

The zip route still works and needs no API: set `BRA_TAXI_ZIP` to an archive holding the
`dsTaxiYYYY.csv` files, and the preparation reads the members directly without unpacking
anything into the project tree.

```{r}
#| label: check-bra-taxi-data

# what is on disk right now, and what to do if nothing is
have_files <- list.files(bra_raw_data_dir, pattern = "^dsTaxi20\\d{2}\\.csv$")
have_zip   <- Sys.getenv("BRA_TAXI_ZIP", unset = "")

status <- if (length(have_files) > 0) {
  tibble::tibble(
    source = paste0(length(have_files), " file(s) in ", bra_raw_data_dir),
    years  = paste(sort(stringr::str_extract(have_files, "\\d{4}")), collapse = ", ")
  )
} else if (nzchar(have_zip) && file.exists(have_zip)) {
  tibble::tibble(source = paste0("zip archive: ", basename(have_zip)), years = "read on demand")
} else {
  tibble::tibble(
    source = "no source data yet",
    years  = paste0("copy dsTaxiYYYY.csv into ", bra_raw_data_dir,
                    ", or set BRA_TAXI_ZIP to an archive holding them")
  )
}
knitr::kable(status)
```

## Source Inventory

```{r}
#| label: tbl-bra-source-inventory
#| tbl-cap: "Inventory of the raw taxi-time source files (dsTaxi)"

files <- list.files(
  bra_raw_data_dir,
  pattern = "^dsTaxi20\\d{2}\\.csv$",
  full.names = TRUE
)

inventory <- purrr::map(files, function(file) {

  year <- stringr::str_extract(basename(file), "\\d{4}") |>
    as.integer()

  if (!(year %in% dsTaxi_years)) {
    return(NULL)
  }

  dates <- vroom::vroom(
    file,
    delim = ";",
    col_select = "dh_bimtra",
    col_types = vroom::cols(dh_bimtra = vroom::col_datetime()),
    show_col_types = FALSE
  )

  tibble::tibble(
    Year = year,
    File = basename(file),
    `Start date` = min(dates$dh_bimtra, na.rm = TRUE),
    `End date` = max(dates$dh_bimtra, na.rm = TRUE),
    Records = scales::comma(nrow(dates)),
    `Size (MB)` = round(file.info(file)$size / 1024^2, 1)
  )

}) |>
  purrr::list_rbind()

if (nrow(inventory) == 0) {
  knitr::kable(tibble::tibble(
    note = paste0("No source files found in ", bra_raw_data_dir,
                  ". Place dsTaxiYYYY.csv there (or set BRA_TAXI_ZIP).")
  ))
} else {
  inventory |>
    dplyr::arrange(Year) |>
    knitr::kable()
}
```

The source files use semicolon-separated fields and include Brazilian airport movements beyond the Brazil-Europe study scope.
The harmonised parquet extracts retain only the airports defined in `_chapter-setup.R`.

## Harmonisation

The payload columns are mapped to the APDF-like fields required by the taxi-time pipeline.
For arrivals, the airport is derived from `addestino`; for departures, the airport is derived from `adpartida`.

| Source field | Harmonised field | Note |
| --- | --- | --- |
| `indicativo` | `FLTID` | Flight identifier |
| `matricula` | `REG` | Aircraft registration |
| `adpartida` | `ADEP` | Departure aerodrome |
| `addestino` | `ADES` | Destination aerodrome |
| `mov` | `PHASE` | `Arr` to `ARR`, `Dep` to `DEP` |
| `dh_bimtra` | `MVT_TIME` | Movement time |
| `dh_vra` | `BLOCK_TIME` | Block time |
| `tipoaeronave` | `ARCTYP` | Aircraft type |
| `box` | `STND` | Stand or gate position |
| `pista` | `RWY` | Runway |
| `match_vra` | `MATCH_VRA` | Payload match flag |
| trailing field when present | `KPI_PAYLOAD` | Retained for later diagnostic checks, not used in the harmonised metric |

The 2024 source file contains a trailing payload field on many rows.
The ingestion therefore reads up to 15 semicolon-separated fields and keeps the additional value as `KPI_PAYLOAD`.
This prevents a column shift in the 2024 data while leaving the supplied KPI calculation out of the PBWG taxi-time metric.

## Preparation Code

The following chunk is the reproducible preparation script used to create the filtered parquet extracts and the daily taxi-time outputs.
It is not evaluated during normal rendering because it reads the large transient source archive.

The taxi-time metric (GANP 20th-percentile reference and daily additional time) is computed in-chunk with base tidyverse, so the script no longer depends on the private PBWG package and runs from the project root on any machine.
Every machine-specific location is read from an environment variable with a project-relative fallback:

- `BRA_TAXI_ZIP` — path to the source zip archive (optional; when unset, the extracted `dsTaxiYYYY.csv` files in `data-raw/` are used as the source).
- `BRA_REPORT_DATA` — data directory of the sibling report project that receives the combined analytic CSVs (defaults to this project's `data/`).

```{r}
#| label: prepare-bra-taxi-data
#| eval: false


# ---- dependencies -----------------------------------------------------------
# Project setup: libraries, project paths and analysis parameters (variant,
# ref_year, data_years, min_n, max_txxt, p_ref, ref_key). The taxi-time metric
# is computed in this chunk, so the PBWG package is no longer required.
source(here::here("_chapter-setup.R"))

# ---- source/report locations (override with environment variables) ----------
# Optional transient source archive (zip). When empty, the extracted
# dsTaxiYYYY.csv files in data-raw/ are used as the source instead.
zip_path        <- Sys.getenv("BRA_TAXI_ZIP", unset = "")
# Report project data dir that receives the combined analytic CSVs; defaults to
# this project's data/ folder (override for a sibling report project).
report_data_dir <- Sys.getenv("BRA_REPORT_DATA", unset = bra_report_data)

dir_create(bra_apdf_dir)
walk(data_years, ~ dir_create(path(bra_output_dir, paste0("txxt-daily-", .x))))

normalise_text <- function(x) {
  x <- as.character(x)
  x <- trimws(x)
  x[x == ""] <- NA_character_
  x
}

parse_bra_datetime <- function(x) {
  x <- normalise_text(x)
  parse_date_time(x, orders = c("ymd HMS", "ymd HMS OS"), tz = "UTC", quiet = TRUE)
}

read_bra_taxi_year <- function(year) {
  member   <- paste0("dsTaxi", year, ".csv")
  csv_path <- path(bra_raw_data_dir, member)

  if (file.exists(csv_path)) {
    message("Reading ", csv_path)
    raw <- fread(
      csv_path,
      sep = ";",
      fill = 15,
      colClasses = "character",
      showProgress = FALSE
    )
  } else if (nzchar(zip_path) && file.exists(zip_path)) {
    message("Reading ", member, " from ", basename(zip_path))
    raw <- fread(
      cmd = sprintf("unzip -p %s %s", shQuote(zip_path), shQuote(member)),
      sep = ";",
      fill = 15,
      colClasses = "character",
      showProgress = FALSE
    )
  } else {
    stop(
      "No source found for ", year, ": expected ", csv_path,
      " or a zip archive via BRA_TAXI_ZIP."
    )
  }

  raw <- as_tibble(raw)

  if (!"V15" %in% names(raw)) {
    raw$V15 <- NA_character_
  }

  harmonised <- raw |>
    transmute(
      FLTID = normalise_text(indicativo),
      REG = normalise_text(matricula),
      ADEP = normalise_text(adpartida),
      ADES = normalise_text(addestino),
      PHASE = recode(normalise_text(mov), Arr = "ARR", Dep = "DEP", .default = NA_character_),
      MVT_TIME = parse_bra_datetime(dh_bimtra),
      BLOCK_TIME = parse_bra_datetime(dh_vra),
      ARCTYP = normalise_text(tipoaeronave),
      STND = normalise_text(box),
      RWY = normalise_text(pista),
      MATCH_VRA = normalise_text(match_vra),
      KPI_PAYLOAD = normalise_text(V15),
      ICAO = case_when(
        PHASE == "ARR" ~ ADES,
        PHASE == "DEP" ~ ADEP,
        TRUE ~ NA_character_
      )
    ) |>
    filter(ICAO %in% bra_apts)

  write_parquet(
    harmonised,
    path(bra_apdf_dir, paste0("PBWG-BRA-dsTaxi-apdf-", year, ".parquet"))
  )

  harmonised
}

# ---- taxi-time metric (reproduces the PBWG GANP p20 methodology) ------------
# Per-movement taxi time (minutes): arrivals = on-block - landing,
# departures = take-off - off-block. The movement date falls back to the
# movement time when the block time is absent.
prepare_txxt <- function(df) {
  df |>
    mutate(
      # Taxi runs in opposite directions per phase, so only the order of the
      # subtraction changes. Spelled out rather than left as the "else" branch of
      # an if_else: DEP is a case in its own right, and any unexpected phase
      # yields NA instead of silently being treated as a departure.
      TXXT = case_when(
        # taxi-in: landing -> on-block
        PHASE == "ARR" ~ as.numeric(difftime(BLOCK_TIME, MVT_TIME, units = "mins")),
        # taxi-out: off-block -> take-off
        PHASE == "DEP" ~ as.numeric(difftime(MVT_TIME, BLOCK_TIME, units = "mins")),
        TRUE ~ NA_real_
      ),
      DATE = as_date(coalesce(BLOCK_TIME, MVT_TIME)),
      VALID_TXXT = !is.na(TXXT) & TXXT > 0 & TXXT <= max_txxt
    )
}

# Reference: 20th-percentile taxi time per ICAO+PHASE+STND+RWY group, built from
# valid reference-year movements with a known stand and runway, keeping only
# groups with at least min_n samples.
build_txxt_reference <- function(mov_ref) {
  ref_start <- ymd_hms(paste0(ref_year, "-01-01 00:00:00"), tz = "UTC")
  ref_end   <- ymd_hms(paste0(ref_year, "-12-31 23:59:59"), tz = "UTC")

  mov_ref |>
    filter(
      VALID_TXXT,
      if_all(all_of(ref_key), ~ !is.na(.x)),
      BLOCK_TIME >= ref_start, BLOCK_TIME <= ref_end
    ) |>
    group_by(across(all_of(ref_key))) |>
    summarise(
      N = n(),
      TXXT_REF = quantile(TXXT, p_ref, type = 7, names = FALSE),
      .groups = "drop"
    ) |>
    filter(N >= min_n) |>
    select(all_of(ref_key), TXXT_REF)
}

# Apply the reference and aggregate to the daily analytic table. A movement is
# valid when it has a usable taxi time and its group has a reference value.
summarise_txxt_daily <- function(mov, reference) {
  mov |>
    filter(!is.na(DATE)) |>
    left_join(reference, by = ref_key) |>
    mutate(VALID = VALID_TXXT & !is.na(TXXT_REF)) |>
    group_by(ICAO, PHASE, DATE) |>
    summarise(
      MVTS_VALID   = sum(VALID),
      MVTS_NA      = sum(!VALID),
      TOT_TXXT     = sum(TXXT[VALID]),
      TOT_REF      = sum(TXXT_REF[VALID]),
      TOT_ADD_TIME = sum(TXXT[VALID] - TXXT_REF[VALID]),
      .groups = "drop"
    ) |>
    arrange(ICAO, PHASE, DATE)
}

# ---- run the pipeline -------------------------------------------------------
movements_by_year <- map(data_years, ~ prepare_txxt(read_bra_taxi_year(.x))) |>
  set_names(as.character(data_years))

reference_data <- build_txxt_reference(movements_by_year[[as.character(ref_year)]])

daily_by_year <- imap(movements_by_year, function(mov, year_chr) {
  year <- as.integer(year_chr)
  message("Applying ", ref_year, " reference to ", year)

  daily <- summarise_txxt_daily(mov, reference_data)

  out_name  <- bra_txxt_analytic_file(year)
  daily_dir <- path(bra_output_dir, paste0("txxt-daily-", year))
  write_csv(daily, path(daily_dir, out_name))
  write_csv(daily, path(report_data_dir, out_name))
  daily
})

# ---- coverage summary (per year and phase) ----------------------------------
coverage_summary <- imap(daily_by_year, function(daily, year_chr) {
  daily |>
    group_by(PHASE) |>
    summarise(
      MVTS_VALID   = sum(MVTS_VALID),
      MVTS_NA      = sum(MVTS_NA),
      TOT_TXXT     = sum(TOT_TXXT),
      TOT_REF      = sum(TOT_REF),
      TOT_ADD_TIME = sum(TOT_ADD_TIME),
      .groups = "drop"
    ) |>
    mutate(YEAR = as.integer(year_chr), MVTS_TOTAL = MVTS_VALID + MVTS_NA)
}) |>
  list_rbind() |>
  select(YEAR, PHASE, MVTS_TOTAL, MVTS_VALID, MVTS_NA,
         TOT_TXXT, TOT_REF, TOT_ADD_TIME)

write_csv(
  coverage_summary,
  path(report_data_dir, bra_coverage_file)
)
```

## Retained Artefacts

```{r}
#| label: tbl-bra-retained-artefacts

parquet_files <- list.files(bra_apdf_dir, pattern = "\\.parquet$", full.names = TRUE)

tibble::tibble(path = parquet_files) |>
  dplyr::mutate(
    file = basename(path),
    size_mb = round(file.info(path)$size / 1024^2, 1)
  ) |>
  dplyr::select(file, size_mb) |>
  knitr::kable()
```

The daily analytic outputs copied into the report project are:

```{r}
#| label: tbl-bra-report-outputs

report_output_files <- list.files(
  bra_report_data,
  pattern = bra_txxt_analytic_pattern,
  full.names = FALSE
)

tibble::tibble(file = report_output_files) |>
  knitr::kable()
```

## Coverage Check

The pipeline records the number of valid taxi-time movements (`MVTS_VALID`) and the number of movements that could not be used for the metric (`MVTS_NA`).
The observed taxi-time total (`TOT_TXXT`), reference total (`TOT_REF`), and additional-time total (`TOT_ADD_TIME`) are calculated only over `MVTS_VALID`.

```{r}
#| label: tbl-bra-txxt-coverage

coverage_path <- here::here("data", bra_coverage_file)

if (file.exists(coverage_path)) {
  readr::read_csv(coverage_path, show_col_types = FALSE) |>
    dplyr::mutate(
      MVTS_NA_SHARE = MVTS_NA / MVTS_TOTAL,
      AVG_TXXT = TOT_TXXT / MVTS_VALID,
      AVG_ADD_TIME = TOT_ADD_TIME / MVTS_VALID
    ) |>
    dplyr::select(YEAR, PHASE, MVTS_TOTAL, MVTS_VALID, MVTS_NA, MVTS_NA_SHARE, AVG_TXXT, AVG_ADD_TIME) |>
    dplyr::mutate(
      dplyr::across(c(MVTS_NA_SHARE, AVG_TXXT, AVG_ADD_TIME), ~ round(.x, 3))
    ) |>
    knitr::kable()
} else {
  knitr::kable(tibble::tibble(
    note = paste0("Coverage summary not found. Run the `prepare-bra-taxi-data` ",
                  "chunk to generate ", bra_coverage_file, " in data/.")
  ))
}
```

The current extracts retain the source KPI payload in the harmonised parquet files for later diagnostic comparison.
It is not used in the 2024 GANP 20th percentile PBWG reference calculation.

## Validation Against the Original Results

`reproduce_txxt.R` rebuilds the daily analytic outputs straight from the raw data and
compares them row by row with the original PBWG results kept in `golden/`. It is the
regression check that proves this repository's metric reproduces the published figures
without the private `PBWG` package.

It reads the raw source, so it is `eval: false` and run on demand.

```{r}
#| label: validate-bra-taxi-golden
#| eval: false

source(here::here("reproduce_txxt.R"))
```

```{r}
#| label: check-bra-taxi-golden

# which reference files are available to validate against
gold <- list.files(here::here("golden"), pattern = bra_txxt_analytic_pattern)
knitr::kable(tibble::tibble(
  `golden files` = if (length(gold) == 0) "none found in golden/" else
    paste(sort(gold), collapse = ", "),
  status = if (length(gold) == 0)
    "nothing to validate against" else
    "run the `validate-bra-taxi-golden` chunk above to compare"
))
```

## Dashboard

`index.html` is the interactive companion to this document: taxi time and additional time
per movement, Brazil against Europe, by year, airport and month.

The page carries its numbers in a generated `index-data.js` beside it, so it opens offline
by double-clicking. Run the chunk below whenever the analytic CSVs change to refresh them.
It is `eval: false` so rendering this document never rewrites the dashboard by surprise.

```{r}
#| label: build-bra-taxi-dashboard
#| eval: false

source(here::here("build_taxi_dashboard.R"))
```

```{r}
#| label: check-bra-taxi-dashboard

# is the page currently carrying data, or does it still need a build?
dash_file <- here::here("index.html")
data_file <- here::here("index-data.js")
dash_state <- if (!file.exists(dash_file)) {
  "index.html not found"
} else if (!file.exists(data_file)) {
  "no data yet — run the `build-bra-taxi-dashboard` chunk above"
} else {
  paste0("built ", format(file.info(data_file)$mtime, "%Y-%m-%d %H:%M"),
         " — open index.html to view the dashboard")
}
knitr::kable(tibble::tibble(dashboard = basename(dash_file), status = dash_state))
```

## The Pipeline End to End

Every step of the taxi-time workflow lives in this document, in the order it runs:

| Step | Section | Chunk | Runs on render? |
| --- | --- | --- | --- |
| 1. Provide the source | Getting the Source Data | `check-bra-taxi-data` | yes (reports only) |
| 2. Inspect what arrived | Source Inventory | `tbl-bra-source-inventory` | yes |
| 3. Harmonise and compute | Preparation Code | `prepare-bra-taxi-data` | **no** — run once, manually |
| 4. Check what was produced | Retained Artefacts / Coverage Check | `tbl-bra-*` | yes |
| 5. Validate against `golden/` | Validation | `validate-bra-taxi-golden` | **no** — reads the raw source |
| 6. Refresh the dashboard | Dashboard | `build-bra-taxi-dashboard` | **no** — writes a file |

The three heavy steps are `eval: false` on purpose: rendering the document is a read-only
report of the current state, never a re-run of the pipeline.
