# Brazil Taxi Data Ingestion

```{r}
source(here::here("_chapter-setup.R"))
```

This note documents the preparation of the Brazilian taxi-time files used for the 2025 Brazil-Europe operational efficiency update.
The raw taxi files were downloaded from the SharePoint (`Datalake-Rafael/decea-2025/TAXI`) to the local Downloads folder as a transient source archive.
To save local storage, the raw CSV files and the raw zip archive are not copied into the project archive.

The retained archive artefacts are filtered, study-airport extracts in an APDF-like parquet format, stored in:

```{r}
bra_apdf_dir    <- here("data", "apdf")
bra_output_dir  <- here("outputs")
bra_report_data <- here("data")
bra_raw_data_dir <- here("data-raw")
```

## Source Inventory

```{r}
#| label: tbl-bra-source-inventory
#| tbl-cap: "Inventory of the raw taxi-time source files (dsTaxi)"

files <- list.files(
  bra_raw_data_dir,
  pattern = "^dsTaxi20\\d{2}\\.csv$",
  full.names = TRUE
)

inventory <- purrr::map(files, function(file) {

  year <- stringr::str_extract(basename(file), "\\d{4}") |>
    as.integer()

  if (!(year %in% dsTaxi_years)) {
    return(NULL)
  }

  dates <- vroom::vroom(
    file,
    delim = ";",
    col_select = "dh_bimtra",
    col_types = vroom::cols(dh_bimtra = vroom::col_datetime()),
    show_col_types = FALSE
  )

  tibble::tibble(
    Year = year,
    File = basename(file),
    `Start date` = min(dates$dh_bimtra, na.rm = TRUE),
    `End date` = max(dates$dh_bimtra, na.rm = TRUE),
    Records = scales::comma(nrow(dates)),
    `Size (MB)` = round(file.info(file)$size / 1024^2, 1)
  )

}) |>
  purrr::list_rbind()

inventory |>
  dplyr::arrange(Year) |>
  knitr::kable()
```

The source files use semicolon-separated fields and include Brazilian airport movements beyond the Brazil-Europe study scope.
The harmonised parquet extracts retain only the airports defined in `_chapter-setup.R`.

## Harmonisation

The payload columns are mapped to the APDF-like fields required by the taxi-time pipeline.
For arrivals, the airport is derived from `addestino`; for departures, the airport is derived from `adpartida`.

| Source field | Harmonised field | Note |
| --- | --- | --- |
| `indicativo` | `FLTID` | Flight identifier |
| `matricula` | `REG` | Aircraft registration |
| `adpartida` | `ADEP` | Departure aerodrome |
| `addestino` | `ADES` | Destination aerodrome |
| `mov` | `PHASE` | `Arr` to `ARR`, `Dep` to `DEP` |
| `dh_bimtra` | `MVT_TIME` | Movement time |
| `dh_vra` | `BLOCK_TIME` | Block time |
| `tipoaeronave` | `ARCTYP` | Aircraft type |
| `box` | `STND` | Stand or gate position |
| `pista` | `RWY` | Runway |
| `match_vra` | `MATCH_VRA` | Payload match flag |
| trailing field when present | `KPI_PAYLOAD` | Retained for later diagnostic checks, not used in the harmonised metric |

The 2024 source file contains a trailing payload field on many rows.
The ingestion therefore reads up to 15 semicolon-separated fields and keeps the additional value as `KPI_PAYLOAD`.
This prevents a column shift in the 2024 data while leaving the supplied KPI calculation out of the PBWG taxi-time metric.

## Preparation Code

The following chunk is the reproducible preparation script used to create the filtered parquet extracts and the daily taxi-time outputs.
It is not evaluated during normal rendering because it reads the large transient source archive.

The script is portable: every machine-specific location is read from an environment variable with a project-relative fallback, so it runs from the project root on any machine without editing the code.

- `PBWG_SRC` — path to a local source checkout of the PBWG package (optional; when unset, the installed `PBWG` package is used).
- `BRA_TAXI_ZIP` — path to the source zip archive (optional; when unset, the extracted `dsTaxiYYYY.csv` files in `data-raw/` are used as the source).
- `BRA_REPORT_DATA` — data directory of the sibling report project that receives the combined analytic CSVs (defaults to this project's `data/`).

```{r}
#| label: prepare-bra-taxi-data
#| eval: false


# ---- dependencies -----------------------------------------------------------
# Local project setup (study airports, libraries, defaults).
source(here::here("_chapter-setup.R"))
library(fs)

# The taxi-time helpers live in the PBWG package. Either install it (loaded here
# with library(PBWG)), or point the PBWG_SRC environment variable at a local
# source checkout to load it with devtools::load_all().
pbwg_src <- Sys.getenv("PBWG_SRC", unset = "")
if (nzchar(pbwg_src)) {
  devtools::load_all(pbwg_src, quiet = TRUE)
} else {
  library(PBWG)
}

# ---- configuration (portable; override with environment variables) ----------
# Optional transient source archive (zip). When empty, the extracted
# dsTaxiYYYY.csv files in data-raw/ are used as the source instead.
zip_path        <- Sys.getenv("BRA_TAXI_ZIP", unset = "")
# Fallback source: already-extracted CSV files.
raw_data_dir    <- here("data-raw")
# Project root that receives the parquet extracts and daily outputs.
archive_root    <- here()
# Report project data dir that receives the combined analytic CSVs. Defaults to
# this project's data/ folder; override for a sibling report project.
report_data_dir <- Sys.getenv("BRA_REPORT_DATA", unset = here("data"))

variant        <- "icao_ganp_p20"
ref_year       <- 2024L
data_years     <- 2023:2025
min_n          <- 5L
max_txxt       <- 120
study_airports <- bra_apts

dir_create(path(archive_root, "data", "apdf"))
dir_create(path(archive_root, "outputs", "txxt-reference-2024"))
walk(data_years, ~ dir_create(path(archive_root, "outputs", paste0("txxt-daily-", .x))))

normalise_text <- function(x) {
  x <- as.character(x)
  x <- trimws(x)
  x[x == ""] <- NA_character_
  x
}

parse_bra_datetime <- function(x) {
  x <- normalise_text(x)
  parse_date_time(x, orders = c("ymd HMS", "ymd HMS OS"), tz = "UTC", quiet = TRUE)
}

read_bra_taxi_year <- function(year) {
  member   <- paste0("dsTaxi", year, ".csv")
  csv_path <- path(raw_data_dir, member)

  if (file.exists(csv_path)) {
    message("Reading ", csv_path)
    raw <- fread(
      csv_path,
      sep = ";",
      fill = 15,
      colClasses = "character",
      showProgress = FALSE
    )
  } else if (nzchar(zip_path) && file.exists(zip_path)) {
    message("Reading ", member, " from ", basename(zip_path))
    raw <- fread(
      cmd = sprintf("unzip -p %s %s", shQuote(zip_path), shQuote(member)),
      sep = ";",
      fill = 15,
      colClasses = "character",
      showProgress = FALSE
    )
  } else {
    stop(
      "No source found for ", year, ": expected ", csv_path,
      " or a zip archive via BRA_TAXI_ZIP."
    )
  }

  raw <- as_tibble(raw)

  if (!"V15" %in% names(raw)) {
    raw$V15 <- NA_character_
  }

  harmonised <- raw |>
    transmute(
      FLTID = normalise_text(indicativo),
      REG = normalise_text(matricula),
      ADEP = normalise_text(adpartida),
      ADES = normalise_text(addestino),
      PHASE = recode(normalise_text(mov), Arr = "ARR", Dep = "DEP", .default = NA_character_),
      MVT_TIME = parse_bra_datetime(dh_bimtra),
      BLOCK_TIME = parse_bra_datetime(dh_vra),
      ARCTYP = normalise_text(tipoaeronave),
      STND = normalise_text(box),
      RWY = normalise_text(pista),
      MATCH_VRA = normalise_text(match_vra),
      KPI_PAYLOAD = normalise_text(V15),
      ICAO = case_when(
        PHASE == "ARR" ~ ADES,
        PHASE == "DEP" ~ ADEP,
        TRUE ~ NA_character_
      )
    ) |>
    filter(ICAO %in% study_airports)

  write_parquet(
    harmonised,
    path(archive_root, "data", "apdf", paste0("PBWG-BRA-dsTaxi-apdf-", year, ".parquet"))
  )

  harmonised
}

samples_by_year <- map(data_years, function(year) {
  read_bra_taxi_year(year) |>
    prepare_txxt_reference_input(max_txxt = max_txxt) |>
    mutate(YEAR = year)
}) |>
  set_names(as.character(data_years))

reference_samples <- samples_by_year[[as.character(ref_year)]]
ref_start <- ymd_hms(paste0(ref_year, "-01-01 00:00:00"), tz = "UTC")
ref_end <- ymd_hms(paste0(ref_year, "-12-31 23:59:59"), tz = "UTC")

reference_data <- build_txxt_reference(
  txxt_samples = reference_samples,
  ref_start = ref_start,
  ref_end = ref_end,
  variant = variant,
  min_n = min_n,
  keep_below_threshold = TRUE,
  include_unknown = FALSE
)

walk(unique(reference_data$ICAO), function(airport) {
  write_pbwg_txxt_reference(
    data = reference_data,
    airport = airport,
    ref_period = as.character(ref_year),
    variant = variant,
    min_n = min_n,
    output_dir = path(archive_root, "outputs", "txxt-reference-2024"),
    region = "BRA"
  )
})

daily_by_year <- imap(samples_by_year, function(samples, year_chr) {
  year <- as.integer(year_chr)
  message("Applying 2024 reference to ", year)

  augmented <- apply_txxt_reference(
    txxt_samples = samples,
    reference_data = reference_data,
    valid_reference_only = TRUE
  )

  daily <- summarise_pbwg_txxt_daily(augmented, year = year)
  daily_dir <- path(archive_root, "outputs", paste0("txxt-daily-", year))

  walk(unique(daily$ICAO), function(airport) {
    write_pbwg_txxt(
      data = daily,
      year = year,
      ref_period = ref_year,
      variant = variant,
      output_dir = daily_dir,
      airport = airport,
      region = "BRA"
    )
  })

  combined_path <- write_pbwg_txxt(
    data = daily,
    year = year,
    ref_period = ref_year,
    variant = variant,
    output_dir = daily_dir,
    airport = NULL,
    region = "BRA"
  )

  file_copy(combined_path, path(report_data_dir, basename(combined_path)), overwrite = TRUE)
  daily
})
```

## Retained Artefacts

```{r}
#| label: tbl-bra-retained-artefacts

parquet_files <- list.files(bra_apdf_dir, pattern = "\\.parquet$", full.names = TRUE)

tibble::tibble(path = parquet_files) |>
  dplyr::mutate(
    file = basename(path),
    size_mb = round(file.info(path)$size / 1024^2, 1)
  ) |>
  dplyr::select(file, size_mb) |>
  knitr::kable()
```

The daily analytic outputs copied into the report project are:

```{r}
#| label: tbl-bra-report-outputs

report_output_files <- list.files(
  bra_report_data,
  pattern = "^PBWG-BRA-txxt-analytic-20[0-9]{2}-ref2024-icao_ganp_p20\\.csv$",
  full.names = FALSE
)

tibble::tibble(file = report_output_files) |>
  knitr::kable()
```

## Coverage Check

The pipeline records the number of valid taxi-time movements (`MVTS_VALID`) and the number of movements that could not be used for the metric (`MVTS_NA`).
The observed taxi-time total (`TOT_TXXT`), reference total (`TOT_REF`), and additional-time total (`TOT_ADD_TIME`) are calculated only over `MVTS_VALID`.

```{r}
#| label: tbl-bra-txxt-coverage

readr::read_csv(
  here::here("data", "BRA-txxt-coverage-summary-2023-2025.csv"),
  show_col_types = FALSE
) |>
  dplyr::mutate(
    MVTS_NA_SHARE = MVTS_NA / MVTS_TOTAL,
    AVG_TXXT = TOT_TXXT / MVTS_VALID,
    AVG_ADD_TIME = TOT_ADD_TIME / MVTS_VALID
  ) |>
  dplyr::select(YEAR, PHASE, MVTS_TOTAL, MVTS_VALID, MVTS_NA, MVTS_NA_SHARE, AVG_TXXT, AVG_ADD_TIME) |>
  dplyr::mutate(
    dplyr::across(c(MVTS_NA_SHARE, AVG_TXXT, AVG_ADD_TIME), ~ round(.x, 3))
  ) |>
  knitr::kable()
```

The current extracts retain the source KPI payload in the harmonised parquet files for later diagnostic comparison.
It is not used in the 2024 GANP 20th percentile PBWG reference calculation.
