# Brazil Taxi Data Ingestion

```{r}
#| label: setup
#| include: false

# Libraries, project paths and analysis parameters. `include: false` keeps the
# rendered document free of the package startup banners and "built under R
# version" warnings that loading the tidyverse emits.
source(here::here("_chapter-setup.R"))
```

This note documents the preparation of the Brazilian taxi-time files used for the 2025 Brazil-Europe operational efficiency update.
The raw taxi files were downloaded from the SharePoint (`Datalake-Rafael/decea-2025/TAXI`) to the local Downloads folder as a transient source archive.
To save local storage, the raw CSV files and the raw zip archive are not copied into the project archive.

The retained archive artefacts are filtered, study-airport extracts in an APDF-like parquet format, stored in `data/apdf`.
The project paths and analysis parameters are defined once in `_chapter-setup.R`.

::: {.callout-important}
## Prerequisites to run this document
The raw `dsTaxiYYYY.csv` files are **not** shipped with the repository. Before rendering, place them in `data-raw/` (or set the `BRA_TAXI_ZIP` environment variable) and run the `prepare-bra-taxi-data` chunk once to generate the analytic outputs. Until then, the tables below render with a short notice instead of failing. See `README.md` for the full setup.
:::

## Getting the Source Data

Unlike the ASMA/KPI08 side, taxi time has **no API**: the `dsTaxiYYYY.csv` files are
delivered as an archive and are not shipped with the repository. Provide them in one of
two ways, then everything below runs from the project root on any machine:

- copy the `dsTaxiYYYY.csv` files into `data-raw/`, or
- set `BRA_TAXI_ZIP` to a zip archive containing them (nothing is unpacked into the
  project tree; the preparation reads the members directly).

```{r}
#| label: check-bra-taxi-data

# what is on disk right now, and what to do if nothing is
have_files <- list.files(bra_raw_data_dir, pattern = "^dsTaxi20\\d{2}\\.csv$")
have_zip   <- Sys.getenv("BRA_TAXI_ZIP", unset = "")

status <- if (length(have_files) > 0) {
  tibble::tibble(
    source = paste0(length(have_files), " file(s) in ", bra_raw_data_dir),
    years  = paste(sort(stringr::str_extract(have_files, "\\d{4}")), collapse = ", ")
  )
} else if (nzchar(have_zip) && file.exists(have_zip)) {
  tibble::tibble(source = paste0("zip archive: ", basename(have_zip)), years = "read on demand")
} else {
  tibble::tibble(
    source = "no source data yet",
    years  = paste0("copy dsTaxiYYYY.csv into ", bra_raw_data_dir,
                    ", or set BRA_TAXI_ZIP to an archive holding them")
  )
}
knitr::kable(status)
```

## Source Inventory

```{r}
#| label: tbl-bra-source-inventory
#| tbl-cap: "Inventory of the raw taxi-time source files (dsTaxi)"

files <- list.files(
  bra_raw_data_dir,
  pattern = "^dsTaxi20\\d{2}\\.csv$",
  full.names = TRUE
)

inventory <- purrr::map(files, function(file) {

  year <- stringr::str_extract(basename(file), "\\d{4}") |>
    as.integer()

  if (!(year %in% dsTaxi_years)) {
    return(NULL)
  }

  dates <- vroom::vroom(
    file,
    delim = ";",
    col_select = "dh_bimtra",
    col_types = vroom::cols(dh_bimtra = vroom::col_datetime()),
    show_col_types = FALSE
  )

  tibble::tibble(
    Year = year,
    File = basename(file),
    `Start date` = min(dates$dh_bimtra, na.rm = TRUE),
    `End date` = max(dates$dh_bimtra, na.rm = TRUE),
    Records = scales::comma(nrow(dates)),
    `Size (MB)` = round(file.info(file)$size / 1024^2, 1)
  )

}) |>
  purrr::list_rbind()

if (nrow(inventory) == 0) {
  knitr::kable(tibble::tibble(
    note = paste0("No source files found in ", bra_raw_data_dir,
                  ". Place dsTaxiYYYY.csv there (or set BRA_TAXI_ZIP).")
  ))
} else {
  inventory |>
    dplyr::arrange(Year) |>
    knitr::kable()
}
```

The source files use semicolon-separated fields and include Brazilian airport movements beyond the Brazil-Europe study scope.
The harmonised parquet extracts retain only the airports defined in `_chapter-setup.R`.

## Harmonisation

The payload columns are mapped to the APDF-like fields required by the taxi-time pipeline.
For arrivals, the airport is derived from `addestino`; for departures, the airport is derived from `adpartida`.

| Source field | Harmonised field | Note |
| --- | --- | --- |
| `indicativo` | `FLTID` | Flight identifier |
| `matricula` | `REG` | Aircraft registration |
| `adpartida` | `ADEP` | Departure aerodrome |
| `addestino` | `ADES` | Destination aerodrome |
| `mov` | `PHASE` | `Arr` to `ARR`, `Dep` to `DEP` |
| `dh_bimtra` | `MVT_TIME` | Movement time |
| `dh_vra` | `BLOCK_TIME` | Block time |
| `tipoaeronave` | `ARCTYP` | Aircraft type |
| `box` | `STND` | Stand or gate position |
| `pista` | `RWY` | Runway |
| `match_vra` | `MATCH_VRA` | Payload match flag |
| trailing field when present | `KPI_PAYLOAD` | Retained for later diagnostic checks, not used in the harmonised metric |

The 2024 source file contains a trailing payload field on many rows.
The ingestion therefore reads up to 15 semicolon-separated fields and keeps the additional value as `KPI_PAYLOAD`.
This prevents a column shift in the 2024 data while leaving the supplied KPI calculation out of the PBWG taxi-time metric.

## Preparation Code

The following chunk is the reproducible preparation script used to create the filtered parquet extracts and the daily taxi-time outputs.
It is not evaluated during normal rendering because it reads the large transient source archive.

The taxi-time metric (GANP 20th-percentile reference and daily additional time) is computed in-chunk with base tidyverse, so the script no longer depends on the private PBWG package and runs from the project root on any machine.
Every machine-specific location is read from an environment variable with a project-relative fallback:

- `BRA_TAXI_ZIP` — path to the source zip archive (optional; when unset, the extracted `dsTaxiYYYY.csv` files in `data-raw/` are used as the source).
- `BRA_REPORT_DATA` — data directory of the sibling report project that receives the combined analytic CSVs (defaults to this project's `data/`).

```{r}
#| label: prepare-bra-taxi-data
#| eval: false


# ---- dependencies -----------------------------------------------------------
# Project setup: libraries, project paths and analysis parameters (variant,
# ref_year, data_years, min_n, max_txxt, p_ref, ref_key). The taxi-time metric
# is computed in this chunk, so the PBWG package is no longer required.
source(here::here("_chapter-setup.R"))

# ---- source/report locations (override with environment variables) ----------
# Optional transient source archive (zip). When empty, the extracted
# dsTaxiYYYY.csv files in data-raw/ are used as the source instead.
zip_path        <- Sys.getenv("BRA_TAXI_ZIP", unset = "")
# Report project data dir that receives the combined analytic CSVs; defaults to
# this project's data/ folder (override for a sibling report project).
report_data_dir <- Sys.getenv("BRA_REPORT_DATA", unset = bra_report_data)

dir_create(bra_apdf_dir)
walk(data_years, ~ dir_create(path(bra_output_dir, paste0("txxt-daily-", .x))))

normalise_text <- function(x) {
  x <- as.character(x)
  x <- trimws(x)
  x[x == ""] <- NA_character_
  x
}

parse_bra_datetime <- function(x) {
  x <- normalise_text(x)
  parse_date_time(x, orders = c("ymd HMS", "ymd HMS OS"), tz = "UTC", quiet = TRUE)
}

read_bra_taxi_year <- function(year) {
  member   <- paste0("dsTaxi", year, ".csv")
  csv_path <- path(bra_raw_data_dir, member)

  if (file.exists(csv_path)) {
    message("Reading ", csv_path)
    raw <- fread(
      csv_path,
      sep = ";",
      fill = 15,
      colClasses = "character",
      showProgress = FALSE
    )
  } else if (nzchar(zip_path) && file.exists(zip_path)) {
    message("Reading ", member, " from ", basename(zip_path))
    raw <- fread(
      cmd = sprintf("unzip -p %s %s", shQuote(zip_path), shQuote(member)),
      sep = ";",
      fill = 15,
      colClasses = "character",
      showProgress = FALSE
    )
  } else {
    stop(
      "No source found for ", year, ": expected ", csv_path,
      " or a zip archive via BRA_TAXI_ZIP."
    )
  }

  raw <- as_tibble(raw)

  if (!"V15" %in% names(raw)) {
    raw$V15 <- NA_character_
  }

  harmonised <- raw |>
    transmute(
      FLTID = normalise_text(indicativo),
      REG = normalise_text(matricula),
      ADEP = normalise_text(adpartida),
      ADES = normalise_text(addestino),
      PHASE = recode(normalise_text(mov), Arr = "ARR", Dep = "DEP", .default = NA_character_),
      MVT_TIME = parse_bra_datetime(dh_bimtra),
      BLOCK_TIME = parse_bra_datetime(dh_vra),
      ARCTYP = normalise_text(tipoaeronave),
      STND = normalise_text(box),
      RWY = normalise_text(pista),
      MATCH_VRA = normalise_text(match_vra),
      KPI_PAYLOAD = normalise_text(V15),
      ICAO = case_when(
        PHASE == "ARR" ~ ADES,
        PHASE == "DEP" ~ ADEP,
        TRUE ~ NA_character_
      )
    ) |>
    filter(ICAO %in% bra_apts)

  write_parquet(
    harmonised,
    path(bra_apdf_dir, paste0("PBWG-BRA-dsTaxi-apdf-", year, ".parquet"))
  )

  harmonised
}

# ---- taxi-time metric (reproduces the PBWG GANP p20 methodology) ------------
# Per-movement taxi time (minutes): arrivals = on-block - landing,
# departures = take-off - off-block. The movement date falls back to the
# movement time when the block time is absent.
prepare_txxt <- function(df) {
  df |>
    mutate(
      # Taxi runs in opposite directions per phase, so only the order of the
      # subtraction changes. Spelled out rather than left as the "else" branch of
      # an if_else: DEP is a case in its own right, and any unexpected phase
      # yields NA instead of silently being treated as a departure.
      TXXT = case_when(
        # taxi-in: landing -> on-block
        PHASE == "ARR" ~ as.numeric(difftime(BLOCK_TIME, MVT_TIME, units = "mins")),
        # taxi-out: off-block -> take-off
        PHASE == "DEP" ~ as.numeric(difftime(MVT_TIME, BLOCK_TIME, units = "mins")),
        TRUE ~ NA_real_
      ),
      DATE = as_date(coalesce(BLOCK_TIME, MVT_TIME)),
      VALID_TXXT = !is.na(TXXT) & TXXT > 0 & TXXT <= max_txxt
    )
}

# Reference: 20th-percentile taxi time per ICAO+PHASE+STND+RWY group, built from
# valid reference-year movements with a known stand and runway, keeping only
# groups with at least min_n samples.
build_txxt_reference <- function(mov_ref) {
  ref_start <- ymd_hms(paste0(ref_year, "-01-01 00:00:00"), tz = "UTC")
  ref_end   <- ymd_hms(paste0(ref_year, "-12-31 23:59:59"), tz = "UTC")

  mov_ref |>
    filter(
      VALID_TXXT,
      if_all(all_of(ref_key), ~ !is.na(.x)),
      BLOCK_TIME >= ref_start, BLOCK_TIME <= ref_end
    ) |>
    group_by(across(all_of(ref_key))) |>
    summarise(
      N = n(),
      TXXT_REF = quantile(TXXT, p_ref, type = 7, names = FALSE),
      .groups = "drop"
    ) |>
    filter(N >= min_n) |>
    select(all_of(ref_key), TXXT_REF)
}

# Apply the reference and aggregate to the daily analytic table. A movement is
# valid when it has a usable taxi time and its group has a reference value.
summarise_txxt_daily <- function(mov, reference) {
  mov |>
    filter(!is.na(DATE)) |>
    left_join(reference, by = ref_key) |>
    mutate(VALID = VALID_TXXT & !is.na(TXXT_REF)) |>
    group_by(ICAO, PHASE, DATE) |>
    summarise(
      MVTS_VALID   = sum(VALID),
      MVTS_NA      = sum(!VALID),
      TOT_TXXT     = sum(TXXT[VALID]),
      TOT_REF      = sum(TXXT_REF[VALID]),
      TOT_ADD_TIME = sum(TXXT[VALID] - TXXT_REF[VALID]),
      .groups = "drop"
    ) |>
    arrange(ICAO, PHASE, DATE)
}

# ---- run the pipeline -------------------------------------------------------
movements_by_year <- map(data_years, ~ prepare_txxt(read_bra_taxi_year(.x))) |>
  set_names(as.character(data_years))

reference_data <- build_txxt_reference(movements_by_year[[as.character(ref_year)]])

daily_by_year <- imap(movements_by_year, function(mov, year_chr) {
  year <- as.integer(year_chr)
  message("Applying ", ref_year, " reference to ", year)

  daily <- summarise_txxt_daily(mov, reference_data)

  out_name  <- bra_txxt_analytic_file(year)
  daily_dir <- path(bra_output_dir, paste0("txxt-daily-", year))
  write_csv(daily, path(daily_dir, out_name))
  write_csv(daily, path(report_data_dir, out_name))
  daily
})

# ---- coverage summary (per year and phase) ----------------------------------
coverage_summary <- imap(daily_by_year, function(daily, year_chr) {
  daily |>
    group_by(PHASE) |>
    summarise(
      MVTS_VALID   = sum(MVTS_VALID),
      MVTS_NA      = sum(MVTS_NA),
      TOT_TXXT     = sum(TOT_TXXT),
      TOT_REF      = sum(TOT_REF),
      TOT_ADD_TIME = sum(TOT_ADD_TIME),
      .groups = "drop"
    ) |>
    mutate(YEAR = as.integer(year_chr), MVTS_TOTAL = MVTS_VALID + MVTS_NA)
}) |>
  list_rbind() |>
  select(YEAR, PHASE, MVTS_TOTAL, MVTS_VALID, MVTS_NA,
         TOT_TXXT, TOT_REF, TOT_ADD_TIME)

write_csv(
  coverage_summary,
  path(report_data_dir, bra_coverage_file)
)
```

## Retained Artefacts

```{r}
#| label: tbl-bra-retained-artefacts

parquet_files <- list.files(bra_apdf_dir, pattern = "\\.parquet$", full.names = TRUE)

tibble::tibble(path = parquet_files) |>
  dplyr::mutate(
    file = basename(path),
    size_mb = round(file.info(path)$size / 1024^2, 1)
  ) |>
  dplyr::select(file, size_mb) |>
  knitr::kable()
```

The daily analytic outputs copied into the report project are:

```{r}
#| label: tbl-bra-report-outputs

report_output_files <- list.files(
  bra_report_data,
  pattern = bra_txxt_analytic_pattern,
  full.names = FALSE
)

tibble::tibble(file = report_output_files) |>
  knitr::kable()
```

## Coverage Check

The pipeline records the number of valid taxi-time movements (`MVTS_VALID`) and the number of movements that could not be used for the metric (`MVTS_NA`).
The observed taxi-time total (`TOT_TXXT`), reference total (`TOT_REF`), and additional-time total (`TOT_ADD_TIME`) are calculated only over `MVTS_VALID`.

```{r}
#| label: tbl-bra-txxt-coverage

coverage_path <- here::here("data", bra_coverage_file)

if (file.exists(coverage_path)) {
  readr::read_csv(coverage_path, show_col_types = FALSE) |>
    dplyr::mutate(
      MVTS_NA_SHARE = MVTS_NA / MVTS_TOTAL,
      AVG_TXXT = TOT_TXXT / MVTS_VALID,
      AVG_ADD_TIME = TOT_ADD_TIME / MVTS_VALID
    ) |>
    dplyr::select(YEAR, PHASE, MVTS_TOTAL, MVTS_VALID, MVTS_NA, MVTS_NA_SHARE, AVG_TXXT, AVG_ADD_TIME) |>
    dplyr::mutate(
      dplyr::across(c(MVTS_NA_SHARE, AVG_TXXT, AVG_ADD_TIME), ~ round(.x, 3))
    ) |>
    knitr::kable()
} else {
  knitr::kable(tibble::tibble(
    note = paste0("Coverage summary not found. Run the `prepare-bra-taxi-data` ",
                  "chunk to generate ", bra_coverage_file, " in data/.")
  ))
}
```

The current extracts retain the source KPI payload in the harmonised parquet files for later diagnostic comparison.
It is not used in the 2024 GANP 20th percentile PBWG reference calculation.

## Validation Against the Original Results

`reproduce_txxt.R` rebuilds the daily analytic outputs straight from the raw data and
compares them row by row with the original PBWG results kept in `golden/`. It is the
regression check that proves this repository's metric reproduces the published figures
without the private `PBWG` package.

It reads the raw source, so it is `eval: false` and run on demand.

```{r}
#| label: validate-bra-taxi-golden
#| eval: false

source(here::here("reproduce_txxt.R"))
```

```{r}
#| label: check-bra-taxi-golden

# which reference files are available to validate against
gold <- list.files(here::here("golden"), pattern = bra_txxt_analytic_pattern)
knitr::kable(tibble::tibble(
  `golden files` = if (length(gold) == 0) "none found in golden/" else
    paste(sort(gold), collapse = ", "),
  status = if (length(gold) == 0)
    "nothing to validate against" else
    "run the `validate-bra-taxi-golden` chunk above to compare"
))
```

## Dashboard

`index.html` is the interactive companion to this document: taxi time and additional time
per movement, Brazil against Europe, by year, airport and month.

The page carries its numbers in a generated `index-data.js` beside it, so it opens offline
by double-clicking. Run the chunk below whenever the analytic CSVs change to refresh them.
It is `eval: false` so rendering this document never rewrites the dashboard by surprise.

```{r}
#| label: build-bra-taxi-dashboard
#| eval: false

source(here::here("build_taxi_dashboard.R"))
```

```{r}
#| label: check-bra-taxi-dashboard

# is the page currently carrying data, or does it still need a build?
dash_file <- here::here("index.html")
data_file <- here::here("index-data.js")
dash_state <- if (!file.exists(dash_file)) {
  "index.html not found"
} else if (!file.exists(data_file)) {
  "no data yet — run the `build-bra-taxi-dashboard` chunk above"
} else {
  paste0("built ", format(file.info(data_file)$mtime, "%Y-%m-%d %H:%M"),
         " — open index.html to view the dashboard")
}
knitr::kable(tibble::tibble(dashboard = basename(dash_file), status = dash_state))
```

## The Pipeline End to End

Every step of the taxi-time workflow lives in this document, in the order it runs:

| Step | Section | Chunk | Runs on render? |
| --- | --- | --- | --- |
| 1. Provide the source | Getting the Source Data | `check-bra-taxi-data` | yes (reports only) |
| 2. Inspect what arrived | Source Inventory | `tbl-bra-source-inventory` | yes |
| 3. Harmonise and compute | Preparation Code | `prepare-bra-taxi-data` | **no** — run once, manually |
| 4. Check what was produced | Retained Artefacts / Coverage Check | `tbl-bra-*` | yes |
| 5. Validate against `golden/` | Validation | `validate-bra-taxi-golden` | **no** — reads the raw source |
| 6. Refresh the dashboard | Dashboard | `build-bra-taxi-dashboard` | **no** — writes a file |

The three heavy steps are `eval: false` on purpose: rendering the document is a read-only
report of the current state, never a re-run of the pipeline.
