# Brazil Taxi Data Ingestion

```{r}
source(here::here("_chapter-setup.R"))
```

This note documents the preparation of the Brazilian taxi-time files used for the 2025 Brazil-Europe operational efficiency update.
The raw taxi files were downloaded from the SharePoint (`Datalake-Rafael/decea-2025/TAXI`) to the local Downloads folder as a transient source archive.
To save local storage, the raw CSV files and the raw zip archive are not copied into the project archive.

The retained archive artefacts are filtered, study-airport extracts in an APDF-like parquet format, stored in `data/apdf`.
The project paths and analysis parameters are defined once in `_chapter-setup.R`.

::: {.callout-important}
## Prerequisites to run this document
The raw `dsTaxiYYYY.csv` files are **not** shipped with the repository. Before rendering, place them in `data-raw/` (or set the `BRA_TAXI_ZIP` environment variable) and run the `prepare-bra-taxi-data` chunk once to generate the analytic outputs. Until then, the tables below render with a short notice instead of failing. See `README.md` for the full setup.
:::

## Source Inventory

```{r}
#| label: tbl-bra-source-inventory
#| tbl-cap: "Inventory of the raw taxi-time source files (dsTaxi)"

files <- list.files(
  bra_raw_data_dir,
  pattern = "^dsTaxi20\\d{2}\\.csv$",
  full.names = TRUE
)

inventory <- purrr::map(files, function(file) {

  year <- stringr::str_extract(basename(file), "\\d{4}") |>
    as.integer()

  if (!(year %in% dsTaxi_years)) {
    return(NULL)
  }

  dates <- vroom::vroom(
    file,
    delim = ";",
    col_select = "dh_bimtra",
    col_types = vroom::cols(dh_bimtra = vroom::col_datetime()),
    show_col_types = FALSE
  )

  tibble::tibble(
    Year = year,
    File = basename(file),
    `Start date` = min(dates$dh_bimtra, na.rm = TRUE),
    `End date` = max(dates$dh_bimtra, na.rm = TRUE),
    Records = scales::comma(nrow(dates)),
    `Size (MB)` = round(file.info(file)$size / 1024^2, 1)
  )

}) |>
  purrr::list_rbind()

if (nrow(inventory) == 0) {
  knitr::kable(tibble::tibble(
    note = paste0("No source files found in ", bra_raw_data_dir,
                  ". Place dsTaxiYYYY.csv there (or set BRA_TAXI_ZIP).")
  ))
} else {
  inventory |>
    dplyr::arrange(Year) |>
    knitr::kable()
}
```

The source files use semicolon-separated fields and include Brazilian airport movements beyond the Brazil-Europe study scope.
The harmonised parquet extracts retain only the airports defined in `_chapter-setup.R`.

## Harmonisation

The payload columns are mapped to the APDF-like fields required by the taxi-time pipeline.
For arrivals, the airport is derived from `addestino`; for departures, the airport is derived from `adpartida`.

| Source field | Harmonised field | Note |
| --- | --- | --- |
| `indicativo` | `FLTID` | Flight identifier |
| `matricula` | `REG` | Aircraft registration |
| `adpartida` | `ADEP` | Departure aerodrome |
| `addestino` | `ADES` | Destination aerodrome |
| `mov` | `PHASE` | `Arr` to `ARR`, `Dep` to `DEP` |
| `dh_bimtra` | `MVT_TIME` | Movement time |
| `dh_vra` | `BLOCK_TIME` | Block time |
| `tipoaeronave` | `ARCTYP` | Aircraft type |
| `box` | `STND` | Stand or gate position |
| `pista` | `RWY` | Runway |
| `match_vra` | `MATCH_VRA` | Payload match flag |
| trailing field when present | `KPI_PAYLOAD` | Retained for later diagnostic checks, not used in the harmonised metric |

The 2024 source file contains a trailing payload field on many rows.
The ingestion therefore reads up to 15 semicolon-separated fields and keeps the additional value as `KPI_PAYLOAD`.
This prevents a column shift in the 2024 data while leaving the supplied KPI calculation out of the PBWG taxi-time metric.

## Preparation Code

The following chunk is the reproducible preparation script used to create the filtered parquet extracts and the daily taxi-time outputs.
It is not evaluated during normal rendering because it reads the large transient source archive.

The taxi-time metric (GANP 20th-percentile reference and daily additional time) is computed in-chunk with base tidyverse, so the script no longer depends on the private PBWG package and runs from the project root on any machine.
Every machine-specific location is read from an environment variable with a project-relative fallback:

- `BRA_TAXI_ZIP` — path to the source zip archive (optional; when unset, the extracted `dsTaxiYYYY.csv` files in `data-raw/` are used as the source).
- `BRA_REPORT_DATA` — data directory of the sibling report project that receives the combined analytic CSVs (defaults to this project's `data/`).

```{r}
#| label: prepare-bra-taxi-data
#| eval: false


# ---- dependencies -----------------------------------------------------------
# Project setup: libraries, project paths and analysis parameters (variant,
# ref_year, data_years, min_n, max_txxt, p_ref, ref_key). The taxi-time metric
# is computed in this chunk, so the PBWG package is no longer required.
source(here::here("_chapter-setup.R"))

# ---- source/report locations (override with environment variables) ----------
# Optional transient source archive (zip). When empty, the extracted
# dsTaxiYYYY.csv files in data-raw/ are used as the source instead.
zip_path        <- Sys.getenv("BRA_TAXI_ZIP", unset = "")
# Report project data dir that receives the combined analytic CSVs; defaults to
# this project's data/ folder (override for a sibling report project).
report_data_dir <- Sys.getenv("BRA_REPORT_DATA", unset = bra_report_data)

dir_create(bra_apdf_dir)
walk(data_years, ~ dir_create(path(bra_output_dir, paste0("txxt-daily-", .x))))

normalise_text <- function(x) {
  x <- as.character(x)
  x <- trimws(x)
  x[x == ""] <- NA_character_
  x
}

parse_bra_datetime <- function(x) {
  x <- normalise_text(x)
  parse_date_time(x, orders = c("ymd HMS", "ymd HMS OS"), tz = "UTC", quiet = TRUE)
}

read_bra_taxi_year <- function(year) {
  member   <- paste0("dsTaxi", year, ".csv")
  csv_path <- path(bra_raw_data_dir, member)

  if (file.exists(csv_path)) {
    message("Reading ", csv_path)
    raw <- fread(
      csv_path,
      sep = ";",
      fill = 15,
      colClasses = "character",
      showProgress = FALSE
    )
  } else if (nzchar(zip_path) && file.exists(zip_path)) {
    message("Reading ", member, " from ", basename(zip_path))
    raw <- fread(
      cmd = sprintf("unzip -p %s %s", shQuote(zip_path), shQuote(member)),
      sep = ";",
      fill = 15,
      colClasses = "character",
      showProgress = FALSE
    )
  } else {
    stop(
      "No source found for ", year, ": expected ", csv_path,
      " or a zip archive via BRA_TAXI_ZIP."
    )
  }

  raw <- as_tibble(raw)

  if (!"V15" %in% names(raw)) {
    raw$V15 <- NA_character_
  }

  harmonised <- raw |>
    transmute(
      FLTID = normalise_text(indicativo),
      REG = normalise_text(matricula),
      ADEP = normalise_text(adpartida),
      ADES = normalise_text(addestino),
      PHASE = recode(normalise_text(mov), Arr = "ARR", Dep = "DEP", .default = NA_character_),
      MVT_TIME = parse_bra_datetime(dh_bimtra),
      BLOCK_TIME = parse_bra_datetime(dh_vra),
      ARCTYP = normalise_text(tipoaeronave),
      STND = normalise_text(box),
      RWY = normalise_text(pista),
      MATCH_VRA = normalise_text(match_vra),
      KPI_PAYLOAD = normalise_text(V15),
      ICAO = case_when(
        PHASE == "ARR" ~ ADES,
        PHASE == "DEP" ~ ADEP,
        TRUE ~ NA_character_
      )
    ) |>
    filter(ICAO %in% bra_apts)

  write_parquet(
    harmonised,
    path(bra_apdf_dir, paste0("PBWG-BRA-dsTaxi-apdf-", year, ".parquet"))
  )

  harmonised
}

# ---- taxi-time metric (reproduces the PBWG GANP p20 methodology) ------------
# Per-movement taxi time (minutes): arrivals = on-block - landing,
# departures = take-off - off-block. The movement date falls back to the
# movement time when the block time is absent.
prepare_txxt <- function(df) {
  df |>
    mutate(
      TXXT = if_else(
        PHASE == "ARR",
        as.numeric(difftime(BLOCK_TIME, MVT_TIME, units = "mins")),
        as.numeric(difftime(MVT_TIME, BLOCK_TIME, units = "mins"))
      ),
      DATE = as_date(coalesce(BLOCK_TIME, MVT_TIME)),
      VALID_TXXT = !is.na(TXXT) & TXXT > 0 & TXXT <= max_txxt
    )
}

# Reference: 20th-percentile taxi time per ICAO+PHASE+STND+RWY group, built from
# valid reference-year movements with a known stand and runway, keeping only
# groups with at least min_n samples.
build_txxt_reference <- function(mov_ref) {
  ref_start <- ymd_hms(paste0(ref_year, "-01-01 00:00:00"), tz = "UTC")
  ref_end   <- ymd_hms(paste0(ref_year, "-12-31 23:59:59"), tz = "UTC")

  mov_ref |>
    filter(
      VALID_TXXT,
      if_all(all_of(ref_key), ~ !is.na(.x)),
      BLOCK_TIME >= ref_start, BLOCK_TIME <= ref_end
    ) |>
    group_by(across(all_of(ref_key))) |>
    summarise(
      N = n(),
      TXXT_REF = quantile(TXXT, p_ref, type = 7, names = FALSE),
      .groups = "drop"
    ) |>
    filter(N >= min_n) |>
    select(all_of(ref_key), TXXT_REF)
}

# Apply the reference and aggregate to the daily analytic table. A movement is
# valid when it has a usable taxi time and its group has a reference value.
summarise_txxt_daily <- function(mov, reference) {
  mov |>
    filter(!is.na(DATE)) |>
    left_join(reference, by = ref_key) |>
    mutate(VALID = VALID_TXXT & !is.na(TXXT_REF)) |>
    group_by(ICAO, PHASE, DATE) |>
    summarise(
      MVTS_VALID   = sum(VALID),
      MVTS_NA      = sum(!VALID),
      TOT_TXXT     = sum(TXXT[VALID]),
      TOT_REF      = sum(TXXT_REF[VALID]),
      TOT_ADD_TIME = sum(TXXT[VALID] - TXXT_REF[VALID]),
      .groups = "drop"
    ) |>
    arrange(ICAO, PHASE, DATE)
}

# ---- run the pipeline -------------------------------------------------------
movements_by_year <- map(data_years, ~ prepare_txxt(read_bra_taxi_year(.x))) |>
  set_names(as.character(data_years))

reference_data <- build_txxt_reference(movements_by_year[[as.character(ref_year)]])

daily_by_year <- imap(movements_by_year, function(mov, year_chr) {
  year <- as.integer(year_chr)
  message("Applying ", ref_year, " reference to ", year)

  daily <- summarise_txxt_daily(mov, reference_data)

  out_name  <- bra_txxt_analytic_file(year)
  daily_dir <- path(bra_output_dir, paste0("txxt-daily-", year))
  write_csv(daily, path(daily_dir, out_name))
  write_csv(daily, path(report_data_dir, out_name))
  daily
})

# ---- coverage summary (per year and phase) ----------------------------------
coverage_summary <- imap(daily_by_year, function(daily, year_chr) {
  daily |>
    group_by(PHASE) |>
    summarise(
      MVTS_VALID   = sum(MVTS_VALID),
      MVTS_NA      = sum(MVTS_NA),
      TOT_TXXT     = sum(TOT_TXXT),
      TOT_REF      = sum(TOT_REF),
      TOT_ADD_TIME = sum(TOT_ADD_TIME),
      .groups = "drop"
    ) |>
    mutate(YEAR = as.integer(year_chr), MVTS_TOTAL = MVTS_VALID + MVTS_NA)
}) |>
  list_rbind() |>
  select(YEAR, PHASE, MVTS_TOTAL, MVTS_VALID, MVTS_NA,
         TOT_TXXT, TOT_REF, TOT_ADD_TIME)

write_csv(
  coverage_summary,
  path(report_data_dir, bra_coverage_file)
)
```

## Retained Artefacts

```{r}
#| label: tbl-bra-retained-artefacts

parquet_files <- list.files(bra_apdf_dir, pattern = "\\.parquet$", full.names = TRUE)

tibble::tibble(path = parquet_files) |>
  dplyr::mutate(
    file = basename(path),
    size_mb = round(file.info(path)$size / 1024^2, 1)
  ) |>
  dplyr::select(file, size_mb) |>
  knitr::kable()
```

The daily analytic outputs copied into the report project are:

```{r}
#| label: tbl-bra-report-outputs

report_output_files <- list.files(
  bra_report_data,
  pattern = bra_txxt_analytic_pattern,
  full.names = FALSE
)

tibble::tibble(file = report_output_files) |>
  knitr::kable()
```

## Coverage Check

The pipeline records the number of valid taxi-time movements (`MVTS_VALID`) and the number of movements that could not be used for the metric (`MVTS_NA`).
The observed taxi-time total (`TOT_TXXT`), reference total (`TOT_REF`), and additional-time total (`TOT_ADD_TIME`) are calculated only over `MVTS_VALID`.

```{r}
#| label: tbl-bra-txxt-coverage

coverage_path <- here::here("data", bra_coverage_file)

if (file.exists(coverage_path)) {
  readr::read_csv(coverage_path, show_col_types = FALSE) |>
    dplyr::mutate(
      MVTS_NA_SHARE = MVTS_NA / MVTS_TOTAL,
      AVG_TXXT = TOT_TXXT / MVTS_VALID,
      AVG_ADD_TIME = TOT_ADD_TIME / MVTS_VALID
    ) |>
    dplyr::select(YEAR, PHASE, MVTS_TOTAL, MVTS_VALID, MVTS_NA, MVTS_NA_SHARE, AVG_TXXT, AVG_ADD_TIME) |>
    dplyr::mutate(
      dplyr::across(c(MVTS_NA_SHARE, AVG_TXXT, AVG_ADD_TIME), ~ round(.x, 3))
    ) |>
    knitr::kable()
} else {
  knitr::kable(tibble::tibble(
    note = paste0("Coverage summary not found. Run the `prepare-bra-taxi-data` ",
                  "chunk to generate ", bra_coverage_file, " in data/.")
  ))
}
```

The current extracts retain the source KPI payload in the harmonised parquet files for later diagnostic comparison.
It is not used in the 2024 GANP 20th percentile PBWG reference calculation.
