# Brazil TOTALBR Ingestion

```{r}
#| label: setup
#| include: false

# Libraries, project paths and parameters — including every TOTALBR setting
# (totalbr_data_years, the directories, the column names). Change the study
# period there, not here.
source(here::here("_chapter-setup.R"))
```

This note documents the preparation of the Brazilian **TOTALBR** files — the all-Brazil
movement table, ODIN table `total_brasil`. It follows the same shape as
`Taxi-BRA-ingestion.qmd` and `ASMA-BRA-ingestion.qmd`: get the data, check what arrived,
inventory it, harmonise it, and verify the coverage before anything downstream uses it.

TOTALBR differs from the other two in what it is *for*. Taxi-time and ASMA are performance
metrics computed at the twelve study airports; TOTALBR is **national traffic volume** — one
row per flight anywhere in Brazil, including aerodromes outside the study set and flights
with no aerodrome at all (`AFIL` departures, `ZZZZ` destinations). It carries no times to
subtract and no unimpeded reference, so there is no reference year and no percentile here.
It is the denominator and the context: how much traffic there was, where, and of what kind.

::: {.callout-important}
## Prerequisites to run this document
The raw `totalbr` CSV files are **not** shipped with the repository. Get them with the
download chunk below (ODIN API), or drop files you already hold into
`data-raw/totalbr/` named `totalbr_<year>.csv`. Until then, the tables below render with
a short notice instead of failing.
:::

## Getting the Source Data

### Check the endpoint first

A full download runs for hours, so confirm the endpoint and its column names with a single
small request before starting one. This is cheap and catches a renamed column immediately
instead of after a wasted afternoon.

```{r}
#| label: probe-bra-totalbr-api
#| eval: false

source(here::here("ODIN", "download_odin.R"))
odin_probe(totalbr_table)      # prints the columns and flags the date-like ones
```

Run on 2026-07-30 this answered with 18 columns —
`pk, id, dt_dia, dh_inicio, dh_fim, co_indicativo, co_addep, co_addes, dh_eobt, dh_eet,
co_matricula, co_modelo, ds_rota, li_orgaos, li_tipovoo, li_regravoo, li_prnav, li_rvsm` —
of which five are date-like (`dt_dia`, `dh_inicio`, `dh_fim`, `dh_eobt`, `dh_eet`). That
confirms the two choices the downloader is built on: the month windows filter on `dt_dia`,
and the unique key is `pk`.

### Download

The chunk below is `eval: false` — the download is slow, so it is not executed on render;
run it manually when you need data.

It is **incremental and resumable**, exactly like the KPI08 downloader (they share the
engine in `ODIN/download_odin.R`): one month per request window, each month saved under
`parts/` the moment it arrives, and a month judged already-downloaded by **the days it
contains** rather than by a file existing. So an interrupted run keeps everything it had,
and a month fetched while it was still the current month is fetched again once it closes
instead of staying frozen half-full.

Years you already hold as files go straight into `data-raw/totalbr/` with the same naming
(`totalbr_2019.csv`); the downloader inspects what each covers and fetches only the gaps.

```{r}
#| label: download-bra-totalbr-data
#| eval: false

# sourcing only defines download_totalbr(); nothing is fetched until it is called
source(here::here("TOTALBR", "download_totalbr.R"))

# the full study period; drop to the current year for the routine refresh
download_totalbr(totalbr_data_years, out_dir = bra_totalbr_raw_dir)
```

## Checking What Arrived

```{r}
#| label: check-bra-totalbr-data

# what is on disk right now, and what to do if nothing is
have_files <- list.files(bra_totalbr_raw_dir,
                         pattern = "^totalbr_[0-9]{4}\\.csv$")

if (length(have_files) == 0) {
  knitr::kable(tibble::tibble(
    status = "no source data yet",
    `what to do` = paste0(
      "Run the `download-bra-totalbr-data` chunk above (or, from a terminal, ",
      "`Rscript TOTALBR/download_totalbr.R ",
      min(totalbr_data_years), " … ", max(totalbr_data_years), "`)."
    )
  ))
} else {
  knitr::kable(tibble::tibble(
    status = paste0(length(have_files), " source file(s) in ", bra_totalbr_raw_dir),
    years  = paste(sort(stringr::str_extract(have_files, "[0-9]{4}")), collapse = ", ")
  ))
}
```

## Source Inventory

```{r}
#| label: tbl-bra-totalbr-source-inventory
#| tbl-cap: "Inventory of the raw TOTALBR source files"

files <- list.files(bra_totalbr_raw_dir, pattern = "^totalbr_[0-9]{4}\\.csv$",
                    full.names = TRUE)

if (length(files) == 0) {
  knitr::kable(tibble::tibble(
    note = paste0("No source files in ", bra_totalbr_raw_dir,
                  " yet. Run the `download-bra-totalbr-data` chunk above.")
  ))
} else {
  purrr::map(files, function(file) {
    tibble::tibble(
      File        = basename(file),
      Year        = stringr::str_extract(basename(file), "[0-9]{4}"),
      `Size (MB)` = round(file.info(file)$size / 1024^2, 1),
      Downloaded  = format(file.info(file)$mtime, "%Y-%m-%d %H:%M")
    )
  }) |>
    purrr::list_rbind() |>
    dplyr::arrange(Year) |>
    knitr::kable()
}
```

The files are semicolon-delimited UTF-8, written by the downloader. The per-month parts it
fetched live under `data-raw/totalbr/parts/` and are what makes a re-run cheap; the year
file is the merge of them, de-duplicated on `pk` and sorted by `dt_dia`.

## What the Columns Mean

| Source field | Meaning | Note |
| --- | --- | --- |
| `pk` | Row hash | **The unique key.** De-duplication keys on this |
| `id` | Callsign + ADEP + ADES | e.g. `PPBLVSBPASBSM`. **Repeats** on every flight of that route — never a key |
| `dt_dia` | Movement day/time | What the month windows filter on |
| `dh_inicio` | Start of the movement | Equal to `dt_dia` in most sampled rows |
| `dh_fim` | End of the movement | Equal to `dh_inicio` when no duration was recorded |
| `co_indicativo` | Callsign | `PPBLV`, `AZU2429` |
| `co_addep` | Departure aerodrome | `AFIL` = flight plan filed airborne, i.e. no departure aerodrome |
| `co_addes` | Destination aerodrome | `ZZZZ` = destination not an ICAO aerodrome |
| `dh_eobt` | Estimated off-block time | From the flight plan |
| `dh_eet` | Estimated elapsed time | Encoded as a timestamp, not a duration — read with care |
| `co_matricula` | Registration | Largely null in the sample |
| `co_modelo` | Aircraft type | `SR22`, `AT76`; `XXXX` where unknown |
| `ds_rota` | Route | Largely null in the sample |
| `li_orgaos` | ATS units involved | JSON array, e.g. `["APPFI"]` |
| `li_tipovoo` | Flight type | JSON array: `S` scheduled, `G` general aviation |
| `li_regravoo` | Flight rule | JSON array: `I` IFR, `V` VFR |
| `li_prnav` | P-RNAV capability | JSON array of 0/1 |
| `li_rvsm` | RVSM capability | JSON array of 0/1 |

Two conventions matter when reading these files:

**The five `li_*` columns are JSON arrays.** They cannot be written to a delimited file as
they stand, so the downloader collapses each to a pipe-separated string. A row with
`["APPFI","ACCCW"]` is stored as `APPFI|ACCCW`. Split them again on read when a row can
carry more than one value:

```r
d$orgaos <- strsplit(d$li_orgaos, "|", fixed = TRUE)
```

**`id` is not an identifier.** It is the callsign glued to the airport pair, so the same
`id` returns on every flight of that route — `PPBLVSBPASBSM` appears on 2019-03-12 and
again months later. De-duplicating on it would delete almost the entire dataset. `pk` is
the key, and the downloader uses it.

## Verifying the Contents

Downloading a file is not the same as having the data. This chunk opens each year file and
answers the three questions that decide whether it is usable: does it cover the whole year,
are the keys unique, and are the essential fields populated.

```{r}
#| label: tbl-bra-totalbr-contents

if (length(files) == 0) {
  knitr::kable(tibble::tibble(
    note = "No source files yet — nothing to verify."
  ))
} else {
  contents <- purrr::map(files, function(file) {
    # only the columns needed for the checks, so a large year file stays cheap
    d <- data.table::fread(
      file = file, sep = ";", colClasses = "character", na.strings = "",
      select = c(totalbr_id_col, totalbr_date_col, "co_addep", "co_addes"),
      showProgress = FALSE
    )
    day <- substr(d[[totalbr_date_col]], 1, 10)
    tibble::tibble(
      Year          = stringr::str_extract(basename(file), "[0-9]{4}"),
      Rows          = nrow(d),
      `Unique pk`   = dplyr::n_distinct(d[[totalbr_id_col]]),
      `Days`        = dplyr::n_distinct(day, na.rm = TRUE),
      `Months`      = dplyr::n_distinct(substr(day, 6, 7), na.rm = TRUE),
      `First day`   = min(day, na.rm = TRUE),
      `Last day`    = max(day, na.rm = TRUE),
      `No ADEP (%)` = round(100 * mean(is.na(d$co_addep) | d$co_addep == "AFIL"), 1),
      `No ADES (%)` = round(100 * mean(is.na(d$co_addes) | d$co_addes == "ZZZZ"), 1)
    )
  }) |> purrr::list_rbind()

  contents |>
    dplyr::mutate(
      Complete = dplyr::case_when(
        Months < 12 ~ paste0("NO — only ", Months, " month(s)"),
        Days   < 360 ~ paste0("NO — only ", Days, " day(s)"),
        TRUE ~ "yes"
      ),
      Duplicates = ifelse(Rows == `Unique pk`, "none",
                          paste0(Rows - `Unique pk`, " duplicate pk"))
    ) |>
    knitr::kable()
}
```

Read the table like this:

- **Complete** reading `NO` for a past year means the download stopped short. Re-run the
  download chunk — it resumes and fetches only the missing months. For the current year an
  incomplete count is expected and correct.
- **Duplicates** should read `none`. The downloader de-duplicates on `pk` when it merges
  the months, so anything else means a file was assembled by other means.
- **No ADEP / No ADES** counts `AFIL` and `ZZZZ`, which are legitimate values, not errors:
  a flight plan filed airborne has no departure aerodrome, and `ZZZZ` marks a destination
  that is not an ICAO aerodrome. They matter because any airport-level use of this table
  must decide what to do with them, and any count of "flights between two aerodromes" that
  ignores them is wrong.

## Monthly Coverage

The completeness check above is per year. This one is per month, which is where a partial
download actually shows up — a month with a handful of days is the signature of an
extraction that was cut off, and it is exactly what produced a bogus collapse in the ASMA
monthly trend before it was caught.

```{r}
#| label: tbl-bra-totalbr-monthly-coverage

if (length(files) == 0) {
  knitr::kable(tibble::tibble(note = "No source files yet — nothing to check."))
} else {
  purrr::map(files, function(file) {
    d <- data.table::fread(file = file, sep = ";", colClasses = "character",
                           na.strings = "", select = totalbr_date_col,
                           showProgress = FALSE)
    day <- substr(d[[1]], 1, 10)
    tibble::tibble(YEAR = substr(day, 1, 4), MONTH = substr(day, 6, 7), DAY = day)
  }) |>
    purrr::list_rbind() |>
    dplyr::filter(!is.na(MONTH), MONTH != "") |>
    dplyr::group_by(YEAR, MONTH) |>
    dplyr::summarise(MOVEMENTS = dplyr::n(),
                     DAYS = dplyr::n_distinct(DAY),
                     .groups = "drop") |>
    dplyr::mutate(
      # a month holding only a few days is a truncated extraction, not a quiet month
      STATUS = dplyr::case_when(
        DAYS >= 28 ~ "complete",
        DAYS >= 20 ~ "partial — check",
        TRUE       ~ "TRUNCATED — re-download"
      )
    ) |>
    knitr::kable()
}
```

A month flagged `TRUNCATED` is re-fetched by simply re-running the download chunk: the
engine judges a month by the days it holds, so a partial month is not treated as done.
Forcing a re-download of specific years is the heavier option:

```r
source(here::here("TOTALBR", "download_totalbr.R"))
download_totalbr(<the offending years>, out_dir = bra_totalbr_raw_dir, force = TRUE)
```

## Traffic Profile

A first look at what the table actually contains, once a year is complete: the mix of
flight rules and flight types, and the busiest aerodromes. This is the sanity check that
the file describes Brazilian traffic and not a filtered subset of it.

```{r}
#| label: tbl-bra-totalbr-profile

if (length(files) == 0) {
  knitr::kable(tibble::tibble(note = "No source files yet — nothing to profile."))
} else {
  prof <- purrr::map(files, function(file) {
    d <- data.table::fread(file = file, sep = ";", colClasses = "character",
                           na.strings = "", showProgress = FALSE,
                           select = c(totalbr_date_col, "li_regravoo", "li_tipovoo"))
    tibble::tibble(
      YEAR = substr(d[[totalbr_date_col]], 1, 4),
      # the li_* columns are pipe-separated after the CSV round-trip; the first
      # value is the flight's own rule/type, further values are later changes
      RULE = sub("\\|.*$", "", d$li_regravoo),
      TYPE = sub("\\|.*$", "", d$li_tipovoo)
    )
  }) |> purrr::list_rbind()

  prof |>
    dplyr::group_by(YEAR) |>
    dplyr::summarise(
      MOVEMENTS = dplyr::n(),
      `IFR (%)` = round(100 * mean(RULE == "I", na.rm = TRUE), 1),
      `VFR (%)` = round(100 * mean(RULE == "V", na.rm = TRUE), 1),
      `Scheduled (%)`  = round(100 * mean(TYPE == "S", na.rm = TRUE), 1),
      `General av. (%)` = round(100 * mean(TYPE == "G", na.rm = TRUE), 1),
      .groups = "drop"
    ) |>
    knitr::kable()
}
```

```{r}
#| label: tbl-bra-totalbr-study-airports
#| tbl-cap: "Movements at the twelve study airports (departures + arrivals)"

if (length(files) == 0) {
  knitr::kable(tibble::tibble(note = "No source files yet — nothing to summarise."))
} else {
  purrr::map(files, function(file) {
    d <- data.table::fread(file = file, sep = ";", colClasses = "character",
                           na.strings = "", showProgress = FALSE,
                           select = c(totalbr_date_col, "co_addep", "co_addes"))
    yr <- substr(d[[totalbr_date_col]], 1, 4)
    # one movement per end of the flight, so a domestic flight between two study
    # airports counts once at each — the same convention as an airport movement count
    dplyr::bind_rows(
      tibble::tibble(YEAR = yr, ICAO = d$co_addep, PHASE = "DEP"),
      tibble::tibble(YEAR = yr, ICAO = d$co_addes, PHASE = "ARR")
    )
  }) |>
    purrr::list_rbind() |>
    dplyr::filter(ICAO %in% bra_apts) |>
    dplyr::count(YEAR, ICAO, name = "MOVEMENTS") |>
    tidyr::pivot_wider(names_from = YEAR, values_from = MOVEMENTS, values_fill = 0) |>
    dplyr::left_join(bra_apts_names, by = "ICAO") |>
    dplyr::relocate(NAME, .after = ICAO) |>
    dplyr::arrange(ICAO) |>
    knitr::kable()
}
```

## Open Points

1. **What a "movement" is in this table.** Each row is a flight, with `co_addep` and
   `co_addes` as its two ends, so an airport count has to decide whether to count the
   flight once or once per end. The table above counts per end; confirm that this matches
   how DECEA reports movements before comparing the numbers to a published figure.
2. **`dh_eet` is stored as a timestamp, not a duration** (e.g. `2019-03-13T08:00:00` for a
   flight on 2019-03-12). It looks like an elapsed time encoded onto an arbitrary date;
   confirm the convention with ICEA before using it.
3. **`dt_dia` versus `dh_inicio`.** They are identical in every sampled row. Confirm they
   are meant to be, or which one is authoritative if they can differ — the month windows
   filter on `dt_dia`, and a divergence would put rows in the wrong month file.
4. **`co_matricula` and `ds_rota` are largely null** in the sample. Confirm whether they
   are populated for later years before any analysis depends on them.
5. **Harmonisation to APDF is not defined yet.** Unlike taxi-time and ASMA, this table has
   no metric attached; what it gets harmonised into depends on what it is used for
   (traffic volumes, an IFR/VFR split, a national denominator). That step is deliberately
   left out of this document until the use is fixed, so nothing here bakes in a wrong
   assumption.
