# Brazil TOTALBR Ingestion

```{r}
#| label: setup
#| include: false

# Libraries, project paths and parameters — including every TOTALBR setting
# (totalbr_data_years, the directories, the column names). Change the study
# period there, not here.
source(here::here("_chapter-setup.R"))
```

This note documents the preparation of the Brazilian **TOTALBR** files — the all-Brazil
movement table, ODIN table `total_brasil`. It follows the same shape as
`Taxi-BRA-ingestion.qmd` and `ASMA-BRA-ingestion.qmd`: get the data, check what arrived,
inventory it, harmonise it, and verify the coverage before anything downstream uses it.

TOTALBR differs from the other two in what it is *for*. Taxi-time and ASMA are performance
metrics computed at the twelve study airports; TOTALBR is **national traffic volume** — one
row per flight anywhere in Brazil, including aerodromes outside the study set and flights
with no aerodrome at all (`AFIL` departures, `ZZZZ` destinations). It carries no times to
subtract and no unimpeded reference, so there is no reference year and no percentile here.
It is the denominator and the context: how much traffic there was, where, and of what kind.

### Two sources, one dataset

TOTALBR reaches this project two ways, and both are first-class:

| Source | What it holds | Where |
| --- | --- | --- |
| **parquet archive** | the history, all airports, up to 2025 | `data-raw/totalbr/*.parquet` (or `BRA_TOTALBR_PARQUET`) |
| **ODIN API** | whatever the archive does not cover | downloaded to `data-raw/totalbr/totalbr_<year>.csv` |

Every check below reads both through `TOTALBR/totalbr_sources.R`, so a year is judged
covered by the days it actually holds regardless of which source holds them. That is what
keeps the download honest: years already in the archive are not fetched again, and the
API is asked only for the gap.

Nothing in the ingestion filters by airport. TOTALBR is the national table and is kept
whole — restricting it to the twelve study airports is a downstream decision, and doing it
on the way in would discard data that could only be recovered by downloading the year
again.

::: {.callout-important}
## Prerequisites to run this document
Neither the parquet archive nor the CSVs are shipped with the repository. Put the archive
in `data-raw/totalbr/` (or point `BRA_TOTALBR_PARQUET` at it if you keep it elsewhere —
it is large), and get the remaining years with the download chunk below. Until then, the
tables render with a short notice instead of failing.
:::

## Getting the Source Data

### Check the endpoint first

A full download runs for hours, so confirm the endpoint and its column names with a single
small request before starting one. This is cheap and catches a renamed column immediately
instead of after a wasted afternoon.

```{r}
#| label: probe-bra-totalbr-api
#| eval: false

source(here::here("ODIN", "download_odin.R"))
odin_probe(totalbr_table)      # prints the columns and flags the date-like ones
```

Run on 2026-07-30 this answered with 18 columns —
`pk, id, dt_dia, dh_inicio, dh_fim, co_indicativo, co_addep, co_addes, dh_eobt, dh_eet,
co_matricula, co_modelo, ds_rota, li_orgaos, li_tipovoo, li_regravoo, li_prnav, li_rvsm` —
of which five are date-like (`dt_dia`, `dh_inicio`, `dh_fim`, `dh_eobt`, `dh_eet`). That
confirms the two choices the downloader is built on: the month windows filter on `dt_dia`,
and the unique key is `pk`.

### Download

The chunk below is `eval: false` — the download is slow, so it is not executed on render;
run it manually when you need data.

It is **incremental and resumable**, exactly like the KPI08 downloader (they share the
engine in `ODIN/download_odin.R`): one month per request window, each month saved under
`parts/` the moment it arrives, and a month judged already-downloaded by **the days it
contains** rather than by a file existing. So an interrupted run keeps everything it had,
and a month fetched while it was still the current month is fetched again once it closes
instead of staying frozen half-full.

Years you already hold as files go straight into `data-raw/totalbr/` with the same naming
(`totalbr_2019.csv`); the downloader inspects what each covers and fetches only the gaps.

```{r}
#| label: download-bra-totalbr-data
#| eval: false

# sourcing only defines the functions; nothing is fetched until they are called
source(here::here("TOTALBR", "download_totalbr.R"))
source(here::here("TOTALBR", "totalbr_sources.R"))

# only the years the parquet archive does not already cover
need <- totalbr_missing_years(totalbr_data_years, raw_dir = bra_totalbr_raw_dir)
message("Years to fetch: ", if (length(need)) paste(need, collapse = ", ") else "none")

if (length(need) > 0)
  download_totalbr(need, out_dir = bra_totalbr_raw_dir)
```

To fetch one year regardless of what is on disk — which is the right way to test the
endpoint before trusting a long run — call the downloader directly:

```r
source(here::here("TOTALBR", "download_totalbr.R"))
download_totalbr(2026, out_dir = bra_totalbr_raw_dir)
```

A word on years that come back nearly empty. Asking for 2019 returns a few dozen rows
(14 in March, 32 in May) and "no rows" for every other month. That is the API answering
truthfully, not a broken filter: `total_brasil` holds only a token sample for the early
years. The history comes from the parquet archive, and the API is there for the present.
The rule to apply is the same either way — trust the day counts in the coverage tables
below, not the fact that a download ran.

## Checking What Arrived

```{r}
#| label: check-bra-totalbr-data

source(here::here("TOTALBR", "totalbr_sources.R"))

src <- totalbr_sources(bra_totalbr_raw_dir)

if (length(src$parquet) == 0 && length(src$csv) == 0) {
  knitr::kable(tibble::tibble(
    status = "no source data yet",
    `what to do` = paste0(
      "Put the parquet archive in ", bra_totalbr_raw_dir,
      " (or set BRA_TOTALBR_PARQUET), and/or run the ",
      "`download-bra-totalbr-data` chunk above."
    )
  ))
} else {
  # the day counts are computed once here and reused by every table below, so a
  # large archive is scanned once per render rather than five times
  totalbr_counts <- totalbr_day_counts(bra_totalbr_raw_dir, totalbr_date_col)

  knitr::kable(tibble::tibble(
    `parquet archive` = if (length(src$parquet) == 0) "none"
                        else paste(basename(src$parquet), collapse = ", "),
    `downloaded CSVs` = if (length(src$csv) == 0) "none"
                        else paste(sort(stringr::str_extract(basename(src$csv),
                                                             "[0-9]{4}")),
                                   collapse = ", "),
    `days held`       = dplyr::n_distinct(totalbr_counts$DATE),
    movements         = format(sum(totalbr_counts$MOVEMENTS), big.mark = ",")
  ))
}
```

## Source Inventory

```{r}
#| label: tbl-bra-totalbr-source-inventory
#| tbl-cap: "Inventory of the raw TOTALBR source files"

files <- c(src$parquet, src$csv)

if (length(files) == 0) {
  knitr::kable(tibble::tibble(
    note = paste0("No source files in ", bra_totalbr_raw_dir,
                  " yet. See the prerequisites above.")
  ))
} else {
  purrr::map(files, function(file) {
    tibble::tibble(
      File        = basename(file),
      Kind        = ifelse(grepl("\\.parquet$", file), "parquet archive",
                           "downloaded CSV"),
      Year        = tidyr::replace_na(
                      stringr::str_extract(basename(file), "[0-9]{4}"), "several"),
      `Size (MB)` = round(file.info(file)$size / 1024^2, 1),
      Modified    = format(file.info(file)$mtime, "%Y-%m-%d %H:%M")
    )
  }) |>
    purrr::list_rbind() |>
    dplyr::arrange(Kind, Year) |>
    knitr::kable()
}
```

The files are semicolon-delimited UTF-8, written by the downloader. The per-month parts it
fetched live under `data-raw/totalbr/parts/` and are what makes a re-run cheap; the year
file is the merge of them, de-duplicated on `pk` and sorted by `dt_dia`.

## What the Columns Mean

| Source field | Meaning | Note |
| --- | --- | --- |
| `pk` | Row hash | **The unique key.** De-duplication keys on this |
| `id` | Callsign + ADEP + ADES | e.g. `PPBLVSBPASBSM`. **Repeats** on every flight of that route — never a key |
| `dt_dia` | Movement day/time | What the month windows filter on |
| `dh_inicio` | Start of the movement | Equal to `dt_dia` in most sampled rows |
| `dh_fim` | End of the movement | Equal to `dh_inicio` when no duration was recorded |
| `co_indicativo` | Callsign | `PPBLV`, `AZU2429` |
| `co_addep` | Departure aerodrome | `AFIL` = flight plan filed airborne, i.e. no departure aerodrome |
| `co_addes` | Destination aerodrome | `ZZZZ` = destination not an ICAO aerodrome |
| `dh_eobt` | Estimated off-block time | From the flight plan |
| `dh_eet` | Estimated elapsed time | Encoded as a timestamp, not a duration — read with care |
| `co_matricula` | Registration | Largely null in the sample |
| `co_modelo` | Aircraft type | `SR22`, `AT76`; `XXXX` where unknown |
| `ds_rota` | Route | Largely null in the sample |
| `li_orgaos` | ATS units involved | JSON array, e.g. `["APPFI"]` |
| `li_tipovoo` | Flight type | JSON array: `S` scheduled, `G` general aviation |
| `li_regravoo` | Flight rule | JSON array: `I` IFR, `V` VFR |
| `li_prnav` | P-RNAV capability | JSON array of 0/1 |
| `li_rvsm` | RVSM capability | JSON array of 0/1 |

Two conventions matter when reading these files:

**The five `li_*` columns are JSON arrays.** They cannot be written to a delimited file as
they stand, so the downloader collapses each to a pipe-separated string. A row with
`["APPFI","ACCCW"]` is stored as `APPFI|ACCCW`. Split them again on read when a row can
carry more than one value:

```r
d$orgaos <- strsplit(d$li_orgaos, "|", fixed = TRUE)
```

**`id` is not an identifier.** It is the callsign glued to the airport pair, so the same
`id` returns on every flight of that route — `PPBLVSBPASBSM` appears on 2019-03-12 and
again months later. De-duplicating on it would delete almost the entire dataset. `pk` is
the key, and the downloader uses it.

## Verifying the Contents

Having a file is not the same as having the data. This table answers the question that
decides whether a year is usable — **how many of its days are actually present** — over
both sources at once, so a year split between the archive and a download is judged on the
union rather than on either half.

```{r}
#| label: tbl-bra-totalbr-contents

if (length(files) == 0) {
  knitr::kable(tibble::tibble(
    note = "No source files yet — nothing to verify."
  ))
} else {
  totalbr_year_coverage(counts = totalbr_counts) |>
    dplyr::transmute(
      Year        = YEAR,
      Sources     = SOURCES,
      Movements   = format(MOVEMENTS, big.mark = ","),
      `Days held` = DAYS_HELD,
      `Days possible` = DAYS_POSSIBLE,
      Complete = dplyr::if_else(
        COVERED, "yes",
        paste0("NO — ", DAYS_POSSIBLE - DAYS_HELD, " day(s) missing"))
    ) |>
    knitr::kable()
}
```

Read the table like this:

- **Days possible** is the whole year for a past year and the days elapsed so far for the
  current one, so 2026 reading fewer days than 2025 is expected, not a gap.
- **Complete** reading `NO` for a past year means that year is short. Re-run the download
  chunk: it fetches only the missing months, and it skips whatever the parquet archive
  already covers.
- **Sources** shows where each year came from. A year reading `parquet + csv` is held by
  both; the day count is de-duplicated across them, so the overlap is not double-counted.

## Monthly Coverage

The completeness check above is per year. This one is per month, which is where a partial
download actually shows up — a month with a handful of days is the signature of an
extraction that was cut off, and it is exactly what produced a bogus collapse in the ASMA
monthly trend before it was caught.

```{r}
#| label: tbl-bra-totalbr-monthly-coverage

if (length(files) == 0) {
  knitr::kable(tibble::tibble(note = "No source files yet — nothing to check."))
} else {
  totalbr_counts |>
    dplyr::group_by(YEAR, MONTH) |>
    dplyr::summarise(MOVEMENTS = sum(MOVEMENTS),
                     # distinct days: a month present in both sources counts once
                     DAYS = dplyr::n_distinct(DATE),
                     .groups = "drop") |>
    dplyr::mutate(
      # a month holding only a few days is a truncated extraction, not a quiet month
      STATUS = dplyr::case_when(
        DAYS >= 28 ~ "complete",
        DAYS >= 20 ~ "partial — check",
        TRUE       ~ "TRUNCATED — re-download"
      )
    ) |>
    knitr::kable()
}
```

A month flagged `TRUNCATED` is re-fetched by simply re-running the download chunk: the
engine judges a month by the days it holds, so a partial month is not treated as done.
Forcing a re-download of specific years is the heavier option:

```r
source(here::here("TOTALBR", "download_totalbr.R"))
download_totalbr(<the offending years>, out_dir = bra_totalbr_raw_dir, force = TRUE)
```

## Duplication Check

ICEA/DECEA have reported duplication in the ODIN data. The check applies the operational
definition, which is deliberately strict — two rows are the same movement only when they
could not be two different flights:

| Test | Rule | Why it is conclusive |
| --- | --- | --- |
| **same `pk`** | the same row hash twice | `pk` identifies the row; no judgement needed |
| **near-duplicate** | same `co_matricula`, same aerodrome pair, and `dh_inicio` **or** `dh_fim` within a few minutes | one airframe cannot be in one place twice at once |

Anything else with a different `pk` is **not** counted. Two rows sharing a callsign, a route
and a day are two flights until the registration and the clock say otherwise: an aircraft
can fly the same route several times a day, and a callsign can be reused by a different
airframe on the same day.

The tolerance is 10 minutes by default (`tol_min`). Both `dh_inicio` and `dh_fim` are tested
because two records of one movement can agree on when it ended while differing on when it
started.

Two details that decide whether the numbers mean anything. The `pk` test reads the
per-month parts in `data-raw/totalbr/parts/` as well, pooled together: the merge
de-duplicates on `pk`, so the year file can no longer show whether the API repeated a row,
and a `pk` repeated *across* two months — what an overlapping request window produces — is
invisible to a per-file check. And `dh_inicio`/`dh_fim` values at the epoch (1970-01-01,
which ODIN uses in place of an empty value) are read as missing, so they cannot pile up at
the start of time and look like a thousand simultaneous movements.

```{r}
#| label: tbl-bra-totalbr-duplicates
#| eval: false

source(here::here("TOTALBR", "check_totalbr_duplicates.R"))

if (length(files) == 0) {
  knitr::kable(tibble::tibble(note = "No source files yet — nothing to check."))
} else {
  # "all" here for the overview; read it per source before concluding anything
  dups <- check_totalbr_duplicates(source   = "all",
                                   raw_dir  = bra_totalbr_raw_dir,
                                   date_col = totalbr_date_col)
  if (nrow(dups) == 0) {
    knitr::kable(tibble::tibble(result = "No duplication found."))
  } else {
    dups |>
      dplyr::select(dplyr::any_of(c(
        Year = "YEAR", Rows = "ROWS",
        `Same pk (merged)` = "SAME_PK", `Same pk (parts)` = "SAME_PK_PARTS",
        `Near-duplicates` = "NEAR_DUP", `Near %` = "NEAR_PCT",
        `No registration` = "NO_REG", `No reg. %` = "NO_REG_PCT"))) |>
      knitr::kable(format.args = list(big.mark = ","))
  }
}
```

This chunk is `eval: false`, unlike the coverage tables above. Those aggregate inside arrow
and never bring the archive into memory; this one has to compare rows, so it pulls the
whole 1 GB archive plus the ~500 MB year files into R — minutes, not seconds. A document
that takes ten minutes to render stops being run, so this is a deliberate manual step.

Run it per source rather than with `"all"` unless you want the pooled figure: `"csv"` skips
the archive entirely and is the fast path. The reads now announce themselves and report
rows and elapsed seconds on finishing, so a long silence after `Reading totalbr.parquet ...`
is a read in progress, not a hang.

The two tests do not overlap. An exact `pk` repeat trivially satisfies the second rule as
well — same aircraft, same place, zero minutes apart — so the near test runs only on
pk-unique rows. `NEAR_DUP` is therefore duplication **beyond** what `SAME_PK` reports, and
the near examples show pairs with different `pk`, which is the case the rule exists to
find. Before this, the examples came back full of pairs sharing one `pk`: true duplicates,
but already counted, and useless as evidence of anything new.

`No registration` and `No time` are there so the near-duplicate figure is read for what it
covers. Rows with an empty `co_matricula`, or with no time of day on either stamp, cannot
be tested by this rule at all. A near-duplicate count of zero over a year that is 90%
untestable says nothing, and `TESTED_PCT` is what makes that visible.

```{r}
#| label: examples-bra-totalbr-duplicates
#| eval: false

# the near-duplicates: rows with DIFFERENT pk that the rule caught, next to the
# rows they repeat, with the gap in minutes. Read the SOURCE column: the function
# walks the sources and returns the first that has anything, so a result can come
# from the parquet archive even when the year you asked about was downloaded.
totalbr_duplicate_examples(2026, "near", raw_dir = bra_totalbr_raw_dir)

# the exact repeats: the same pk twice
totalbr_duplicate_examples(2026, "pk", raw_dir = bra_totalbr_raw_dir)

# EVERY flagged row, all columns, from ONE source — the list to inspect or send
dup26 <- totalbr_duplicate_rows(2026, "csv", raw_dir = bra_totalbr_raw_dir)

# ... and written out, semicolon-delimited like the raw files
totalbr_duplicate_rows(2026, "csv", raw_dir = bra_totalbr_raw_dir,
                       out_file = here::here("outputs", "totalbr-2026-duplicates.csv"))

# the API's own answer, with nothing from the archive mixed in
check_totalbr_duplicates(2026, "csv", raw_dir = bra_totalbr_raw_dir)

# widen or tighten what counts as "almost the same time"
check_totalbr_duplicates(2026, "csv", tol_min = 5,  raw_dir = bra_totalbr_raw_dir)
check_totalbr_duplicates(2026, "csv", tol_min = 30, raw_dir = bra_totalbr_raw_dir)
```

Run the examples before reporting anything. The gap in minutes is what turns a count into
evidence, and it is what ICEA need in order to find the records on their side.

`totalbr_duplicate_examples()` is a sample; `totalbr_duplicate_rows()` is the complete list
with every column, and it takes a **source** so the answer is about one producer only:

| `source` | Reads |
| --- | --- |
| `"csv"` | only the years downloaded from the API — the usual choice |
| `"parquet"` | only the handed-over archive |
| `"all"` | both |

That argument exists because mixing the two is how a clean 2026 download came to look
duplicated. The rows flagged then were the archive's, spilled into 2026-01-01, and no
amount of searching the downloaded file would have found them.

Each row comes back with `REASON` (`same pk` or `near`), a `PAIR` number shared by the rows
that belong together, `ROLE` (`repeat` or `original`) and `GAP_MIN`. So a repeat is always
next to the row it repeats, which is what makes the output usable as evidence rather than
as a list of record ids.

### The T that hid every timestamp

The first run of `check_totalbr_duplicates(2025, "csv")` came back with `NO_TIME_PCT` of
**100** — every downloaded row supposedly without a time — while `totalbr_midnight_share()`
put the same figure at 0.1%. Two measurements of one thing disagreeing by three orders of
magnitude is a bug, not a finding.

The CSVs store stamps as `2025-01-01T00:03:00`. `as.POSIXct()`'s default formats do **not**
include the ISO `T` separator: it falls through to the date-only format and returns
**midnight**, for every row, without an error or a warning. The midnight share was computed
on the text and was right; the duplicate check parsed to POSIXct and was wrong.

The consequence is that the near-duplicate test never compared anything on a downloaded
file. Any earlier "no duplicates" for a CSV source was the test skipping every row as
timeless, not a clean result. The parser now normalises the separator, and a file where
more than 90% of rows have no usable time raises a warning instead of reporting a
comfortable zero.

Two lessons that outlive this bug: a silent fallback is worse than an error, and the reason
this was caught at all is that the same quantity was measured twice by different routes. The
`TESTED_PCT` column existed precisely to make an untestable file visible — it earned its
place on the first run.

(It also showed `TESTED_PCT` of −41.5, which is impossible: the untestable classes overlap,
a row can lack both a registration and a time, and adding them double-counts. It is now the
union.)

### Result for the 2026 API download

Re-run this after pulling the parser fix — the earlier "none" was the bug, not an answer.



What does stand independently of the parser: filtering the downloaded 2026 file by hand for
the callsigns flagged earlier finds nothing — those records are the archive's, not the
API's. The `pk` test is also unaffected, since it never touches a timestamp.

Read any near-duplicate count next to `TESTED_PCT` from `check_totalbr_duplicates(2026, "csv")` before
concluding anything. If most of 2026 has no `co_matricula`, or no time of day on either
stamp, then "no duplicates" means "the rule could not be applied", not "there are none".
The two statements look identical in a summary table and are not the same finding.

So the duplication ICEA reported is in the handed-over archive, not in what the API serves
today — which is a different message to send back, and a better one.

### The two sources overlap at the year boundary

The archive labelled 2023–2025 carries rows dated **2026-01-01** in the first hours UTC —
the evening of 31 December in Brazil — so its extraction was cut at a local-time boundary
and spilled into the next year. Those same rows are in the downloaded 2026 file. Anything
reading both sources counts them twice.

It surfaced by accident: the near-duplicate examples came back full of pairs sharing one
`pk`, all stamped 2026-01-01 between 00:29 and 00:57, all long-haul international. The
uppercase hashes gave away which source they came from.

Which is the second half of the finding. **The row hash arrives uppercase from the archive
and lowercase from the API.** Compared as they come, the same row from the two sources
looks like two different rows, and no de-duplication on `pk` matches across them. Every
comparison here now folds the case first.

```{r}
#| label: tbl-bra-totalbr-overlap
#| eval: false

totalbr_source_overlap(raw_dir = bra_totalbr_raw_dir)
```

`IN_BOTH` above zero for a year means that year is held twice, and the movement counts for
it are inflated by that much. The fix is a decision about precedence — for the overlapping
window, keep the API rows or the archive rows, not both — and it belongs wherever the two
sources are combined for analysis, not in the ingestion.

### Why the earlier, looser tests were dropped

An earlier version of this check also counted rows sharing `id` + `dt_dia`, and rows sharing
callsign + route + day. Those are not duplication and the numbers were misleading — nearly
4% of the archive at one point. Two findings came out of chasing them, and both are about
field quality rather than duplication:

#### `dt_dia` does not resolve a movement

Pulling the rows showed the shape of it. A group flagged by the old, looser test:

```
pk        id      dt_dia               dh_inicio             dh_fim
6104BE1E… PPMFM…  2025-02-16 00:00:00  2025-02-16 20:56:03   2025-02-16 21:07:58
E2F1E8D2… PPMFM…  2025-02-16 00:00:00  2025-02-16 18:12:15   2025-02-16 18:12:19
82B9BD65… PPMFM…  2025-02-16 00:00:00  2025-02-16 18:46:24   2025-02-16 18:47:40
…ten rows, dh_inicio spanning 13:55 to 21:53
```

Ten **different** flights by one aircraft (`ZZZZ` → `ZZZZ`, local flying) on one day, each
with its own start and end, all sharing one `dt_dia`. Any key built on `dt_dia` collapses
them into one. They are not duplicates, and the strict rule agrees: their `dh_inicio` are
hours apart.

How widespread that is needs measuring rather than eyeballing, and the first attempt at it
was wrong. A histogram of `minute(dt_dia)` put 11% of rows at minute zero, which is **not**
the same as midnight — 13:00:00 and 20:00:00 are minute zero too. Measured properly,
`dt_dia` is a bare date on only about 0.1% of rows:

| Year | `dt_dia` at midnight | of rows |
| --- | --- | --- |
| 2023 | 1,932 | 1,818,076 |
| 2024 | 1,912 | 1,959,619 |
| 2025 | 1,747 | 2,109,623 |
| 2026 | 337 | 1,232,397 |

So the bare-date rows are real but far too few to explain 3.6% of a year. What the minute
histogram does show is a stamp **rounded to the hour** on a large share of rows, and a
stamp that resolves only to the hour cannot separate two movements of one aircraft inside
that hour any more than a bare date can separate them inside the day.

`totalbr_midnight_share()` now reports both — bare dates and on-the-hour stamps — for all
three time columns, and `totalbr_stamp_agreement()` answers the question underneath: how
often `dt_dia` is the movement time at all.

```{r}
#| label: tbl-bra-totalbr-stamps
#| eval: false

totalbr_midnight_share(raw_dir = bra_totalbr_raw_dir)   # per column, per year
totalbr_stamp_agreement(raw_dir = bra_totalbr_raw_dir)  # dt_dia vs dh_inicio
```

In the API payload `dt_dia` and `dh_inicio` are equal on every sampled row. Wherever they
diverge, `dt_dia` is something else — a filing time, a rounded stamp, a bare date — and is
not a movement time.

**The rule for the pipeline:** `dt_dia` is safe for the monthly download windows, because
even a bare date lands on the right day. It is not safe as a movement time, and neither
`dh_inicio` nor `dh_fim` can be assumed to carry a time of day without checking — the near
duplicate test excludes the rows where they do not and reports them as `NO_TIME`.

**The fallback is not confined to `dt_dia`.** Running the near-duplicate check on 2026
turned up rows where `dh_inicio` and `dh_fim` are also bare dates:

```
co_matricula co_indicativo co_addep co_addes dh_inicio            dh_fim               GAP_MIN
00403A       UAF1225       GMMX     SBEG     2026-06-27 00:00:00  2026-06-27 00:00:00       NA
00403A       UAF1225       GMMX     SBEG     2026-06-27 00:00:00  2026-06-27 00:00:00        0
1002         GAF922        SBBR     EDDB     2026-07-03 00:00:00  2026-07-03 00:00:00        0
```

A `GAP_MIN` of 0 there is not two records of one movement — it is two empty stamps. Left
alone, every such row of an aircraft on one day would be flagged as a duplicate of the
others, and the check would measure the missing clock rather than the traffic. Those rows
are excluded from the test and counted as `NO_TIME`; `TESTED_PCT` says what share of the
year the rule could actually judge.

Worth checking whether the zeroed rows are a particular class of traffic — the examples
above are foreign and military (`UAF1225` GMMX→SBEG, `GAF922` SBBR→EDDB), which would
suggest movements known to Brazil only as a flight plan, with no observed times. That would
make the absence expected rather than a fault.

#### `dh_eet`: correct in the API year, wrong in the 2019 sample

Read as an estimated arrival time, `dh_eet` minus `dh_eobt` is the estimated elapsed time.
In the 2026 download that comes out right. GLO2028, SBGL→SBFL:

```
dh_eobt              dh_fim               dh_eet
2026-01-01 00:03:00  2026-01-01 01:13:41  2026-01-01 01:18:00
```

Seventy-five minutes Rio to Florianópolis, against an actual `dh_fim` of 01:13 — the field
is doing exactly what its name says.

In the 2019 API sample it does not. There the difference from `dh_eobt` equals the EOBT's
**own time of day**, on every row sampled — 16:00 filed, 16h00 "elapsed"; 18:50 filed,
18h50 "elapsed" — which is not a flight a light single makes.

An earlier reading of this column concluded the field was broken outright, from a `summary()`
whose quartiles were all zero. That was wrong, and the way it was wrong is worth keeping:
rows where both stamps are a **bare date** give a difference of zero and pile up at the
centre of the distribution, so the median said "zero" about the empty rows rather than about
the field. A statistic over a column with a large timeless class describes the class, not
the column.

`totalbr_eet_check()` judges it the only way that means anything — by whether the implied
duration is a flight — and excludes the timeless rows instead of averaging over them:

```{r}
#| label: tbl-bra-totalbr-eet
#| eval: false

totalbr_eet_check(raw_dir = bra_totalbr_raw_dir)
```

`PLAUSIBLE_PCT` is the share landing between zero and twenty hours. High for a source means
`dh_eet` can be used there as an estimated arrival; low means it cannot, whatever the median
says.

Nothing here repairs anything, deliberately. De-duplicating a national movement table
changes every count computed from it, and which copy to keep is DECEA's call, not a choice
this pipeline should make silently.

### One cause we removed on our side first

Before measuring anyone else's duplication it was worth removing ours. The downloader
paginated with `order=dt_dia.asc`, and thousands of rows share a `dt_dia` — with ties, the
order between pages is whatever the database happens to produce per request, so a row can
come back on two consecutive pages (or on neither). The ordering now includes the unique
id, which makes the sort total and the pages a genuine partition of the window.

Re-downloading 2026 after the fix confirmed it: the same-`pk` count in the parts fell from
194 to 134 and the year gained exactly 60 rows. Sixty rows had been served twice, each in
place of one never served at all. What remains is the source's.

If your parts predate that fix, re-fetch a month with `force = TRUE` and compare.

## Archive vs API: What Differs

The parquet archive is the file used in the report — the golden — and the same years can now
be pulled from the ODIN API. `TOTALBR/compare_totalbr_sources.R` puts them side by side:
what matches, what exists on one side only, and where they disagree on rows they both hold.

### Match on what, exactly

The comparison is only as good as the key, and the obvious key is the wrong one to trust
blindly. If the two sources compute the row hash differently, **every row looks unique to
its own side** and a comparison on `pk` reports "nothing in common" about data that is in
fact identical. That is a fact about the identifier, not a difference in the traffic — and
it would be read as a catastrophic mismatch.

So there are three keys, and the first step is running all three:

| `key` | Matches on | Use it to |
| --- | --- | --- |
| `"pk"` | the row hash, case-folded | see whether the hash is comparable across sources at all |
| `"flight"` | callsign + aerodrome pair + `dh_inicio` to the minute | the real comparison |
| `"flight_day"` | the same, but only to the day | catch a systematic timezone shift: if this matches far better than `"flight"`, the stamps are offset |

```{r}
#| label: compare-bra-totalbr-sources
#| eval: false

source(here::here("TOTALBR", "compare_totalbr_sources.R"))

# run all three before believing any of them
for (k in c("pk", "flight", "flight_day"))
  print(compare_totalbr_sources(2023:2025, k, raw_dir = bra_totalbr_raw_dir))
```

`MATCH_PCT` is the share of the archive's rows found in the download. Reading the result:

- **`pk` low, `flight` high** — the hashes are source-specific. Drop `pk` as a cross-source
  key and compare on `flight`; nothing is wrong with the data.
- **`flight` low, `flight_day` high** — the two disagree on the *time*, not on the flights.
  Almost certainly a timezone: the archive cut at Brazilian local time, as its spill into
  2026-01-01 already suggested.
- **both low** — then the sources genuinely hold different traffic, and the examples below
  say which.

### When nothing matches, suspect the key

The first run said it plainly:

| Year | key | archive | download | matched |
| --- | --- | --- | --- | --- |
| 2024 | `pk` | 1,959,619 | 1,948,449 | **0** |
| 2024 | `flight` | 1,959,619 | 1,948,449 | **8** |
| 2025 | `flight` | 2,109,623 | 2,108,473 | 16 |

Row counts within 0.6% of each other and essentially nothing in common. Two files of the
same traffic cannot disagree on every record, so the key is wrong, not the data. `pk` at
zero was expected — the hash is computed per source. `flight` at eight is a different
problem, and guessing which part of it fails is a waste of a run:

```{r}
#| label: compare-bra-totalbr-diagnose
#| eval: false

compare_totalbr_diagnose(2024, raw_dir = bra_totalbr_raw_dir)
```

It reports the overlap of each piece of the key on its own — callsign, departure,
destination, calendar day — and then measures the clock difference directly: for every
downloaded flight it finds the **nearest** archive record of the same callsign and route
and reports how far apart they are. A rolling join, because a key built on time cannot
discover its own offset.

#### What it found

```
PIECE                       OVERLAP_PCT
co_indicativo                      89.5
co_addep                           99.7
co_addes                           99.7
callsign + route                   95.5
calendar day of dh_inicio          99.7
callsign + route + day             93.2

OFFSET_MIN        N
        50  1779079     <- 91% of all matched rows
        80     2271
        51     2070
```

The flights are the same: 93.2% share callsign, route and day. What differs is the clock,
by **exactly 50 minutes on 91% of rows**.

Fifty minutes is not a timezone — timezones are whole hours. A constant that clean, on
nearly every row, is a **definition difference**: the two sources stamp different moments of
the same flight in `dh_inicio`. In the API payload `dh_inicio` equals `dh_eobt` (the filed
off-block time); in the archive it is evidently something else.

That points at the way out. `dh_eobt` is copied from the flight plan rather than observed,
so if the two sources agree there, it can carry the comparison:

```{r}
#| label: compare-bra-totalbr-eobt
#| eval: false

compare_totalbr_sources(2024, "eobt", raw_dir = bra_totalbr_raw_dir)
compare_totalbr_fields(2024, key = "eobt", raw_dir = bra_totalbr_raw_dir)
```

`compare_totalbr_diagnose()` now measures the offset on `dh_eobt` as well as on
`dh_inicio`, so one run says whether that key will work before you spend anything on it.

Reading the offsets:

- **one dominant value on the hour, e.g. −180** — a timezone. Nothing is missing; one side
  needs shifting before any comparison.
- **one dominant value NOT on the hour, as here (50)** — the two sources mean different
  things by the field. Shifting would be a fudge; use a field they define the same way
  (`key = "eobt"`) and treat `dh_inicio` as a column that *differs*, which is a finding in
  its own right.
- **zero dominant, but a spread of a few minutes** — the two sources record slightly
  different moments for the same event. Match on `flight_day` and compare fields instead.
- **no callsign+route in common at all** — then the identifiers differ, not the clock, and
  the first table says which one.

#### Seeing which values are missing

An overlap of 89.5% on `co_indicativo` looks alarming and usually is not, because that is a
share of **distinct values**, not of traffic. Most distinct callsigns in TOTALBR are
general-aviation registrations flown a few times a year, so a tenth of the values can be a
three-hundredth of the flights.

```{r}
#| label: compare-bra-totalbr-missing-values
#| eval: false

compare_totalbr_missing_values("co_indicativo", 2024, raw_dir = bra_totalbr_raw_dir)
compare_totalbr_missing_values("co_addes", 2024, raw_dir = bra_totalbr_raw_dir)
```

It reports both shares — `VALUES_PCT` and `ROWS_PCT` — and then lists the one-sided values
**sorted by rows**, so anything that carries real traffic is at the top. What to look for:

- **`ROWS_PCT` near zero** — a long tail of rare callsigns. Ordinary, and no reason to
  distrust either source.
- **A scheduled airline callsign high in the list** — that is a real gap, and worth
  isolating: check whether it is one month, one aerodrome, or the whole year.
- **Values differing only in form** — trailing spaces, a zero-padded number, a different
  case. That is a formatting difference masquerading as missing data, and it would also be
  depressing every key built on the column.

#### The result, and why there is no minute-level key

| Year | key | archive | download | matched | only archive | only download |
| --- | --- | --- | --- | --- | --- | --- |
| 2024 | `pk` | 1,959,619 | 1,948,449 | **0** | — | — |
| 2024 | `flight_day` | 1,959,619 | 1,948,449 | **1,715,760** | 124,898 | 18,384 |
| 2025 | `flight_day` | 2,109,623 | 2,108,473 | 1,839,992 | 130,054 | 21,019 |

`flight_day` matches 93%. The minute-level key matched **eight rows in a year** and has been
removed: with `dh_inicio` stamped ~50 minutes apart by the two sources, a key finer than a
day cannot match, so it measured the offset rather than the traffic. The offset itself is
still reported, by `compare_totalbr_diagnose()`, which is where it belongs.

(2023 shows zero on the download side simply because that year has not been fetched from
the API yet.)

#### Zero pk matches: fact or bug?

Two sources holding 93% of the same flights and **no** identical hash is either a property
of the hash or a fault in this code, and no count can tell those apart. Look at the same
flight from both sides:

```{r}
#| label: compare-bra-totalbr-pk
#| eval: false

compare_totalbr_pk_examples(2024, raw_dir = bra_totalbr_raw_dir)
```

It first reports the **shape** of `pk` on each side — length range and whether every
character is hex — which catches a truncation or an encoding difference without reading a
single example. Then it prints the two `pk` values for the same flight, matched via
`flight_day`.

- **Two different 32-character hex strings** — the hash is computed per source, over
  different fields or with a different salt. Nothing to fix, and `pk` is simply not a
  cross-source identifier. Worth asking ICEA what it is computed over.
- **One truncated, padded, quoted, or empty** — that is a reading fault, and ours.

### The rows on each side

```{r}
#| label: compare-bra-totalbr-examples
#| eval: false

# in the archive, absent from the API
compare_totalbr_examples(2024, "only_parquet", key = "flight",
                         raw_dir = bra_totalbr_raw_dir)

# in the API, absent from the archive
compare_totalbr_examples(2024, "only_csv", key = "flight",
                         raw_dir = bra_totalbr_raw_dir)

# matched rows, one from each source, side by side
compare_totalbr_examples(2024, "both", key = "flight",
                         raw_dir = bra_totalbr_raw_dir)
```

Look at what the one-sided rows have in common before concluding the API lost data. A
missing set concentrated in one month, one aerodrome, or one flight type is a filter
difference; scattered singles are ordinary late-arriving records.

### Where they disagree on the same flight

```{r}
#| label: compare-bra-totalbr-fields
#| eval: false

compare_totalbr_fields(2024, raw_dir = bra_totalbr_raw_dir)

# to include ds_rota and the li_* arrays as well (holds every column of both
# sources in memory — several GB over three years)
compare_totalbr_fields(2024, all_cols = TRUE, raw_dir = bra_totalbr_raw_dir)
```

For rows both sources carry, this reports per column the share where they differ. It
separates a genuine disagreement (`DIFFERENT_PCT`) from a field one side simply does not
populate (`ONLY_PARQUET_PCT` / `ONLY_CSV_PCT`) — the second is a coverage difference and
needs a different answer from the first.

A row count that matches proves nothing about content: the same flight can be present on
both sides with a different registration, a different aircraft type, or a stamp minutes
apart. This is the table that finds that.

## Traffic Profile

A first look at what the table actually contains, once a year is complete: the mix of
flight rules and flight types, and the busiest aerodromes. This is the sanity check that
the file describes Brazilian traffic and not a filtered subset of it.

```{r}
#| label: tbl-bra-totalbr-profile

if (length(files) == 0) {
  knitr::kable(tibble::tibble(note = "No source files yet — nothing to profile."))
} else {
  prof <- totalbr_count_by(c("li_regravoo", "li_tipovoo"),
                           bra_totalbr_raw_dir, totalbr_date_col)

  if (nrow(prof) == 0) {
    knitr::kable(tibble::tibble(
      note = "Neither source exposes li_regravoo / li_tipovoo as plain columns."
    ))
  } else {
    prof |>
      # the li_* columns are pipe-separated after the CSV round-trip; the first
      # value is the flight's own rule/type, further values are later changes
      dplyr::mutate(RULE = sub("\\|.*$", "", li_regravoo),
                    TYPE = sub("\\|.*$", "", li_tipovoo)) |>
      dplyr::group_by(YEAR) |>
      dplyr::summarise(
        MOVEMENTS = sum(MOVEMENTS),
        `IFR (%)` = round(100 * sum(MOVEMENTS[RULE == "I"]) / sum(MOVEMENTS), 1),
        `VFR (%)` = round(100 * sum(MOVEMENTS[RULE == "V"]) / sum(MOVEMENTS), 1),
        `Scheduled (%)`   = round(100 * sum(MOVEMENTS[TYPE == "S"]) / sum(MOVEMENTS), 1),
        `General av. (%)` = round(100 * sum(MOVEMENTS[TYPE == "G"]) / sum(MOVEMENTS), 1),
        .groups = "drop"
      ) |>
      dplyr::mutate(MOVEMENTS = format(MOVEMENTS, big.mark = ",")) |>
      knitr::kable()
  }
}
```

```{r}
#| label: tbl-bra-totalbr-study-airports
#| tbl-cap: "Movements at the twelve study airports (departures + arrivals)"

if (length(files) == 0) {
  knitr::kable(tibble::tibble(note = "No source files yet — nothing to summarise."))
} else {
  # one movement per END of the flight, so a domestic flight between two study
  # airports counts once at each — the usual airport movement convention. The two
  # ends are counted separately and stacked, never joined.
  dep <- totalbr_count_by("co_addep", bra_totalbr_raw_dir, totalbr_date_col) |>
    dplyr::rename(ICAO = co_addep)
  arr <- totalbr_count_by("co_addes", bra_totalbr_raw_dir, totalbr_date_col) |>
    dplyr::rename(ICAO = co_addes)

  movements <- dplyr::bind_rows(dep, arr)

  if (nrow(movements) == 0) {
    knitr::kable(tibble::tibble(
      note = "Neither source exposes co_addep / co_addes."
    ))
  } else {
    movements |>
      dplyr::filter(ICAO %in% bra_apts) |>
      dplyr::group_by(YEAR, ICAO) |>
      dplyr::summarise(MOVEMENTS = sum(MOVEMENTS), .groups = "drop") |>
      tidyr::pivot_wider(names_from = YEAR, values_from = MOVEMENTS,
                         values_fill = 0) |>
      dplyr::left_join(bra_apts_names, by = "ICAO") |>
      dplyr::relocate(NAME, .after = ICAO) |>
      dplyr::arrange(ICAO) |>
      knitr::kable(format.args = list(big.mark = ","))
  }
}
```

The same two counts also say how much of the table is *not* airport-to-airport traffic:

```{r}
#| label: tbl-bra-totalbr-no-aerodrome

if (length(files) == 0 || !exists("movements") || nrow(movements) == 0) {
  knitr::kable(tibble::tibble(note = "No source files yet — nothing to summarise."))
} else {
  dplyr::bind_rows(
    dplyr::mutate(dep, END = "departure"),
    dplyr::mutate(arr, END = "destination")
  ) |>
    dplyr::group_by(YEAR, END) |>
    dplyr::summarise(
      # AFIL = flight plan filed airborne (no departure aerodrome);
      # ZZZZ = the end is not an ICAO aerodrome. Both are valid values.
      `No aerodrome (%)` = round(
        100 * sum(MOVEMENTS[is.na(ICAO) | ICAO %in% c("AFIL", "ZZZZ")]) /
          sum(MOVEMENTS), 1),
      .groups = "drop"
    ) |>
    tidyr::pivot_wider(names_from = END, values_from = `No aerodrome (%)`) |>
    knitr::kable()
}
```

`AFIL` and `ZZZZ` are legitimate values, not errors: a flight plan filed airborne has no
departure aerodrome, and `ZZZZ` marks an end that is not an ICAO aerodrome. They matter
because any count of "flights between two aerodromes" that ignores them is wrong, and any
airport-level use of this table has to decide what to do with them.

## Open Points

1. **What a "movement" is in this table.** Each row is a flight, with `co_addep` and
   `co_addes` as its two ends, so an airport count has to decide whether to count the
   flight once or once per end. The table above counts per end; confirm that this matches
   how DECEA reports movements before comparing the numbers to a published figure.
2. **`dh_eet` behaves differently by era.** In the 2026 download it is a sound estimated
   arrival time (GLO2028 SBGL→SBFL: 75 minutes, against an actual 71). In the 2019 sample
   the difference from `dh_eobt` is the EOBT's own time of day, which is not a flight.
   Some rows also carry 1970-01-01 as a null sentinel, so any code touching the column must
   exclude epoch dates rather than trust `na.rm`. Worth asking ICEA when the encoding
   changed, rather than reporting the field as broken.
3. **`dt_dia` versus `dh_inicio`.** They are equal on rows that carry a real time, but
   around 11% of rows have `dt_dia` at midnight while `dh_inicio` holds the actual movement
   time. Use `dh_inicio` for anything time-of-day; `dt_dia` is still safe for the month
   windows, because a midnight stamp still lands on the right day.
4. **`co_matricula` and `ds_rota` are largely null** in the sample. Confirm whether they
   are populated for later years before any analysis depends on them.
5. **Harmonisation to APDF is not defined yet.** Unlike taxi-time and ASMA, this table has
   no metric attached; what it gets harmonised into depends on what it is used for
   (traffic volumes, an IFR/VFR split, a national denominator). That step is deliberately
   left out of this document until the use is fixed, so nothing here bakes in a wrong
   assumption.
