# Brazil TOTALBR Ingestion

```{r}
#| label: setup
#| include: false

# Libraries, project paths and parameters — including every TOTALBR setting
# (totalbr_data_years, the directories, the column names). Change the study
# period there, not here.
source(here::here("_chapter-setup.R"))
```

This note documents the preparation of the Brazilian **TOTALBR** files — the all-Brazil
movement table, ODIN table `total_brasil`. It follows the same shape as
`Taxi-BRA-ingestion.qmd` and `ASMA-BRA-ingestion.qmd`: get the data, check what arrived,
inventory it, harmonise it, and verify the coverage before anything downstream uses it.

TOTALBR differs from the other two in what it is *for*. Taxi-time and ASMA are performance
metrics computed at the twelve study airports; TOTALBR is **national traffic volume** — one
row per flight anywhere in Brazil, including aerodromes outside the study set and flights
with no aerodrome at all (`AFIL` departures, `ZZZZ` destinations). It carries no times to
subtract and no unimpeded reference, so there is no reference year and no percentile here.
It is the denominator and the context: how much traffic there was, where, and of what kind.

### Two sources, one dataset

TOTALBR reaches this project two ways, and both are first-class:

| Source | What it holds | Where |
| --- | --- | --- |
| **parquet archive** | the history, all airports, up to 2025 | `data-raw/totalbr/*.parquet` (or `BRA_TOTALBR_PARQUET`) |
| **ODIN API** | whatever the archive does not cover | downloaded to `data-raw/totalbr/totalbr_<year>.csv` |

Every check below reads both through `TOTALBR/totalbr_sources.R`, so a year is judged
covered by the days it actually holds regardless of which source holds them. That is what
keeps the download honest: years already in the archive are not fetched again, and the
API is asked only for the gap.

Nothing in the ingestion filters by airport. TOTALBR is the national table and is kept
whole — restricting it to the twelve study airports is a downstream decision, and doing it
on the way in would discard data that could only be recovered by downloading the year
again.

::: {.callout-important}
## Prerequisites to run this document
Neither the parquet archive nor the CSVs are shipped with the repository. Put the archive
in `data-raw/totalbr/` (or point `BRA_TOTALBR_PARQUET` at it if you keep it elsewhere —
it is large), and get the remaining years with the download chunk below. Until then, the
tables render with a short notice instead of failing.
:::

## Getting the Source Data

### Check the endpoint first

A full download runs for hours, so confirm the endpoint and its column names with a single
small request before starting one. This is cheap and catches a renamed column immediately
instead of after a wasted afternoon.

```{r}
#| label: probe-bra-totalbr-api
#| eval: false

source(here::here("ODIN", "download_odin.R"))
odin_probe(totalbr_table)      # prints the columns and flags the date-like ones
```

Run on 2026-07-30 this answered with 18 columns —
`pk, id, dt_dia, dh_inicio, dh_fim, co_indicativo, co_addep, co_addes, dh_eobt, dh_eet,
co_matricula, co_modelo, ds_rota, li_orgaos, li_tipovoo, li_regravoo, li_prnav, li_rvsm` —
of which five are date-like (`dt_dia`, `dh_inicio`, `dh_fim`, `dh_eobt`, `dh_eet`). That
confirms the two choices the downloader is built on: the month windows filter on `dt_dia`,
and the unique key is `pk`.

### Download

The chunk below is `eval: false` — the download is slow, so it is not executed on render;
run it manually when you need data.

It is **incremental and resumable**, exactly like the KPI08 downloader (they share the
engine in `ODIN/download_odin.R`): one month per request window, each month saved under
`parts/` the moment it arrives, and a month judged already-downloaded by **the days it
contains** rather than by a file existing. So an interrupted run keeps everything it had,
and a month fetched while it was still the current month is fetched again once it closes
instead of staying frozen half-full.

Years you already hold as files go straight into `data-raw/totalbr/` with the same naming
(`totalbr_2019.csv`); the downloader inspects what each covers and fetches only the gaps.

::: {.callout-note}
## On a corporate network
If the API answers in a browser but R fails with `CONNECT tunnel failed, response 407` — or
the vaguer `Failure when receiving data from the peer` — the proxy is refusing an
unauthenticated request. The browser authenticates with the logged-in account and never says
so; curl, under `httr2`, does not try unless told how. Write the settings once:

```{r}
#| label: setup-env-proxy
#| eval: false

source(here::here("setup_renviron.R"))
setup_renviron()   # creates/updates .Renviron with BRA_PROXY, BRA_PROXY_AUTH, ...
```

and fill in, e.g. `BRA_PROXY=http://proxy.example.int:9512`, `BRA_PROXY_AUTH=ntlm`,
`BRA_PROXY_USERPWD=:`. On Windows, `netsh winhttp show proxy` shows the address; the bare
`:` authenticates as the logged-in account, so no password is stored anywhere. Restart R
afterwards — `.Renviron` is read at startup only.
:::

```{r}
#| label: download-bra-totalbr-data
#| eval: false

# sourcing only defines the functions; nothing is fetched until they are called
source(here::here("TOTALBR", "download_totalbr.R"))
source(here::here("TOTALBR", "totalbr_sources.R"))

# What is already held, year by year, BEFORE deciding anything. Printing this is
# the point: a year that is skipped and a year that was never looked at produce
# the same silence otherwise, and "it did not download 2024" then reads as a
# failure when it is the archive already covering it.
cov <- totalbr_year_coverage(totalbr_data_years, raw_dir = bra_totalbr_raw_dir)
print(cov, n = Inf)

# only the years nothing on disk covers yet
need <- totalbr_missing_years(totalbr_data_years, raw_dir = bra_totalbr_raw_dir)
message("Years to fetch: ", if (length(need)) paste(need, collapse = ", ") else "none")

if (length(need) > 0)
  download_totalbr(need, out_dir = bra_totalbr_raw_dir)
```

The coverage table printed first is what makes the decision auditable. `COVERED` is why a
year is not fetched: 2024 and 2025 are in the **parquet archive**, so `need` comes back as
2026 alone. That is the intended behaviour — the archive is the reference for the history,
and re-downloading a year it already holds would spend hours to produce a second, competing
copy of the same year.

Read `DAYS_HELD` against `DAYS_POSSIBLE` in that table before trusting it. A year marked
covered with noticeably fewer days than the calendar has is a hole in the archive, and the
filter would keep hiding it — the same failure mode that kept 2025-10-31 invisible in the
taxi source for as long as nobody counted days.

To fetch a year **regardless** of what is on disk — to test the endpoint before trusting a
long run, or to hold the API's own copy of a year beside the archive so the two can be
compared — call the downloader directly:

```r
source(here::here("TOTALBR", "download_totalbr.R"))

download_totalbr(2026, out_dir = bra_totalbr_raw_dir)        # one year
download_totalbr(2024:2025, out_dir = bra_totalbr_raw_dir)   # beside the archive
```

Nothing is overwritten: the download writes `totalbr_<year>.csv`, never the parquet, so a
year can be held by both sources and each one still says what its own producer served.

A word on years that come back nearly empty. Asking for 2019 returns a few dozen rows
(14 in March, 32 in May) and "no rows" for every other month. That is the API answering
truthfully, not a broken filter: `total_brasil` holds only a token sample for the early
years. The history comes from the parquet archive, and the API is there for the present.
The rule to apply is the same either way — trust the day counts in the coverage tables
below, not the fact that a download ran.

## Checking What Arrived

```{r}
#| label: check-bra-totalbr-data

source(here::here("TOTALBR", "totalbr_sources.R"))

src <- totalbr_sources(bra_totalbr_raw_dir)

if (length(src$parquet) == 0 && length(src$csv) == 0) {
  knitr::kable(tibble::tibble(
    status = "no source data yet",
    `what to do` = paste0(
      "Put the parquet archive in ", bra_totalbr_raw_dir,
      " (or set BRA_TOTALBR_PARQUET), and/or run the ",
      "`download-bra-totalbr-data` chunk above."
    )
  ))
} else {
  # the day counts are computed once here and reused by every table below, so a
  # large archive is scanned once per render rather than five times
  totalbr_counts <- totalbr_day_counts(bra_totalbr_raw_dir, totalbr_date_col)

  knitr::kable(tibble::tibble(
    `parquet archive` = if (length(src$parquet) == 0) "none"
                        else paste(basename(src$parquet), collapse = ", "),
    `downloaded CSVs` = if (length(src$csv) == 0) "none"
                        else paste(sort(stringr::str_extract(basename(src$csv),
                                                             "[0-9]{4}")),
                                   collapse = ", "),
    `days held`       = dplyr::n_distinct(totalbr_counts$DATE),
    movements         = format(sum(totalbr_counts$MOVEMENTS), big.mark = ",")
  ))
}
```

## Source Inventory

```{r}
#| label: tbl-bra-totalbr-source-inventory
#| tbl-cap: "Inventory of the raw TOTALBR source files"

files <- c(src$parquet, src$csv)

if (length(files) == 0) {
  knitr::kable(tibble::tibble(
    note = paste0("No source files in ", bra_totalbr_raw_dir,
                  " yet. See the prerequisites above.")
  ))
} else {
  purrr::map(files, function(file) {
    tibble::tibble(
      File        = basename(file),
      Kind        = ifelse(grepl("\\.parquet$", file), "parquet archive",
                           "downloaded CSV"),
      Year        = tidyr::replace_na(
                      stringr::str_extract(basename(file), "[0-9]{4}"), "several"),
      `Size (MB)` = round(file.info(file)$size / 1024^2, 1),
      Modified    = format(file.info(file)$mtime, "%Y-%m-%d %H:%M")
    )
  }) |>
    purrr::list_rbind() |>
    dplyr::arrange(Kind, Year) |>
    knitr::kable()
}
```

The files are semicolon-delimited UTF-8, written by the downloader. The per-month parts it
fetched live under `data-raw/totalbr/parts/` and are what makes a re-run cheap; the year
file is the merge of them, de-duplicated on `pk` and sorted by `dt_dia`.

## What the Columns Mean

| Source field | Meaning | Note |
| --- | --- | --- |
| `pk` | Row hash | **The unique key.** De-duplication keys on this |
| `id` | Callsign + ADEP + ADES | e.g. `PPBLVSBPASBSM`. **Repeats** on every flight of that route — never a key |
| `dt_dia` | Movement day/time | What the month windows filter on |
| `dh_inicio` | Start of the movement | Equal to `dt_dia` in most sampled rows |
| `dh_fim` | End of the movement | Equal to `dh_inicio` when no duration was recorded |
| `co_indicativo` | Callsign | `PPBLV`, `AZU2429` |
| `co_addep` | Departure aerodrome | `AFIL` = flight plan filed airborne, i.e. no departure aerodrome |
| `co_addes` | Destination aerodrome | `ZZZZ` = destination not an ICAO aerodrome |
| `dh_eobt` | Estimated off-block time | From the flight plan |
| `dh_eet` | Estimated elapsed time | Encoded as a timestamp, not a duration — read with care |
| `co_matricula` | Registration | Largely null in the sample |
| `co_modelo` | Aircraft type | `SR22`, `AT76`; `XXXX` where unknown |
| `ds_rota` | Route | Largely null in the sample |
| `li_orgaos` | ATS units involved | JSON array, e.g. `["APPFI"]` |
| `li_tipovoo` | Flight type | JSON array: `S` scheduled, `G` general aviation |
| `li_regravoo` | Flight rule | JSON array: `I` IFR, `V` VFR |
| `li_prnav` | P-RNAV capability | JSON array of 0/1 |
| `li_rvsm` | RVSM capability | JSON array of 0/1 |

Two conventions matter when reading these files:

**The five `li_*` columns are JSON arrays.** They cannot be written to a delimited file as
they stand, so the downloader collapses each to a pipe-separated string. A row with
`["APPFI","ACCCW"]` is stored as `APPFI|ACCCW`. Split them again on read when a row can
carry more than one value:

```r
d$orgaos <- strsplit(d$li_orgaos, "|", fixed = TRUE)
```

**`id` is not an identifier.** It is the callsign glued to the airport pair, so the same
`id` returns on every flight of that route — `PPBLVSBPASBSM` appears on 2019-03-12 and
again months later. De-duplicating on it would delete almost the entire dataset. `pk` is
the key, and the downloader uses it.

## Verifying the Contents

Having a file is not the same as having the data. This table answers the question that
decides whether a year is usable — **how many of its days are actually present** — over
both sources at once, so a year split between the archive and a download is judged on the
union rather than on either half.

```{r}
#| label: tbl-bra-totalbr-contents

if (length(files) == 0) {
  knitr::kable(tibble::tibble(
    note = "No source files yet — nothing to verify."
  ))
} else {
  totalbr_year_coverage(counts = totalbr_counts) |>
    dplyr::transmute(
      Year        = YEAR,
      Sources     = SOURCES,
      Movements   = format(MOVEMENTS, big.mark = ","),
      `Days held` = DAYS_HELD,
      `Days possible` = DAYS_POSSIBLE,
      Complete = dplyr::if_else(
        COVERED, "yes",
        paste0("NO — ", DAYS_POSSIBLE - DAYS_HELD, " day(s) missing"))
    ) |>
    knitr::kable()
}
```

Read the table like this:

- **Days possible** is the whole year for a past year and the days elapsed so far for the
  current one, so 2026 reading fewer days than 2025 is expected, not a gap.
- **Complete** reading `NO` for a past year means that year is short. Re-run the download
  chunk: it fetches only the missing months, and it skips whatever the parquet archive
  already covers.
- **Sources** shows where each year came from. A year reading `parquet + csv` is held by
  both; the day count is de-duplicated across them, so the overlap is not double-counted.

## Monthly Coverage

The completeness check above is per year. This one is per month, which is where a partial
download actually shows up — a month with a handful of days is the signature of an
extraction that was cut off, and it is exactly what produced a bogus collapse in the ASMA
monthly trend before it was caught.

```{r}
#| label: tbl-bra-totalbr-monthly-coverage

if (length(files) == 0) {
  knitr::kable(tibble::tibble(note = "No source files yet — nothing to check."))
} else {
  totalbr_counts |>
    dplyr::group_by(YEAR, MONTH) |>
    dplyr::summarise(MOVEMENTS = sum(MOVEMENTS),
                     # distinct days: a month present in both sources counts once
                     DAYS = dplyr::n_distinct(DATE),
                     .groups = "drop") |>
    dplyr::mutate(
      # a month holding only a few days is a truncated extraction, not a quiet month
      STATUS = dplyr::case_when(
        DAYS >= 28 ~ "complete",
        DAYS >= 20 ~ "partial — check",
        TRUE       ~ "TRUNCATED — re-download"
      )
    ) |>
    knitr::kable()
}
```

A month flagged `TRUNCATED` is re-fetched by simply re-running the download chunk: the
engine judges a month by the days it holds, so a partial month is not treated as done.
Forcing a re-download of specific years is the heavier option:

```r
source(here::here("TOTALBR", "download_totalbr.R"))
download_totalbr(<the offending years>, out_dir = bra_totalbr_raw_dir, force = TRUE)
```

## Two Validations, One Month at a Time

Everything from here on works on **one month**. That is not a limitation to apologise for, it
is the unit the checks are meant to run in: a month of the API download reads in about ten
seconds and a month of the archive in about three, against minutes and a gigabyte of memory
for a whole year. A check that takes ten minutes stops being run, and a check that is not run
tells you nothing.

Every function below therefore takes `month`, `totalbr_read_year()` included, so the filter
happens while reading rather than after a year is in memory. Leave it out and it judges the
whole year — worth doing once a year is closed, and not worth doing while working something
out.

The snippets below all start from **`d`**: the raw dataset. `totalbr_read_year()` returns the
rows as the source served them and `totalbr_normalise()` only fixes types — timestamps parsed
to POSIXct, `pk` to one case, because the archive writes it uppercase and the API lowercase.
Nothing is filtered, nothing is aggregated, no column is added. It is the material both
validations work on, and one month of it is about 180,000 rows against 1.2 million for the
year.

Two validations, in this order, because the second only makes sense once the first is done:

| | Test | Verdict | What to do |
| --- | --- | --- | --- |
| **1** | the same `pk` twice | one movement, recorded twice with different payloads | **collapse** the copies into one |
| **2** | several rows, one flight | each row is one unit's view of it | **merge** them into one |

Both end in a merge, and the difference is what has to be decided first. In validation 1 there
is nothing to decide: `pk` already asserts the rows are one movement, and measurement confirms
every identity column agrees. In validation 2 that is exactly what has to be established, from
proximity in time, the units, and the route.

Nothing in this pipeline is deleted. An earlier version dropped the second copy of a repeated
`pk` on the grounds that `pk` hashes the whole row — it does not, and the measurement is
below.

The functions, so the rest of this chapter can be read as a map:

| | Count them | See the rows | Resolve them |
| --- | --- | --- | --- |
| **1** | `check_totalbr_duplicates()` → `SAME_PK` | `totalbr_duplicate_examples(…, "pk")` | `totalbr_merge_duplicate_pk()` |
| **2** | `check_totalbr_duplicates()` → `NEAR_DUP` | `totalbr_duplicate_examples(…, "near")` | `totalbr_merge_flights(d, gap_min =)` |

In practice neither resolving function is called by hand: `totalbr_prepare()` runs
`totalbr_clean_units()`, then both, in that order, and reports what each step did.
`totalbr_drop_duplicate_pk()` still exists but is **not** the path — it is kept only to
reproduce the old behaviour, and it is what loses the 162 `dh_eet` values measured below.

## Validation 1 — The Same `pk` Twice

Nothing else: does a `pk` appear more than once? One number answers it, and no judgement is
needed to *find* them. What to do with them turned out to need a rule — see below.

```{r}
#| label: tbl-bra-totalbr-duplicates
#| eval: false

source(here::here("TOTALBR", "check_totalbr_duplicates.R"))

# one source at a time, one month at a time
dup <- check_totalbr_duplicates(2026, "csv", month = 1, raw_dir = bra_totalbr_raw_dir)

# this validation, and only this validation
dup |> dplyr::select(YEAR, ROWS, SAME_PK, dplyr::any_of("SAME_PK_PARTS"))

# the archive answers for itself
#check_totalbr_duplicates(2025, "parquet", month = 1, raw_dir = bra_totalbr_raw_dir)

# the whole year, when a year is closed and the wait is worth it
#check_totalbr_duplicates(2026, "csv", raw_dir = bra_totalbr_raw_dir)
```

`check_totalbr_duplicates()` returns the columns for both validations in one table, because
both are answered from one read and the read is the expensive part. Select what you are asking
about, or the two get read as one figure. The rest of the columns belong to validation 2.

- **`SAME_PK`** — the surplus copies in the year file: every one after the first. A count, not
  a verdict; what happens to them is the next subsection.
- **`SAME_PK_PARTS`** — the same test on the per-month files in `data-raw/totalbr/parts/`,
  pooled. The year file is built by de-duplicating on `pk`, so by the time it exists it can no
  longer show whether the API served a row twice — and a `pk` repeated *across* two months,
  which is what an overlapping request window produces, is invisible to a per-file check. This
  is where an exact repeat actually shows up.

Always name a **source**. The archive and the API are different producers with different
faults, and `"all"` pools them into a figure that describes neither. Each call is one pass over
the file, so ask for the source and the month you actually want rather than reading the same
file twice.

The rows themselves, when the count is not zero:

```{r}
#| label: examples-bra-totalbr-pk
#| eval: false

totalbr_duplicate_examples(2026, "pk", month = 1, raw_dir = bra_totalbr_raw_dir)
```

It has no `source` argument: it walks the sources and returns the **first that has anything**,
so **read the `SOURCE` column before concluding**. A month of the download with no repeats
falls through to the archive, whose rows spill into 2026-01-01 — which is exactly how a clean
2026 download once looked duplicated.

**One cause was ours, and it is fixed.** The downloader paginated with `order=dt_dia.asc`, and
thousands of rows share a `dt_dia`; with ties, the order between pages is whatever the database
happens to produce per request, so a row could come back on two consecutive pages, or on
neither. The ordering now includes the unique id, which makes the sort total and the pages a
genuine partition of the window. Re-downloading 2026 after the fix took the same-`pk` count in
the parts from 194 to 134 and the year gained exactly 60 rows: sixty rows had been served
twice, each in place of one never served at all. If your parts predate that fix, re-fetch a
month with `force = TRUE` and compare.

### The copies are not interchangeable

The obvious thing to do with a repeated `pk` is keep one and drop the rest, on the grounds
that `pk` is the row hash and the copies are therefore the same row. **That premise is false**,
and the archive says so plainly. All 455 repeated `pk` there produce 910 rows, and all 910 stay
distinct when every column is compared — two versions each, never identical.

What never differs is the movement's **identity**: `id`, `dt_dia`, `dh_inicio`, `dh_fim`,
`co_indicativo`, `co_addep`, `co_addes`. What differs is the **payload** — `li_orgaos` in 303
of the 455 pairs, `dh_eet` in 291:

```
li_orgaos     dh_eet
APPAN         NA
APPAN,APPAN   2023-03-02 21:20:00
```

So `pk` hashes the movement, not the row, and "keep the first copy" discards a filled field at
random. Measured over those 455 pairs:

| Column | Keeping the first copy | Collapsing the pair |
| --- | --- | --- |
| `dh_eet` filled | 291 | **453** |
| `co_matricula` filled | 124 | **127** |

162 estimated-arrival values, thrown away by an arbitrary choice of copy.

The rule is therefore the same one validation 2 uses, in its degenerate case — the identity is
already equal, so there is nothing to chain and nothing to judge: **union the `li_*` columns,
take the first filled value everywhere else**. Nothing is lost either way.

```{r}
#| label: merge-pk-bra-totalbr
#| eval: false

# prepare_totalbr.R sources the other three in order, so one line is enough --
# sourcing merge_totalbr_duplicates.R alone leaves totalbr_read_year() and
# totalbr_normalise() undefined, or worse, at whatever version the session
# already held
source(here::here("TOTALBR", "prepare_totalbr.R"))

# d is the RAW dataset: the rows as downloaded, with times parsed and pk
# case-normalised, nothing filtered and nothing aggregated
d <- totalbr_normalise(totalbr_read_year(2026, month = 1, source = "csv",
                                         raw_dir = bra_totalbr_raw_dir))
d <- totalbr_clean_units(d)          # canonical li_* lists, before any grouping
d <- totalbr_merge_duplicate_pk(d)   # adds N_PK_MERGED; deletes nothing

# the resolved month, written out — semicolon-delimited like the raw files.
# NEVER over a raw download's name: that file records what the API served.
data.table::fwrite(d, here::here("outputs", "totalbr-2026-01-pk-resolved.csv"),
                   sep = ";", na = "", quote = TRUE)
```

`N_PK_MERGED` says how many source rows each row stands for, so the written file carries its
own evidence: `N_PK_MERGED > 1` is exactly the set that had a repeated `pk`. Write it to
`outputs/`, never over the raw download — that file is the record of what the API served and
is not edited in place.

In practice you will not call this by hand: `totalbr_prepare()` runs it as step 3 and reports
how many rows it collapsed, and its `out_file` writes the fully resolved dataset in one go. `totalbr_drop_duplicate_pk()` still exists for comparing against the old
behaviour, and is no longer used by the pipeline.

The API download shows `SAME_PK` of zero for every month tested, so this is an archive
property so far. It is measured on the archive and applied to both, because a rule that keeps
information cannot be wrong on a source that never triggers it.

## Validation 2 — One Flight, Several Rows

Everything that is not an identical `pk` lands here, and here there **is** judgement to make.
Two rows can be the same movement recorded twice, or one flight handed from sector to sector,
or two genuine flights of the same service — and the answer is read off proximity in time,
the units that saw them, and the route.

A flight is recorded by **every unit that saw it** — APPSP, ACCCW, APPRJ, APPVT — and each
record covers part of what those units captured. A row where `dh_fim` equals `dh_inicio` is not
an intention and not a filed plan: it is the flight caught at a single instant.

A record is **not** one unit's view, though it is tempting to read it that way. Only about a
third of raw rows name a single unit; two thirds already name two to six, in both sources. So
a record is a *partial* aggregation, and merging unions the partials — it does not assemble
one row per unit.

What is a genuine fault is the **same** unit listed twice in one row, `APPAN,APPAN`: 1.1% of
the API's rows and 2.7% of the archive's. Several *different* units in one row are normal and
are all kept. The two sources do not even agree on the separator — comma in the archive, pipe
in the download — so `totalbr_clean_units()` imposes one canonical form before anything is
grouped: distinct, sorted, pipe-separated. `totalbr_prepare()` runs it as part of step 2, and
after it no `li_*` value repeats a token, before or after merging.

### Proximity: what the near test measures

The near test is the strict part of this validation. Two rows are the same movement only when
they could not be two different flights: same `co_indicativo`, same aerodrome pair, and
`dh_inicio` **or** `dh_fim` within a few minutes. Anything else is two flights until the clock
says otherwise — an aircraft can fly the same route several times a day.

Both stamps are tested because two records of one movement can agree on when it ended while
differing on when it started. Stamps at the epoch (1970-01-01, which ODIN uses in place of an
empty value) are read as missing, so they cannot pile up at the start of time and look like a
thousand simultaneous movements.

These are the columns of the same table that answer it:

- **`NEAR_DUP`** — rows the rule caught, split into `PLAN_DUP` and `REPEAT_DUP`. The test runs
  only on `pk`-unique rows, so this is duplication **beyond** validation 1, never the same
  rows counted twice.
- **`PLAN_DUP`** — a zero-duration row paired with a row of real duration, over a **wide**
  window (`plan_tol_min`, 180 min). **`REPEAT_DUP`** — two rows of the same shape a few minutes
  apart, narrow window (`tol_min`, 10 min); there the two records *are* meant to agree, so a
  real gap means two flights.
- **`PLAN_ROWS`** — every zero-duration row, paired or not.
- **`NO_TIME_PCT` and `TESTED_PCT`** — how much of the month the rule could look at. A count of
  zero over a month that is 90% untestable is not a result, and a `NO_TIME_PCT` near 100 is a
  parsing fault rather than a property of the data; the function says so on screen.

```{r}
#| label: examples-bra-totalbr-near
#| eval: false

dup |> dplyr::select(YEAR, ROWS, NEAR_DUP, PLAN_DUP, REPEAT_DUP, PLAN_ROWS, TESTED_PCT)

# the rows, with the group each flagged row belongs to
ex <- totalbr_duplicate_examples(2026, "near", month = 1, raw_dir = bra_totalbr_raw_dir)
View(ex)                          # or: print(ex, n = Inf, width = Inf)

# EVERY flagged row, all columns, from ONE named source — to inspect or to send
totalbr_duplicate_rows(2026, "csv", month = 1, raw_dir = bra_totalbr_raw_dir,
                       out_file = here::here("outputs", "totalbr-2026-01-duplicates.csv"))
```

`totalbr_duplicate_examples()` shows the whole group a flagged row belongs to, with `FLAGGED`
marking the rows the rule actually caught and the others there to be compared against; `n` is
the number of groups, not rows. It carries **`ds_rota` and `li_orgaos`**, because those are
what the judgement rests on: two records of one flight name different units over the same
route, two genuine flights do not.

`totalbr_duplicate_rows()` is the complete list rather than a sample. Each row carries `REASON`
(`same pk` or `near`), a `PAIR` number shared by the rows that belong together, `ROLE`
(`repeat` or `original`) and `GAP_MIN`, so a repeat always sits next to the row it repeats —
which is what makes it usable as evidence, and what ICEA need to find the records on their
side.

### Rebuilding the flight

The rows of one flight are found by **key plus time chaining**:

- same `co_indicativo`, `co_addep` and `co_addes`; **and**
- the record starts within `gap_min` of the **latest end already in the group**.

Chaining on the running end — not on the group's first record — is what matters. PRATC on
2026-01-01 was captured by APPSP at 05:43 and tracked by ACCCW, APPRJ and APPVT from 06:07 to
06:59: one flight, four records, joined because 05:43 and 06:07 are 24 minutes apart. Before
the chaining it came out as two flights, because the earlier matching was one-to-one — one
record answered for one partner, and the APPSP capture was left over as a phantom second
flight.


| Field | Value in the merged row |
| --- | --- |
| `dh_inicio`, `dt_dia` | from the **earliest** record |
| `dh_fim` | from the **latest** record |
| `li_orgaos` and the other `li_*` | the **union** of every record's values |
| `co_matricula`, aerodromes, everything else | the first value actually filled in — a blank never overwrites a value |

`li_orgaos` has to be the union: it is precisely what differs between the records, so keeping
one value would erase the units whose reports justified merging. Three columns are added —
`N_MERGED`, `SPAN_MIN` and `MERGED_PK` (every source `pk`, pipe-separated) — so a merged row
can always be traced back.

### Two ways to run it, and when to use which

The next two chunks do **exactly the same work**. The difference is whether it happens in one
call or in four, and that decides which one you want:

| | `merge-bra-totalbr` | `merge-bra-totalbr-steps` |
| --- | --- | --- |
| What it is | `totalbr_prepare()` — the four steps packaged | the same four steps, written out |
| Prints | the step table: rows read, `pk` collapsed, records merged, flights out | each function's own message |
| Writes a file | yes, `out_file=`, and it refuses a raw download's name | no |
| Lets you stop midway | no — you get the end | **yes** — `d` exists between steps |
| Use it for | the routine run, and anything whose numbers you will quote | inspecting, testing a different `gap_min`, debugging |

Everything is packaged for a reason: the order matters (`totalbr_clean_units()` before
anything is grouped, validation 1 before validation 2), and the step table is what lets you
say what a number cost rather than guess. Reach for the long form when you need the raw `d`
in your hands — to count something yourself, to look at `totalbr_flight_edges()` before it is
consumed, or to try two tolerances without re-reading the file.

**Both must be given the same `gap_min`, or their results are not comparable.** The default
lives in one place, `TOTALBR_GAP_MIN` (60 minutes) in `merge_totalbr_duplicates.R`, precisely
so the two cannot drift: an earlier version of this document inspected a month at 45 and wrote
the file at 60, and the two counts had no reason to agree.

Both validations are one call. Read the month, collapse the repeated `pk`, merge each flight's
records, and what comes back is the dataset to work on:

```{r}
#| label: merge-bra-totalbr
#| eval: false

source(here::here("TOTALBR", "prepare_totalbr.R"))

# the working unit: one month
jan <- totalbr_prepare(2026, month = 1, raw_dir = bra_totalbr_raw_dir)

# the whole year, when that is what is wanted
#flights <- totalbr_prepare(2026, raw_dir = bra_totalbr_raw_dir)

# both validations resolved, straight to a NEW file
# gap_min is not passed: the default is TOTALBR_GAP_MIN, the same value the line
# above used, so the file written matches the dataset just inspected
totalbr_prepare(2026, month = 1, raw_dir = bra_totalbr_raw_dir,
                out_file = here::here("outputs", "totalbr-2026-01-flights.csv"))

# check one aircraft end to end
jan |>
  dplyr::filter(co_indicativo == "CMP205") |>
  dplyr::select(co_indicativo , co_addep, co_addes, dh_inicio, dh_fim, ds_rota , SPAN_MIN, N_MERGED, li_orgaos)
```

It prints what each step did — rows read, rows dropped as a repeated `pk`, records absorbed
by merging, flights out — so the effect on any count is stated rather than guessed. The same
table comes back as `attr(jan, "steps")`. For PRATC in January 2026 it returns 13 flights,
one row each, including the `SDIM`–`SBVT` leg as a single 05:43–06:59 row carrying all four
units.

Renaming columns, filtering to the study airports, deriving metrics: all of that happens
**after** this, on the result. Nothing is written to disk unless `out_file` is given, and it
refuses a raw download's file name — a download records what the API served and is not edited
in place.

The individual steps stay available for when something needs to be looked at on its own:

```{r}
#| label: merge-bra-totalbr-steps
#| eval: false

# prepare_totalbr.R sources the other three in order, so one line is enough --
# sourcing merge_totalbr_duplicates.R alone leaves totalbr_read_year() and
# totalbr_normalise() undefined, or worse, at whatever version the session
# already held
source(here::here("TOTALBR", "prepare_totalbr.R"))

# d is the RAW dataset: the rows as downloaded, with times parsed and pk
# case-normalised, nothing filtered and nothing aggregated
d <- totalbr_normalise(totalbr_read_year(2026, month = 1, source = "csv",
                                         raw_dir = bra_totalbr_raw_dir))
d <- totalbr_clean_units(d)                        # canonical li_* lists
d <- totalbr_merge_duplicate_pk(d)                 # validation 1
merged <- totalbr_merge_flights(d)                 # validation 2, TOTALBR_GAP_MIN
```

### Every month, and the months joined

One month is the working unit; the study period is many of them. Two functions cover it, and
neither re-does a month already written:

```{r}
#| label: months-bra-totalbr
#| eval: false

source(here::here("TOTALBR", "prepare_totalbr.R"))

# every CLOSED month of the year, one file each, skipping what is already there
totalbr_prepare_months(2026, out_dir = here::here("outputs"),
                       raw_dir = bra_totalbr_raw_dir)

# just one month, when that is what changed
totalbr_prepare_months(2026, months = 2, raw_dir = bra_totalbr_raw_dir)

# the months read back as one dataset
flights <- totalbr_bind_months(out_dir = here::here("outputs"))
```

`totalbr_prepare_months()` stops at the **last closed month**: the current one is still
receiving rows, so a file written from it disagrees with the next run for no good reason.
Ask for it by number if you want it anyway. Months already written are skipped, so calling
this again after a month closes costs one month of work, not a year of it.

**`totalbr_bind_months()` is not `rbind`.** A flight that starts on 31 January at 23:40 and
lands on 1 February has its records split across two months: each month merged its own half,
and the flight appears twice, each time incomplete. The bind therefore runs the flight merge
once more over the joined data, which joins exactly those straddling records and leaves
everything else alone — the rule is the same one, so a flight that is already whole has
nothing within `gap_min` to merge with. It reports how many rows the boundary cost:

```
1,247,301 flight(s) in, 1,247,142 out: 159 straddled a month boundary.
```

`MERGED_PK` is unioned across that second pass and `N_MERGED` is recomputed from it, so both
keep meaning "the source rows behind this flight" rather than "what the last pass did".
`remerge = FALSE` concatenates without it, for when you want the months exactly as written.

**Choosing `gap_min`.** It is the shortest turnaround that still separates two real flights of
the same aircraft on the same route. Too wide and consecutive legs of a shuttle merge; too
narrow and one flight splits. Read it off the data rather than assuming:

```{r}
#| label: cluster-profile-bra-totalbr
#| eval: false

totalbr_cluster_profile(d)   # flights, merges and span at 15 … 180 minutes
```

The tolerance to avoid is the one where the span starts climbing while the flight count barely
moves: that is legs being swallowed, not records being joined.

::: {.callout-warning}
## Measured on the API download
The rules were read off the CSVs. The **parquet archive has not been analysed the same way**
and may record flights differently. Measure it first, and write the result to a new file: a
reference archive cannot be downloaded again.
:::

### What proximity cannot see

Proximity and chaining are different relations, and it is worth being explicit about it
because the two counts look comparable and are not.

`NEAR_DUP` looks for **simultaneity**: two records whose `dh_inicio` fall within `tol_min` of
each other. That is the right shape for a record served twice. Fragmentation by unit is
**sequential** — each record ends roughly where the next begins — so the fragments of one
flight are never near each other in that sense, whatever the tolerance. No setting of
`tol_min` reaches them. Only the merge does, because it chains a record's start against the
group's running **end**.

CMP486, MPTO→SBGL, as the API served it:

```
co_matricula  dh_inicio            dh_fim               li_orgaos
HP9926        2026-01-01 02:15:00  2026-01-01 04:03:02  ACCAZ
HP9928        2026-01-01 04:58:00  2026-01-01 06:32:00  ACCCW|ACCBS
<NA>          2026-01-01 07:01:33  2026-01-01 07:01:33  APPRJ
HP9922        2026-01-02 01:49:00  2026-01-02 06:35:00  ACCBS|APPRJ|ACCCW|ACCAZ
```

The first three rows are one flight. Two independent things say so: the span, 02:15 to
07:01:33, is **4h46**, and the units are **disjoint** — ACCAZ, then ACCCW and ACCBS, then
APPRJ, in the geographic order of the route. The fourth row is the control. It is the same
daily service on the next day, arriving as a **single row** of 4h46 carrying exactly those
four units. One flight, recorded two different ways on two consecutive days.

The near rule flagged only the third row, and not against the other two: rows one and two
start **163 minutes** apart, sixteen times `tol_min`. They were never candidates.

Two things follow.

**`co_matricula` does not identify an airframe here.** The same daily service carries HP9926,
HP9928, HP9922 and a blank across four records — the real Copa format is `HP-1717CMP`. The
merge keys on `co_indicativo` plus the aerodrome pair and never consults the registration,
which is what lets it work; but its first-value-filled rule then reports HP9926 for the merged
flight and drops the rest silently. Do not read the registration of a merged row as fact.

**This flight is why the default is 60.** Rows one and two are 55 minutes apart end to start.
At 60 the three become one row spanning 286.6 minutes, matching the control exactly; at 45 —
the earlier default — they came out as two flights plus a fragment.
`totalbr_cluster_profile()` does not settle it either way (the P99 span is flat from 45 to
120), so the case, not the profile, is the argument. `TOTALBR_GAP_MIN` in
`merge_totalbr_duplicates.R` now holds that decision for the whole pipeline; change it there
if a later case argues otherwise.

### Finding more of them

The merged result is itself the search: a flight built from several records, over several
units, is what to look at.

```{r}
#| label: fragmented-bra-totalbr
#| eval: false

source(here::here("TOTALBR", "prepare_totalbr.R"))

# d is the RAW dataset: the rows as downloaded, with times parsed and pk
# case-normalised, nothing filtered and nothing aggregated
d <- totalbr_normalise(totalbr_read_year(2026, month = 1, source = "csv",
                                         raw_dir = bra_totalbr_raw_dir))
m <- totalbr_merge_flights(totalbr_merge_duplicate_pk(totalbr_clean_units(d)))

frag <- m |>
  dplyr::filter(N_MERGED > 1) |>
  # disjoint units are the signature: one flight handed from sector to sector,
  # not one record served twice
  dplyr::mutate(N_ORGAOS = lengths(strsplit(li_orgaos, "|", fixed = TRUE))) |>
  dplyr::arrange(dplyr::desc(N_ORGAOS), dplyr::desc(SPAN_MIN)) |>
  dplyr::select(co_indicativo, co_addep, co_addes, dh_inicio, SPAN_MIN,
                N_MERGED, N_ORGAOS, li_orgaos, MERGED_PK)

View(frag)
```

January 2026 alone: **13,369** flights rebuilt from more than one record, **9,510** of them
spanning three or more units and **1,017** spanning five or more. The heaviest are long-haul
crossings of Brazilian airspace, which is what one would expect — the more sectors a flight
transits, the more records it arrives in:

```
co_indicativo  co_addep  co_addes  SPAN_MIN  N_MERGED  li_orgaos
KLM701         EHAM      SAEZ         295.0         2  ACCAO|ACCAZ|ACCBS|ACCCW|ACCRE|APPAN|APPFI
AFR411A        SAEZ      LFPG         274.9         2  ACCAO|ACCAZ|ACCBS|ACCCW|ACCRE|APPAN|APPFI
PTOLW          SBTV      SBBI         270.5         2  ACCBS|ACCCW|ACCRE|APPCT|APPPS|APPRJ|APPSP
```

`MERGED_PK` carries every source `pk`, so any case found here can be traced back to the exact
records — which is the form ICEA need it in.

Set against `NEAR_DUP` at 4,289 for the same month, these are not overlapping figures and must
not be added or compared. They count different faults.


## Traffic Profile

A first look at what the table actually contains, once a year is complete: the mix of
flight rules and flight types, and the busiest aerodromes. This is the sanity check that
the file describes Brazilian traffic and not a filtered subset of it.

```{r}
#| label: tbl-bra-totalbr-profile

if (length(files) == 0) {
  knitr::kable(tibble::tibble(note = "No source files yet — nothing to profile."))
} else {
  prof <- totalbr_count_by(c("li_regravoo", "li_tipovoo"),
                           bra_totalbr_raw_dir, totalbr_date_col)

  if (nrow(prof) == 0) {
    knitr::kable(tibble::tibble(
      note = "Neither source exposes li_regravoo / li_tipovoo as plain columns."
    ))
  } else {
    prof |>
      # the li_* columns are pipe-separated after the CSV round-trip; the first
      # value is the flight's own rule/type, further values are later changes
      dplyr::mutate(RULE = sub("\\|.*$", "", li_regravoo),
                    TYPE = sub("\\|.*$", "", li_tipovoo)) |>
      dplyr::group_by(YEAR) |>
      dplyr::summarise(
        MOVEMENTS = sum(MOVEMENTS),
        `IFR (%)` = round(100 * sum(MOVEMENTS[RULE == "I"]) / sum(MOVEMENTS), 1),
        `VFR (%)` = round(100 * sum(MOVEMENTS[RULE == "V"]) / sum(MOVEMENTS), 1),
        `Scheduled (%)`   = round(100 * sum(MOVEMENTS[TYPE == "S"]) / sum(MOVEMENTS), 1),
        `General av. (%)` = round(100 * sum(MOVEMENTS[TYPE == "G"]) / sum(MOVEMENTS), 1),
        .groups = "drop"
      ) |>
      dplyr::mutate(MOVEMENTS = format(MOVEMENTS, big.mark = ",")) |>
      knitr::kable()
  }
}
```

```{r}
#| label: tbl-bra-totalbr-study-airports
#| tbl-cap: "Movements at the twelve study airports (departures + arrivals)"

if (length(files) == 0) {
  knitr::kable(tibble::tibble(note = "No source files yet — nothing to summarise."))
} else {
  # one movement per END of the flight, so a domestic flight between two study
  # airports counts once at each — the usual airport movement convention. The two
  # ends are counted separately and stacked, never joined.
  dep <- totalbr_count_by("co_addep", bra_totalbr_raw_dir, totalbr_date_col) |>
    dplyr::rename(ICAO = co_addep)
  arr <- totalbr_count_by("co_addes", bra_totalbr_raw_dir, totalbr_date_col) |>
    dplyr::rename(ICAO = co_addes)

  movements <- dplyr::bind_rows(dep, arr)

  if (nrow(movements) == 0) {
    knitr::kable(tibble::tibble(
      note = "Neither source exposes co_addep / co_addes."
    ))
  } else {
    movements |>
      dplyr::filter(ICAO %in% bra_apts) |>
      dplyr::group_by(YEAR, ICAO) |>
      dplyr::summarise(MOVEMENTS = sum(MOVEMENTS), .groups = "drop") |>
      tidyr::pivot_wider(names_from = YEAR, values_from = MOVEMENTS,
                         values_fill = 0) |>
      dplyr::left_join(bra_apts_names, by = "ICAO") |>
      dplyr::relocate(NAME, .after = ICAO) |>
      dplyr::arrange(ICAO) |>
      knitr::kable(format.args = list(big.mark = ","))
  }
}
```

The same two counts also say how much of the table is *not* airport-to-airport traffic:

```{r}
#| label: tbl-bra-totalbr-no-aerodrome

if (length(files) == 0 || !exists("movements") || nrow(movements) == 0) {
  knitr::kable(tibble::tibble(note = "No source files yet — nothing to summarise."))
} else {
  dplyr::bind_rows(
    dplyr::mutate(dep, END = "departure"),
    dplyr::mutate(arr, END = "destination")
  ) |>
    dplyr::group_by(YEAR, END) |>
    dplyr::summarise(
      # AFIL = flight plan filed airborne (no departure aerodrome);
      # ZZZZ = the end is not an ICAO aerodrome. Both are valid values.
      `No aerodrome (%)` = round(
        100 * sum(MOVEMENTS[is.na(ICAO) | ICAO %in% c("AFIL", "ZZZZ")]) /
          sum(MOVEMENTS), 1),
      .groups = "drop"
    ) |>
    tidyr::pivot_wider(names_from = END, values_from = `No aerodrome (%)`) |>
    knitr::kable()
}
```

`AFIL` and `ZZZZ` are legitimate values, not errors: a flight plan filed airborne has no
departure aerodrome, and `ZZZZ` marks an end that is not an ICAO aerodrome. They matter
because any count of "flights between two aerodromes" that ignores them is wrong, and any
airport-level use of this table has to decide what to do with them.

## Open Points

1. **What a "movement" is in this table.** Each row is a flight, with `co_addep` and
   `co_addes` as its two ends, so an airport count has to decide whether to count the
   flight once or once per end. The table above counts per end; confirm that this matches
   how DECEA reports movements before comparing the numbers to a published figure.
2. **`dh_eet` behaves differently by era.** In the 2026 download it is a sound estimated
   arrival time (GLO2028 SBGL→SBFL: 75 minutes, against an actual 71). In the 2019 sample
   the difference from `dh_eobt` is the EOBT's own time of day, which is not a flight.
   Some rows also carry 1970-01-01 as a null sentinel, so any code touching the column must
   exclude epoch dates rather than trust `na.rm`. Worth asking ICEA when the encoding
   changed, rather than reporting the field as broken.
3. **`dt_dia` versus `dh_inicio`.** They are equal on rows that carry a real time, but
   around 11% of rows have `dt_dia` at midnight while `dh_inicio` holds the actual movement
   time. Use `dh_inicio` for anything time-of-day; `dt_dia` is still safe for the month
   windows, because a midnight stamp still lands on the right day.
4. **`co_matricula` and `ds_rota` are largely null** in the sample. Confirm whether they
   are populated for later years before any analysis depends on them.
5. **Harmonisation to APDF is not defined yet.** Unlike taxi-time and ASMA, this table has
   no metric attached; what it gets harmonised into depends on what it is used for
   (traffic volumes, an IFR/VFR split, a national denominator). That step is deliberately
   left out of this document until the use is fixed, so nothing here bakes in a wrong
   assumption.


## Function Reference

Every function this chapter uses, and what it is for. Four files, sourced in order by
`prepare_totalbr.R` — sourcing one of the later ones on its own leaves the earlier ones
undefined, or at whatever version the session happened to hold.

### The pipeline: `TOTALBR/prepare_totalbr.R`

| Function | What it does |
| --- | --- |
| `totalbr_prepare(years, month, gap_min, source, out_file)` | The whole cycle in one call: read → normalise → `totalbr_clean_units()` → validation 1 → validation 2. Prints the step table and returns it as `attr(x, "steps")`. Writes a file only when `out_file` is given, and refuses a raw download's name. |
| `totalbr_prepare_months(years, months, out_dir)` | Runs the above for every **closed** month, one file per month, skipping months already written. Re-runnable after a month closes. |
| `totalbr_bind_months(out_dir, remerge)` | The monthly files as one dataset, re-merging the flights that straddle a month boundary. `remerge = FALSE` concatenates as written. |

### Getting the data: `TOTALBR/download_totalbr.R` and `ODIN/download_odin.R`

| Function | What it does |
| --- | --- |
| `download_totalbr(years, out_dir, force)` | Downloads `total_brasil` into `totalbr_<year>.csv`, one month per request, resuming what is on disk. Years default to `totalbr_data_years`. |
| `odin_probe(table)` | One small request: prints the table's columns and flags the date-like ones. Run before trusting a long download. |
| `odin_tables()` | Lists every table the ODIN API exposes. |
| `odin_count(table, date_col, from, to)` | How many rows the API **has** in a window, without downloading them. Tells a download gap from a source gap. |

### Reading and coverage: `TOTALBR/totalbr_sources.R`

| Function | What it does |
| --- | --- |
| `totalbr_read_year(year, month, source, raw_dir)` | The rows of one year — or one month of it, filtered while reading. Builds the bounds in the column's own timezone, half-open, so no day is lost at either end. **This is where `d` comes from.** |
| `totalbr_sources(raw_dir)` | Which files exist: the parquet archive, the downloaded CSVs, the month parts. |
| `totalbr_year_coverage(years, raw_dir)` | Per year: days held against days possible, movements, which sources hold them, and whether it counts as covered. |
| `totalbr_missing_years(years, raw_dir)` | Just the years nothing on disk covers — what the download chunk asks for. |
| `totalbr_day_counts(raw_dir)` | Movements per day per source. The raw material of every coverage table here. |
| `totalbr_count_by(...)` | Counts grouped by any column, for profiling traffic. |
| `totalbr_year_span()`, `totalbr_year_edges()`, `totalbr_year_boundary_rows()` | The first and last rows of a year, for checking that a window did not cut a day. |
| `totalbr_schema_info(raw_dir)` | The arrow type of each time column **with its timezone** — how the spurious `Europe/Paris` label on the archive was found. |
| `totalbr_year_count(years, official, ...)` | Counts a year under several date definitions at once, to reconcile against an official figure. |
| `totalbr_stored_tz(ds, col)` | The timezone a parquet column is stored with. Use it to build bounds that match the file. |

### Measuring duplication: `TOTALBR/check_totalbr_duplicates.R`

| Function | What it does |
| --- | --- |
| `check_totalbr_duplicates(years, source, month)` | Both validations counted, per source: `SAME_PK`, `NEAR_DUP`, and the rows the tests could **not** be applied to, so a number is read for what it covers. |
| `totalbr_duplicate_examples(years, what, month)` | The rows themselves — `what = "pk"` for validation 1, `"near"` for validation 2 — each flagged row shown beside the records it was flagged against. |
| `totalbr_duplicate_rows(...)` | Every flagged row, all columns, from one named source: the file to inspect or to send to ICEA. |
| `totalbr_near_pairs(d, tol_min)` | The pairing itself: which rows the near test considers the same flight, with the gap in minutes. |
| `totalbr_normalise(d)` | Times parsed to POSIXct, `pk` to one case, the 1970 sentinel to `NA`. Everything else assumes this has run. |
| `totalbr_parse_time(v)` | One column of stamps to POSIXct, handling the ISO `T` separator and the epoch sentinel. |
| `totalbr_midnight_share()`, `totalbr_stamp_agreement()`, `totalbr_eet_check()` | Field-level diagnostics: how often `dt_dia` is midnight, whether the stamps agree, whether `dh_eet` behaves as an arrival estimate. |
| `totalbr_source_overlap()` | Where the archive and the download cover the same days. |
| `totalbr_drop_duplicate_pk(d)` | **Not the pipeline.** Deletes repeated `pk` instead of collapsing them — kept only to reproduce the old behaviour, and it loses 162 filled `dh_eet` values on the archive. |

### Resolving it: `TOTALBR/merge_totalbr_duplicates.R`

| Function | What it does |
| --- | --- |
| `totalbr_clean_units(d)` | One canonical form for the `li_*` lists: de-duplicated inside the row, sorted, pipe-separated. Run **before** anything is grouped — a unit listed twice makes every later count wrong, and the two sources do not agree on the separator. |
| `totalbr_merge_duplicate_pk(d)` | **Validation 1.** Rows sharing a `pk` collapsed into one: `li_*` unioned, first filled value elsewhere. Adds `N_PK_MERGED`. |
| `totalbr_merge_flights(d, gap_min)` | **Validation 2.** The records of one flight — same callsign and aerodrome pair, chained within `gap_min` of the group's latest end — become one row. Adds `N_MERGED`, `SPAN_MIN`, `MERGED_PK`, `ZERO_SPAN_ONLY`. |
| `totalbr_flight_edges(d, gap_min)` | The grouping alone, as pairs, without merging. For seeing *why* two records were joined. |
| `totalbr_merge_duplicates(d, pairs)` | The collapse itself, given any pairing. `totalbr_merge_flights()` is this plus the chaining. |
| `totalbr_cluster_profile(d, gaps)` | Flights, merges and span at several tolerances — how `gap_min` is chosen from the data. |
| `totalbr_gap_profile(pairs)` | The distribution of the gap between paired records, per kind. |
| `totalbr_merge_example(d, pairs, merged, callsign)` | One aircraft's source rows and merged result side by side. The check worth running before trusting a month. |

### Comparing the two sources: `TOTALBR/compare_totalbr_sources.R`

| Function | What it does |
| --- | --- |
| `compare_totalbr_sources(years, key)` | Archive against API download: what matches, what exists on one side only. |
| `totalbr_match_flights(years, tol_min)` | Nearest-neighbour one-to-one pairing between the sources, when the keys do not match exactly. |
| `totalbr_unmatched(years, key, out_file)` | Every unmatched row, with the key used, so the comparison is auditable. |
| `totalbr_lookup(callsign, day)` | One flight in both sources, side by side — is it absent, or present with a different stamp? |
| `totalbr_pk_overlap()` | Whether the same `pk` appears in both sources at all. |

### Constants worth knowing

| Constant | Where | Meaning |
| --- | --- | --- |
| `TOTALBR_GAP_MIN` | `merge_totalbr_duplicates.R` | 60 minutes — how far apart two records of one flight may be. One value for the whole pipeline. |
| `TOTALBR_UNION_COLS` | `merge_totalbr_duplicates.R` | The `li_*` columns that are unioned rather than picked. |
| `TOTALBR_NEAR_MIN` | `check_totalbr_duplicates.R` | The near test's tolerance, in minutes. |
| `TOTALBR_DUP_COLS` | `check_totalbr_duplicates.R` | The columns the checks need; a source missing one is skipped. |
| `TOTALBR_CONTEXT_COLS` | `check_totalbr_duplicates.R` | `ds_rota`, `li_orgaos` — read when present, for reading the result. |
