# Brazil TOTALBR Ingestion

```{r}
#| label: setup
#| include: false

# Libraries, project paths and parameters — including every TOTALBR setting
# (totalbr_data_years, the directories, the column names). Change the study
# period there, not here.
source(here::here("_chapter-setup.R"))
```

This note documents the preparation of the Brazilian **TOTALBR** files — the all-Brazil
movement table, ODIN table `total_brasil`. It follows the same shape as
`Taxi-BRA-ingestion.qmd` and `ASMA-BRA-ingestion.qmd`: get the data, check what arrived,
inventory it, harmonise it, and verify the coverage before anything downstream uses it.

TOTALBR differs from the other two in what it is *for*. Taxi-time and ASMA are performance
metrics computed at the twelve study airports; TOTALBR is **national traffic volume** — one
row per flight anywhere in Brazil, including aerodromes outside the study set and flights
with no aerodrome at all (`AFIL` departures, `ZZZZ` destinations). It carries no times to
subtract and no unimpeded reference, so there is no reference year and no percentile here.
It is the denominator and the context: how much traffic there was, where, and of what kind.

### Two sources, one dataset

TOTALBR reaches this project two ways, and both are first-class:

| Source | What it holds | Where |
| --- | --- | --- |
| **parquet archive** | the history, all airports, up to 2025 | `data-raw/totalbr/*.parquet` (or `BRA_TOTALBR_PARQUET`) |
| **ODIN API** | whatever the archive does not cover | downloaded to `data-raw/totalbr/totalbr_<year>.csv` |

Every check below reads both through `TOTALBR/totalbr_sources.R`, so a year is judged
covered by the days it actually holds regardless of which source holds them. That is what
keeps the download honest: years already in the archive are not fetched again, and the
API is asked only for the gap.

Nothing in the ingestion filters by airport. TOTALBR is the national table and is kept
whole — restricting it to the twelve study airports is a downstream decision, and doing it
on the way in would discard data that could only be recovered by downloading the year
again.

::: {.callout-important}
## Prerequisites to run this document
Neither the parquet archive nor the CSVs are shipped with the repository. Put the archive
in `data-raw/totalbr/` (or point `BRA_TOTALBR_PARQUET` at it if you keep it elsewhere —
it is large), and get the remaining years with the download chunk below. Until then, the
tables render with a short notice instead of failing.
:::

## Getting the Source Data

### Check the endpoint first

A full download runs for hours, so confirm the endpoint and its column names with a single
small request before starting one. This is cheap and catches a renamed column immediately
instead of after a wasted afternoon.

```{r}
#| label: probe-bra-totalbr-api
#| eval: false

source(here::here("ODIN", "download_odin.R"))
odin_probe(totalbr_table)      # prints the columns and flags the date-like ones
```

Run on 2026-07-30 this answered with 18 columns —
`pk, id, dt_dia, dh_inicio, dh_fim, co_indicativo, co_addep, co_addes, dh_eobt, dh_eet,
co_matricula, co_modelo, ds_rota, li_orgaos, li_tipovoo, li_regravoo, li_prnav, li_rvsm` —
of which five are date-like (`dt_dia`, `dh_inicio`, `dh_fim`, `dh_eobt`, `dh_eet`). That
confirms the two choices the downloader is built on: the month windows filter on `dt_dia`,
and the unique key is `pk`.

### Download

The chunk below is `eval: false` — the download is slow, so it is not executed on render;
run it manually when you need data.

It is **incremental and resumable**, exactly like the KPI08 downloader (they share the
engine in `ODIN/download_odin.R`): one month per request window, each month saved under
`parts/` the moment it arrives, and a month judged already-downloaded by **the days it
contains** rather than by a file existing. So an interrupted run keeps everything it had,
and a month fetched while it was still the current month is fetched again once it closes
instead of staying frozen half-full.

Years you already hold as files go straight into `data-raw/totalbr/` with the same naming
(`totalbr_2019.csv`); the downloader inspects what each covers and fetches only the gaps.

```{r}
#| label: download-bra-totalbr-data
#| eval: false

# sourcing only defines the functions; nothing is fetched until they are called
source(here::here("TOTALBR", "download_totalbr.R"))
source(here::here("TOTALBR", "totalbr_sources.R"))

# only the years the parquet archive does not already cover
need <- totalbr_missing_years(totalbr_data_years, raw_dir = bra_totalbr_raw_dir)
message("Years to fetch: ", if (length(need)) paste(need, collapse = ", ") else "none")

if (length(need) > 0)
  download_totalbr(need, out_dir = bra_totalbr_raw_dir)
```

To fetch one year regardless of what is on disk — which is the right way to test the
endpoint before trusting a long run — call the downloader directly:

```r
source(here::here("TOTALBR", "download_totalbr.R"))
download_totalbr(2026, out_dir = bra_totalbr_raw_dir)
```

A word on years that come back nearly empty. Asking for 2019 returns a few dozen rows
(14 in March, 32 in May) and "no rows" for every other month. That is the API answering
truthfully, not a broken filter: `total_brasil` holds only a token sample for the early
years. The history comes from the parquet archive, and the API is there for the present.
The rule to apply is the same either way — trust the day counts in the coverage tables
below, not the fact that a download ran.

## Checking What Arrived

```{r}
#| label: check-bra-totalbr-data

source(here::here("TOTALBR", "totalbr_sources.R"))

src <- totalbr_sources(bra_totalbr_raw_dir)

if (length(src$parquet) == 0 && length(src$csv) == 0) {
  knitr::kable(tibble::tibble(
    status = "no source data yet",
    `what to do` = paste0(
      "Put the parquet archive in ", bra_totalbr_raw_dir,
      " (or set BRA_TOTALBR_PARQUET), and/or run the ",
      "`download-bra-totalbr-data` chunk above."
    )
  ))
} else {
  # the day counts are computed once here and reused by every table below, so a
  # large archive is scanned once per render rather than five times
  totalbr_counts <- totalbr_day_counts(bra_totalbr_raw_dir, totalbr_date_col)

  knitr::kable(tibble::tibble(
    `parquet archive` = if (length(src$parquet) == 0) "none"
                        else paste(basename(src$parquet), collapse = ", "),
    `downloaded CSVs` = if (length(src$csv) == 0) "none"
                        else paste(sort(stringr::str_extract(basename(src$csv),
                                                             "[0-9]{4}")),
                                   collapse = ", "),
    `days held`       = dplyr::n_distinct(totalbr_counts$DATE),
    movements         = format(sum(totalbr_counts$MOVEMENTS), big.mark = ",")
  ))
}
```

## Source Inventory

```{r}
#| label: tbl-bra-totalbr-source-inventory
#| tbl-cap: "Inventory of the raw TOTALBR source files"

files <- c(src$parquet, src$csv)

if (length(files) == 0) {
  knitr::kable(tibble::tibble(
    note = paste0("No source files in ", bra_totalbr_raw_dir,
                  " yet. See the prerequisites above.")
  ))
} else {
  purrr::map(files, function(file) {
    tibble::tibble(
      File        = basename(file),
      Kind        = ifelse(grepl("\\.parquet$", file), "parquet archive",
                           "downloaded CSV"),
      Year        = tidyr::replace_na(
                      stringr::str_extract(basename(file), "[0-9]{4}"), "several"),
      `Size (MB)` = round(file.info(file)$size / 1024^2, 1),
      Modified    = format(file.info(file)$mtime, "%Y-%m-%d %H:%M")
    )
  }) |>
    purrr::list_rbind() |>
    dplyr::arrange(Kind, Year) |>
    knitr::kable()
}
```

The files are semicolon-delimited UTF-8, written by the downloader. The per-month parts it
fetched live under `data-raw/totalbr/parts/` and are what makes a re-run cheap; the year
file is the merge of them, de-duplicated on `pk` and sorted by `dt_dia`.

## What the Columns Mean

| Source field | Meaning | Note |
| --- | --- | --- |
| `pk` | Row hash | **The unique key.** De-duplication keys on this |
| `id` | Callsign + ADEP + ADES | e.g. `PPBLVSBPASBSM`. **Repeats** on every flight of that route — never a key |
| `dt_dia` | Movement day/time | What the month windows filter on |
| `dh_inicio` | Start of the movement | Equal to `dt_dia` in most sampled rows |
| `dh_fim` | End of the movement | Equal to `dh_inicio` when no duration was recorded |
| `co_indicativo` | Callsign | `PPBLV`, `AZU2429` |
| `co_addep` | Departure aerodrome | `AFIL` = flight plan filed airborne, i.e. no departure aerodrome |
| `co_addes` | Destination aerodrome | `ZZZZ` = destination not an ICAO aerodrome |
| `dh_eobt` | Estimated off-block time | From the flight plan |
| `dh_eet` | Estimated elapsed time | Encoded as a timestamp, not a duration — read with care |
| `co_matricula` | Registration | Largely null in the sample |
| `co_modelo` | Aircraft type | `SR22`, `AT76`; `XXXX` where unknown |
| `ds_rota` | Route | Largely null in the sample |
| `li_orgaos` | ATS units involved | JSON array, e.g. `["APPFI"]` |
| `li_tipovoo` | Flight type | JSON array: `S` scheduled, `G` general aviation |
| `li_regravoo` | Flight rule | JSON array: `I` IFR, `V` VFR |
| `li_prnav` | P-RNAV capability | JSON array of 0/1 |
| `li_rvsm` | RVSM capability | JSON array of 0/1 |

Two conventions matter when reading these files:

**The five `li_*` columns are JSON arrays.** They cannot be written to a delimited file as
they stand, so the downloader collapses each to a pipe-separated string. A row with
`["APPFI","ACCCW"]` is stored as `APPFI|ACCCW`. Split them again on read when a row can
carry more than one value:

```r
d$orgaos <- strsplit(d$li_orgaos, "|", fixed = TRUE)
```

**`id` is not an identifier.** It is the callsign glued to the airport pair, so the same
`id` returns on every flight of that route — `PPBLVSBPASBSM` appears on 2019-03-12 and
again months later. De-duplicating on it would delete almost the entire dataset. `pk` is
the key, and the downloader uses it.

## Verifying the Contents

Having a file is not the same as having the data. This table answers the question that
decides whether a year is usable — **how many of its days are actually present** — over
both sources at once, so a year split between the archive and a download is judged on the
union rather than on either half.

```{r}
#| label: tbl-bra-totalbr-contents

if (length(files) == 0) {
  knitr::kable(tibble::tibble(
    note = "No source files yet — nothing to verify."
  ))
} else {
  totalbr_year_coverage(counts = totalbr_counts) |>
    dplyr::transmute(
      Year        = YEAR,
      Sources     = SOURCES,
      Movements   = format(MOVEMENTS, big.mark = ","),
      `Days held` = DAYS_HELD,
      `Days possible` = DAYS_POSSIBLE,
      Complete = dplyr::if_else(
        COVERED, "yes",
        paste0("NO — ", DAYS_POSSIBLE - DAYS_HELD, " day(s) missing"))
    ) |>
    knitr::kable()
}
```

Read the table like this:

- **Days possible** is the whole year for a past year and the days elapsed so far for the
  current one, so 2026 reading fewer days than 2025 is expected, not a gap.
- **Complete** reading `NO` for a past year means that year is short. Re-run the download
  chunk: it fetches only the missing months, and it skips whatever the parquet archive
  already covers.
- **Sources** shows where each year came from. A year reading `parquet + csv` is held by
  both; the day count is de-duplicated across them, so the overlap is not double-counted.

## Monthly Coverage

The completeness check above is per year. This one is per month, which is where a partial
download actually shows up — a month with a handful of days is the signature of an
extraction that was cut off, and it is exactly what produced a bogus collapse in the ASMA
monthly trend before it was caught.

```{r}
#| label: tbl-bra-totalbr-monthly-coverage

if (length(files) == 0) {
  knitr::kable(tibble::tibble(note = "No source files yet — nothing to check."))
} else {
  totalbr_counts |>
    dplyr::group_by(YEAR, MONTH) |>
    dplyr::summarise(MOVEMENTS = sum(MOVEMENTS),
                     # distinct days: a month present in both sources counts once
                     DAYS = dplyr::n_distinct(DATE),
                     .groups = "drop") |>
    dplyr::mutate(
      # a month holding only a few days is a truncated extraction, not a quiet month
      STATUS = dplyr::case_when(
        DAYS >= 28 ~ "complete",
        DAYS >= 20 ~ "partial — check",
        TRUE       ~ "TRUNCATED — re-download"
      )
    ) |>
    knitr::kable()
}
```

A month flagged `TRUNCATED` is re-fetched by simply re-running the download chunk: the
engine judges a month by the days it holds, so a partial month is not treated as done.
Forcing a re-download of specific years is the heavier option:

```r
source(here::here("TOTALBR", "download_totalbr.R"))
download_totalbr(<the offending years>, out_dir = bra_totalbr_raw_dir, force = TRUE)
```

## Duplication Check

ICEA/DECEA have reported that the ODIN data contains duplication. "Duplicate" means four
different things here, with different causes and different consequences, so they are
measured separately — a single number would hide which problem is present.

| Level | Key | What it means |
| --- | --- | --- |
| `key` | `pk` | the same row returned twice. `pk` is the row hash, so this is a delivery fault |
| `record` | `id` + the three timestamps | the same movement stored twice under **different** `pk` — the source does not recognise them as the same row |
| `movement` | `id` + `dt_dia` | the same flight at the same timestamp, other fields free to differ |
| `daily` | callsign + route + calendar day | **candidates only** — an aircraft can genuinely fly a route twice a day |

The `key` level is read from the per-month parts in `data-raw/totalbr/parts/`, not from the
year file. The merge de-duplicates on `pk`, so by the time a year file exists the evidence
of a repeated `pk` has already been removed; asking the year file could only ever answer
"none". All the CSVs are also grouped together rather than one at a time, because a key
repeated *across* two month files is exactly what an overlapping request window would
produce and a per-file check would never see it.

```{r}
#| label: tbl-bra-totalbr-duplicates

source(here::here("TOTALBR", "check_totalbr_duplicates.R"))

if (length(files) == 0) {
  knitr::kable(tibble::tibble(note = "No source files yet — nothing to check."))
} else {
  dups <- check_totalbr_duplicates(raw_dir = bra_totalbr_raw_dir,
                                   date_col = totalbr_date_col)
  if (nrow(dups) == 0) {
    knitr::kable(tibble::tibble(
      result = "No duplication found at any level."
    ))
  } else {
    dups |>
      dplyr::select(Level = LEVEL, Source = SOURCE, Year = YEAR,
                    `Repeated keys` = GROUPS, `Surplus rows` = EXTRA_ROWS,
                    `Year rows` = TOTAL_ROWS, `% of year` = PCT) |>
      knitr::kable(format.args = list(big.mark = ","))
  }
}
```

This is the heaviest table in the document — it groups the whole archive once per level —
so expect it to take a while on the full parquet.

How to read the result:

- **`key` above zero** is a delivery fault: the API returned the same row twice. Our year
  files are unaffected because the merge de-duplicates on `pk`, but any consumer reading
  the raw parts would double-count. This is the one worth reporting back to ICEA with the
  example rows.
- **`record` or `movement` above zero** is duplication in the source itself: the same
  movement is present twice and `pk` does not identify it as such, so no de-duplication key
  can remove it without a rule about which copy to keep.
- **`daily` on its own means nothing.** Two rows for the same callsign and route on one day
  are usually two genuine rotations. The minute gap is what separates them from a repeat,
  which is what the examples below report.

```{r}
#| label: examples-bra-totalbr-duplicates
#| eval: false

# the actual rows behind the most-repeated keys, with GAP_MIN between copies:
# a gap near zero is the same movement stored twice, hours apart is real traffic
totalbr_duplicate_examples("movement", n = 5, raw_dir = bra_totalbr_raw_dir)
totalbr_duplicate_examples("key",      n = 5, raw_dir = bra_totalbr_raw_dir)

# WHAT differs between the copies — the question the counts cannot answer
totalbr_duplicate_anatomy("movement", raw_dir = bra_totalbr_raw_dir)
```

The anatomy table is the one that decides whether the `movement` level is duplication at
all. If `li_orgaos` varies in most groups, the source is emitting **one row per ATS unit
crossed** and the flight is a single flight — the rows must be collapsed before anything
counts movements, but nothing is duplicated. If only `dh_inicio`/`dh_fim` vary, the same
movement really is stored twice.

### What the first run showed

Measured over 2023–2026 (the archive for 2023–2025, the API for 2026):

| Level | 2023–2025 | 2026 |
| --- | --- | --- |
| `key` | 13–285 rows (≤0.016%) | 194 rows in the parts |
| `record` | same rows as `key` | 134 rows (0.012%) |
| `movement` | **~3.6% of every year** | 0.024% |
| `daily` | 5.8–6.6% | 13.8% |

Three readings, in order of how much they matter:

1. **`movement` is ~3.6% in the archive and ~0.02% in the API year.** A 150-fold difference
   between two producers of the same table is not an operational change, it is a difference
   in how the rows are built. Until the anatomy table says what varies between the copies,
   **do not compare a movement count from the archive with one from the API** — that is the
   same trap that produced a negative additional time in the ASMA work, in a new place.
2. **`key` and `record` agree exactly for 2023–2025**, which confirms those copies are
   identical in every field: a plain repeated row, and rare enough to be ignorable.
3. **`daily` doubles in 2026** while `movement` collapses, so the 2026 repeats are at
   *different* times of day. That is what genuine repeated rotations look like — but it can
   equally be circuit training (`co_addep == co_addes`) that the archive excluded. Worth
   splitting those out before drawing any conclusion.

Nothing here repairs anything, deliberately. De-duplicating a national movement table
changes every count computed from it, and which copy to keep is DECEA's call, not a choice
this pipeline should make silently.

### One cause we removed on our side first

Before measuring anyone else's duplication it was worth removing ours. The downloader
paginated with `order=dt_dia.asc`, and thousands of rows share a `dt_dia` — with ties, the
order between pages is whatever the database happens to produce per request, so a row can
come back on two consecutive pages (or on neither). The ordering now includes the unique
id, which makes the sort total and the pages a genuine partition of the window.

If your parts were downloaded before that fix, re-fetch one month with `force = TRUE` and
compare the `key` level: any difference was ours, not the source's.

## Traffic Profile

A first look at what the table actually contains, once a year is complete: the mix of
flight rules and flight types, and the busiest aerodromes. This is the sanity check that
the file describes Brazilian traffic and not a filtered subset of it.

```{r}
#| label: tbl-bra-totalbr-profile

if (length(files) == 0) {
  knitr::kable(tibble::tibble(note = "No source files yet — nothing to profile."))
} else {
  prof <- totalbr_count_by(c("li_regravoo", "li_tipovoo"),
                           bra_totalbr_raw_dir, totalbr_date_col)

  if (nrow(prof) == 0) {
    knitr::kable(tibble::tibble(
      note = "Neither source exposes li_regravoo / li_tipovoo as plain columns."
    ))
  } else {
    prof |>
      # the li_* columns are pipe-separated after the CSV round-trip; the first
      # value is the flight's own rule/type, further values are later changes
      dplyr::mutate(RULE = sub("\\|.*$", "", li_regravoo),
                    TYPE = sub("\\|.*$", "", li_tipovoo)) |>
      dplyr::group_by(YEAR) |>
      dplyr::summarise(
        MOVEMENTS = sum(MOVEMENTS),
        `IFR (%)` = round(100 * sum(MOVEMENTS[RULE == "I"]) / sum(MOVEMENTS), 1),
        `VFR (%)` = round(100 * sum(MOVEMENTS[RULE == "V"]) / sum(MOVEMENTS), 1),
        `Scheduled (%)`   = round(100 * sum(MOVEMENTS[TYPE == "S"]) / sum(MOVEMENTS), 1),
        `General av. (%)` = round(100 * sum(MOVEMENTS[TYPE == "G"]) / sum(MOVEMENTS), 1),
        .groups = "drop"
      ) |>
      dplyr::mutate(MOVEMENTS = format(MOVEMENTS, big.mark = ",")) |>
      knitr::kable()
  }
}
```

```{r}
#| label: tbl-bra-totalbr-study-airports
#| tbl-cap: "Movements at the twelve study airports (departures + arrivals)"

if (length(files) == 0) {
  knitr::kable(tibble::tibble(note = "No source files yet — nothing to summarise."))
} else {
  # one movement per END of the flight, so a domestic flight between two study
  # airports counts once at each — the usual airport movement convention. The two
  # ends are counted separately and stacked, never joined.
  dep <- totalbr_count_by("co_addep", bra_totalbr_raw_dir, totalbr_date_col) |>
    dplyr::rename(ICAO = co_addep)
  arr <- totalbr_count_by("co_addes", bra_totalbr_raw_dir, totalbr_date_col) |>
    dplyr::rename(ICAO = co_addes)

  movements <- dplyr::bind_rows(dep, arr)

  if (nrow(movements) == 0) {
    knitr::kable(tibble::tibble(
      note = "Neither source exposes co_addep / co_addes."
    ))
  } else {
    movements |>
      dplyr::filter(ICAO %in% bra_apts) |>
      dplyr::group_by(YEAR, ICAO) |>
      dplyr::summarise(MOVEMENTS = sum(MOVEMENTS), .groups = "drop") |>
      tidyr::pivot_wider(names_from = YEAR, values_from = MOVEMENTS,
                         values_fill = 0) |>
      dplyr::left_join(bra_apts_names, by = "ICAO") |>
      dplyr::relocate(NAME, .after = ICAO) |>
      dplyr::arrange(ICAO) |>
      knitr::kable(format.args = list(big.mark = ","))
  }
}
```

The same two counts also say how much of the table is *not* airport-to-airport traffic:

```{r}
#| label: tbl-bra-totalbr-no-aerodrome

if (length(files) == 0 || !exists("movements") || nrow(movements) == 0) {
  knitr::kable(tibble::tibble(note = "No source files yet — nothing to summarise."))
} else {
  dplyr::bind_rows(
    dplyr::mutate(dep, END = "departure"),
    dplyr::mutate(arr, END = "destination")
  ) |>
    dplyr::group_by(YEAR, END) |>
    dplyr::summarise(
      # AFIL = flight plan filed airborne (no departure aerodrome);
      # ZZZZ = the end is not an ICAO aerodrome. Both are valid values.
      `No aerodrome (%)` = round(
        100 * sum(MOVEMENTS[is.na(ICAO) | ICAO %in% c("AFIL", "ZZZZ")]) /
          sum(MOVEMENTS), 1),
      .groups = "drop"
    ) |>
    tidyr::pivot_wider(names_from = END, values_from = `No aerodrome (%)`) |>
    knitr::kable()
}
```

`AFIL` and `ZZZZ` are legitimate values, not errors: a flight plan filed airborne has no
departure aerodrome, and `ZZZZ` marks an end that is not an ICAO aerodrome. They matter
because any count of "flights between two aerodromes" that ignores them is wrong, and any
airport-level use of this table has to decide what to do with them.

## Open Points

1. **What a "movement" is in this table.** Each row is a flight, with `co_addep` and
   `co_addes` as its two ends, so an airport count has to decide whether to count the
   flight once or once per end. The table above counts per end; confirm that this matches
   how DECEA reports movements before comparing the numbers to a published figure.
2. **`dh_eet` is stored as a timestamp, not a duration** (e.g. `2019-03-13T08:00:00` for a
   flight on 2019-03-12). It looks like an elapsed time encoded onto an arbitrary date;
   confirm the convention with ICEA before using it.
3. **`dt_dia` versus `dh_inicio`.** They are identical in every sampled row. Confirm they
   are meant to be, or which one is authoritative if they can differ — the month windows
   filter on `dt_dia`, and a divergence would put rows in the wrong month file.
4. **`co_matricula` and `ds_rota` are largely null** in the sample. Confirm whether they
   are populated for later years before any analysis depends on them.
5. **Harmonisation to APDF is not defined yet.** Unlike taxi-time and ASMA, this table has
   no metric attached; what it gets harmonised into depends on what it is used for
   (traffic volumes, an IFR/VFR split, a national denominator). That step is deliberately
   left out of this document until the use is fixed, so nothing here bakes in a wrong
   assumption.
