# Brazil TOTALBR Ingestion

```{r}
#| label: setup
#| include: false

# Libraries, project paths and parameters — including every TOTALBR setting
# (totalbr_data_years, the directories, the column names). Change the study
# period there, not here.
source(here::here("_chapter-setup.R"))
```

This note documents the preparation of the Brazilian **TOTALBR** files — the all-Brazil
movement table, ODIN table `total_brasil`. It follows the same shape as
`Taxi-BRA-ingestion.qmd` and `ASMA-BRA-ingestion.qmd`: get the data, check what arrived,
inventory it, harmonise it, and verify the coverage before anything downstream uses it.

TOTALBR differs from the other two in what it is *for*. Taxi-time and ASMA are performance
metrics computed at the twelve study airports; TOTALBR is **national traffic volume** — one
row per flight anywhere in Brazil, including aerodromes outside the study set and flights
with no aerodrome at all (`AFIL` departures, `ZZZZ` destinations). It carries no times to
subtract and no unimpeded reference, so there is no reference year and no percentile here.
It is the denominator and the context: how much traffic there was, where, and of what kind.

### Two sources, one dataset

TOTALBR reaches this project two ways, and both are first-class:

| Source | What it holds | Where |
| --- | --- | --- |
| **parquet archive** | the history, all airports, up to 2025 | `data-raw/totalbr/*.parquet` (or `BRA_TOTALBR_PARQUET`) |
| **ODIN API** | whatever the archive does not cover | downloaded to `data-raw/totalbr/totalbr_<year>.csv` |

Every check below reads both through `TOTALBR/totalbr_sources.R`, so a year is judged
covered by the days it actually holds regardless of which source holds them. That is what
keeps the download honest: years already in the archive are not fetched again, and the
API is asked only for the gap.

Nothing in the ingestion filters by airport. TOTALBR is the national table and is kept
whole — restricting it to the twelve study airports is a downstream decision, and doing it
on the way in would discard data that could only be recovered by downloading the year
again.

::: {.callout-important}
## Prerequisites to run this document
Neither the parquet archive nor the CSVs are shipped with the repository. Put the archive
in `data-raw/totalbr/` (or point `BRA_TOTALBR_PARQUET` at it if you keep it elsewhere —
it is large), and get the remaining years with the download chunk below. Until then, the
tables render with a short notice instead of failing.
:::

## Getting the Source Data

### Check the endpoint first

A full download runs for hours, so confirm the endpoint and its column names with a single
small request before starting one. This is cheap and catches a renamed column immediately
instead of after a wasted afternoon.

```{r}
#| label: probe-bra-totalbr-api
#| eval: false

source(here::here("ODIN", "download_odin.R"))
odin_probe(totalbr_table)      # prints the columns and flags the date-like ones
```

Run on 2026-07-30 this answered with 18 columns —
`pk, id, dt_dia, dh_inicio, dh_fim, co_indicativo, co_addep, co_addes, dh_eobt, dh_eet,
co_matricula, co_modelo, ds_rota, li_orgaos, li_tipovoo, li_regravoo, li_prnav, li_rvsm` —
of which five are date-like (`dt_dia`, `dh_inicio`, `dh_fim`, `dh_eobt`, `dh_eet`). That
confirms the two choices the downloader is built on: the month windows filter on `dt_dia`,
and the unique key is `pk`.

### Download

The chunk below is `eval: false` — the download is slow, so it is not executed on render;
run it manually when you need data.

It is **incremental and resumable**, exactly like the KPI08 downloader (they share the
engine in `ODIN/download_odin.R`): one month per request window, each month saved under
`parts/` the moment it arrives, and a month judged already-downloaded by **the days it
contains** rather than by a file existing. So an interrupted run keeps everything it had,
and a month fetched while it was still the current month is fetched again once it closes
instead of staying frozen half-full.

Years you already hold as files go straight into `data-raw/totalbr/` with the same naming
(`totalbr_2019.csv`); the downloader inspects what each covers and fetches only the gaps.

```{r}
#| label: download-bra-totalbr-data
#| eval: false

# sourcing only defines the functions; nothing is fetched until they are called
source(here::here("TOTALBR", "download_totalbr.R"))
source(here::here("TOTALBR", "totalbr_sources.R"))

# only the years the parquet archive does not already cover
need <- totalbr_missing_years(totalbr_data_years, raw_dir = bra_totalbr_raw_dir)
message("Years to fetch: ", if (length(need)) paste(need, collapse = ", ") else "none")

if (length(need) > 0)
  download_totalbr(need, out_dir = bra_totalbr_raw_dir)
```

To fetch one year regardless of what is on disk — which is the right way to test the
endpoint before trusting a long run — call the downloader directly:

```r
source(here::here("TOTALBR", "download_totalbr.R"))
download_totalbr(2026, out_dir = bra_totalbr_raw_dir)
```

A word on years that come back nearly empty. Asking for 2019 returns a few dozen rows
(14 in March, 32 in May) and "no rows" for every other month. That is the API answering
truthfully, not a broken filter: `total_brasil` holds only a token sample for the early
years. The history comes from the parquet archive, and the API is there for the present.
The rule to apply is the same either way — trust the day counts in the coverage tables
below, not the fact that a download ran.

## Checking What Arrived

```{r}
#| label: check-bra-totalbr-data

source(here::here("TOTALBR", "totalbr_sources.R"))

src <- totalbr_sources(bra_totalbr_raw_dir)

if (length(src$parquet) == 0 && length(src$csv) == 0) {
  knitr::kable(tibble::tibble(
    status = "no source data yet",
    `what to do` = paste0(
      "Put the parquet archive in ", bra_totalbr_raw_dir,
      " (or set BRA_TOTALBR_PARQUET), and/or run the ",
      "`download-bra-totalbr-data` chunk above."
    )
  ))
} else {
  # the day counts are computed once here and reused by every table below, so a
  # large archive is scanned once per render rather than five times
  totalbr_counts <- totalbr_day_counts(bra_totalbr_raw_dir, totalbr_date_col)

  knitr::kable(tibble::tibble(
    `parquet archive` = if (length(src$parquet) == 0) "none"
                        else paste(basename(src$parquet), collapse = ", "),
    `downloaded CSVs` = if (length(src$csv) == 0) "none"
                        else paste(sort(stringr::str_extract(basename(src$csv),
                                                             "[0-9]{4}")),
                                   collapse = ", "),
    `days held`       = dplyr::n_distinct(totalbr_counts$DATE),
    movements         = format(sum(totalbr_counts$MOVEMENTS), big.mark = ",")
  ))
}
```

## Source Inventory

```{r}
#| label: tbl-bra-totalbr-source-inventory
#| tbl-cap: "Inventory of the raw TOTALBR source files"

files <- c(src$parquet, src$csv)

if (length(files) == 0) {
  knitr::kable(tibble::tibble(
    note = paste0("No source files in ", bra_totalbr_raw_dir,
                  " yet. See the prerequisites above.")
  ))
} else {
  purrr::map(files, function(file) {
    tibble::tibble(
      File        = basename(file),
      Kind        = ifelse(grepl("\\.parquet$", file), "parquet archive",
                           "downloaded CSV"),
      Year        = tidyr::replace_na(
                      stringr::str_extract(basename(file), "[0-9]{4}"), "several"),
      `Size (MB)` = round(file.info(file)$size / 1024^2, 1),
      Modified    = format(file.info(file)$mtime, "%Y-%m-%d %H:%M")
    )
  }) |>
    purrr::list_rbind() |>
    dplyr::arrange(Kind, Year) |>
    knitr::kable()
}
```

The files are semicolon-delimited UTF-8, written by the downloader. The per-month parts it
fetched live under `data-raw/totalbr/parts/` and are what makes a re-run cheap; the year
file is the merge of them, de-duplicated on `pk` and sorted by `dt_dia`.

## What the Columns Mean

| Source field | Meaning | Note |
| --- | --- | --- |
| `pk` | Row hash | **The unique key.** De-duplication keys on this |
| `id` | Callsign + ADEP + ADES | e.g. `PPBLVSBPASBSM`. **Repeats** on every flight of that route — never a key |
| `dt_dia` | Movement day/time | What the month windows filter on |
| `dh_inicio` | Start of the movement | Equal to `dt_dia` in most sampled rows |
| `dh_fim` | End of the movement | Equal to `dh_inicio` when no duration was recorded |
| `co_indicativo` | Callsign | `PPBLV`, `AZU2429` |
| `co_addep` | Departure aerodrome | `AFIL` = flight plan filed airborne, i.e. no departure aerodrome |
| `co_addes` | Destination aerodrome | `ZZZZ` = destination not an ICAO aerodrome |
| `dh_eobt` | Estimated off-block time | From the flight plan |
| `dh_eet` | Estimated elapsed time | Encoded as a timestamp, not a duration — read with care |
| `co_matricula` | Registration | Largely null in the sample |
| `co_modelo` | Aircraft type | `SR22`, `AT76`; `XXXX` where unknown |
| `ds_rota` | Route | Largely null in the sample |
| `li_orgaos` | ATS units involved | JSON array, e.g. `["APPFI"]` |
| `li_tipovoo` | Flight type | JSON array: `S` scheduled, `G` general aviation |
| `li_regravoo` | Flight rule | JSON array: `I` IFR, `V` VFR |
| `li_prnav` | P-RNAV capability | JSON array of 0/1 |
| `li_rvsm` | RVSM capability | JSON array of 0/1 |

Two conventions matter when reading these files:

**The five `li_*` columns are JSON arrays.** They cannot be written to a delimited file as
they stand, so the downloader collapses each to a pipe-separated string. A row with
`["APPFI","ACCCW"]` is stored as `APPFI|ACCCW`. Split them again on read when a row can
carry more than one value:

```r
d$orgaos <- strsplit(d$li_orgaos, "|", fixed = TRUE)
```

**`id` is not an identifier.** It is the callsign glued to the airport pair, so the same
`id` returns on every flight of that route — `PPBLVSBPASBSM` appears on 2019-03-12 and
again months later. De-duplicating on it would delete almost the entire dataset. `pk` is
the key, and the downloader uses it.

## Verifying the Contents

Having a file is not the same as having the data. This table answers the question that
decides whether a year is usable — **how many of its days are actually present** — over
both sources at once, so a year split between the archive and a download is judged on the
union rather than on either half.

```{r}
#| label: tbl-bra-totalbr-contents

if (length(files) == 0) {
  knitr::kable(tibble::tibble(
    note = "No source files yet — nothing to verify."
  ))
} else {
  totalbr_year_coverage(counts = totalbr_counts) |>
    dplyr::transmute(
      Year        = YEAR,
      Sources     = SOURCES,
      Movements   = format(MOVEMENTS, big.mark = ","),
      `Days held` = DAYS_HELD,
      `Days possible` = DAYS_POSSIBLE,
      Complete = dplyr::if_else(
        COVERED, "yes",
        paste0("NO — ", DAYS_POSSIBLE - DAYS_HELD, " day(s) missing"))
    ) |>
    knitr::kable()
}
```

Read the table like this:

- **Days possible** is the whole year for a past year and the days elapsed so far for the
  current one, so 2026 reading fewer days than 2025 is expected, not a gap.
- **Complete** reading `NO` for a past year means that year is short. Re-run the download
  chunk: it fetches only the missing months, and it skips whatever the parquet archive
  already covers.
- **Sources** shows where each year came from. A year reading `parquet + csv` is held by
  both; the day count is de-duplicated across them, so the overlap is not double-counted.

## Monthly Coverage

The completeness check above is per year. This one is per month, which is where a partial
download actually shows up — a month with a handful of days is the signature of an
extraction that was cut off, and it is exactly what produced a bogus collapse in the ASMA
monthly trend before it was caught.

```{r}
#| label: tbl-bra-totalbr-monthly-coverage

if (length(files) == 0) {
  knitr::kable(tibble::tibble(note = "No source files yet — nothing to check."))
} else {
  totalbr_counts |>
    dplyr::group_by(YEAR, MONTH) |>
    dplyr::summarise(MOVEMENTS = sum(MOVEMENTS),
                     # distinct days: a month present in both sources counts once
                     DAYS = dplyr::n_distinct(DATE),
                     .groups = "drop") |>
    dplyr::mutate(
      # a month holding only a few days is a truncated extraction, not a quiet month
      STATUS = dplyr::case_when(
        DAYS >= 28 ~ "complete",
        DAYS >= 20 ~ "partial — check",
        TRUE       ~ "TRUNCATED — re-download"
      )
    ) |>
    knitr::kable()
}
```

A month flagged `TRUNCATED` is re-fetched by simply re-running the download chunk: the
engine judges a month by the days it holds, so a partial month is not treated as done.
Forcing a re-download of specific years is the heavier option:

```r
source(here::here("TOTALBR", "download_totalbr.R"))
download_totalbr(<the offending years>, out_dir = bra_totalbr_raw_dir, force = TRUE)
```

## Duplication Check

ICEA/DECEA have reported duplication in the ODIN data. The check applies the operational
definition, which is deliberately strict — two rows are the same movement only when they
could not be two different flights:

| Test | Rule | Why it is conclusive |
| --- | --- | --- |
| **same `pk`** | the same row hash twice | `pk` identifies the row; no judgement needed |
| **near-duplicate** | same `co_matricula`, same aerodrome pair, and `dh_inicio` **or** `dh_fim` within a few minutes | one airframe cannot be in one place twice at once |

Anything else with a different `pk` is **not** counted. Two rows sharing a callsign, a route
and a day are two flights until the registration and the clock say otherwise: an aircraft
can fly the same route several times a day, and a callsign can be reused by a different
airframe on the same day.

The tolerance is 10 minutes by default (`tol_min`). Both `dh_inicio` and `dh_fim` are tested
because two records of one movement can agree on when it ended while differing on when it
started.

Two details that decide whether the numbers mean anything. The `pk` test reads the
per-month parts in `data-raw/totalbr/parts/` as well, pooled together: the merge
de-duplicates on `pk`, so the year file can no longer show whether the API repeated a row,
and a `pk` repeated *across* two months — what an overlapping request window produces — is
invisible to a per-file check. And `dh_inicio`/`dh_fim` values at the epoch (1970-01-01,
which ODIN uses in place of an empty value) are read as missing, so they cannot pile up at
the start of time and look like a thousand simultaneous movements.

```{r}
#| label: tbl-bra-totalbr-duplicates

source(here::here("TOTALBR", "check_totalbr_duplicates.R"))

if (length(files) == 0) {
  knitr::kable(tibble::tibble(note = "No source files yet — nothing to check."))
} else {
  dups <- check_totalbr_duplicates(raw_dir = bra_totalbr_raw_dir,
                                   date_col = totalbr_date_col)
  if (nrow(dups) == 0) {
    knitr::kable(tibble::tibble(result = "No duplication found."))
  } else {
    dups |>
      dplyr::select(dplyr::any_of(c(
        Year = "YEAR", Rows = "ROWS",
        `Same pk (merged)` = "SAME_PK", `Same pk (parts)` = "SAME_PK_PARTS",
        `Near-duplicates` = "NEAR_DUP", `Near %` = "NEAR_PCT",
        `No registration` = "NO_REG", `No reg. %` = "NO_REG_PCT"))) |>
      knitr::kable(format.args = list(big.mark = ","))
  }
}
```

`No registration` is there so the near-duplicate figure is read for what it covers: rows
with an empty `co_matricula` cannot be tested by this rule at all, and in the early samples
that field was mostly null. A near-duplicate count of zero over a year that is 90% without
registration says nothing.

```{r}
#| label: examples-bra-totalbr-duplicates
#| eval: false

# the flagged rows next to the ones they repeat, with the gap in minutes
totalbr_duplicate_examples(2026, raw_dir = bra_totalbr_raw_dir)

# widen or tighten what counts as "almost the same time"
check_totalbr_duplicates(2026, tol_min = 5,  raw_dir = bra_totalbr_raw_dir)
check_totalbr_duplicates(2026, tol_min = 30, raw_dir = bra_totalbr_raw_dir)
```

Run the examples before reporting anything. The gap in minutes is what turns a count into
evidence, and it is what ICEA need in order to find the records on their side.

### Why the earlier, looser tests were dropped

An earlier version of this check also counted rows sharing `id` + `dt_dia`, and rows sharing
callsign + route + day. Those are not duplication and the numbers were misleading — nearly
4% of the archive at one point. Two findings came out of chasing them, and both are about
field quality rather than duplication:

#### `dt_dia` does not resolve a movement

Pulling the rows showed the shape of it. A group flagged by the old, looser test:

```
pk        id      dt_dia               dh_inicio             dh_fim
6104BE1E… PPMFM…  2025-02-16 00:00:00  2025-02-16 20:56:03   2025-02-16 21:07:58
E2F1E8D2… PPMFM…  2025-02-16 00:00:00  2025-02-16 18:12:15   2025-02-16 18:12:19
82B9BD65… PPMFM…  2025-02-16 00:00:00  2025-02-16 18:46:24   2025-02-16 18:47:40
…ten rows, dh_inicio spanning 13:55 to 21:53
```

Ten **different** flights by one aircraft (`ZZZZ` → `ZZZZ`, local flying) on one day, each
with its own start and end, all sharing one `dt_dia`. Any key built on `dt_dia` collapses
them into one. They are not duplicates, and the strict rule agrees: their `dh_inicio` are
hours apart.

How widespread that is needs measuring rather than eyeballing, and the first attempt at it
was wrong. A histogram of `minute(dt_dia)` put 11% of rows at minute zero, which is **not**
the same as midnight — 13:00:00 and 20:00:00 are minute zero too. Measured properly,
`dt_dia` is a bare date on only about 0.1% of rows:

| Year | `dt_dia` at midnight | of rows |
| --- | --- | --- |
| 2023 | 1,932 | 1,818,076 |
| 2024 | 1,912 | 1,959,619 |
| 2025 | 1,747 | 2,109,623 |
| 2026 | 337 | 1,232,397 |

So the bare-date rows are real but far too few to explain 3.6% of a year. What the minute
histogram does show is a stamp **rounded to the hour** on a large share of rows, and a
stamp that resolves only to the hour cannot separate two movements of one aircraft inside
that hour any more than a bare date can separate them inside the day.

`totalbr_midnight_share()` now reports both — bare dates and on-the-hour stamps — for all
three time columns, and `totalbr_stamp_agreement()` answers the question underneath: how
often `dt_dia` is the movement time at all.

```{r}
#| label: tbl-bra-totalbr-stamps
#| eval: false

totalbr_midnight_share(raw_dir = bra_totalbr_raw_dir)   # per column, per year
totalbr_stamp_agreement(raw_dir = bra_totalbr_raw_dir)  # dt_dia vs dh_inicio
```

In the API payload `dt_dia` and `dh_inicio` are equal on every sampled row. Wherever they
diverge, `dt_dia` is something else — a filing time, a rounded stamp, a bare date — and is
not a movement time.

**The rule for the pipeline:** `dt_dia` is safe for the monthly download windows, because
even a bare date lands on the right day. It is not safe as a movement time, and neither
`dh_inicio` nor `dh_fim` can be assumed to carry a time of day without checking — the near
duplicate test excludes the rows where they do not and reports them as `NO_TIME`.

**The fallback is not confined to `dt_dia`.** Running the near-duplicate check on 2026
turned up rows where `dh_inicio` and `dh_fim` are also bare dates:

```
co_matricula co_indicativo co_addep co_addes dh_inicio            dh_fim               GAP_MIN
00403A       UAF1225       GMMX     SBEG     2026-06-27 00:00:00  2026-06-27 00:00:00       NA
00403A       UAF1225       GMMX     SBEG     2026-06-27 00:00:00  2026-06-27 00:00:00        0
1002         GAF922        SBBR     EDDB     2026-07-03 00:00:00  2026-07-03 00:00:00        0
```

A `GAP_MIN` of 0 there is not two records of one movement — it is two empty stamps. Left
alone, every such row of an aircraft on one day would be flagged as a duplicate of the
others, and the check would measure the missing clock rather than the traffic. Those rows
are excluded from the test and counted as `NO_TIME`; `TESTED_PCT` says what share of the
year the rule could actually judge.

Worth checking whether the zeroed rows are a particular class of traffic — the examples
above are foreign and military (`UAF1225` GMMX→SBEG, `GAF922` SBBR→EDDB), which would
suggest movements known to Brazil only as a flight plan, with no observed times. That would
make the absence expected rather than a fault.

#### `dh_eet` is not a timestamp, and is encoded differently in each source

In the API payload, `dh_eet` minus `dh_eobt` equals **the time of day of `dh_eobt`**, on
every sampled row:

| `dh_eobt` | `dh_eet` | difference | time of day |
| --- | --- | --- | --- |
| 2019-03-12 16:00 | 2019-03-13 08:00 | 16h00 | 16:00 |
| 2019-03-12 18:50 | 2019-03-13 13:40 | 18h50 | 18:50 |
| 2019-05-19 13:01 | 2019-05-20 02:02 | 13h01 | 13:01 |

That is not an elapsed time — a light single from Porto Alegre to Santa Maria does not fly
for sixteen hours. It reads as the EOBT having been added to itself as though it were a
duration. In the parquet archive the field behaves differently again: `dh_eet` equals
`dh_eobt` exactly, a difference of zero.

Either way the real EET is not recoverable from this column, and it must never be read as a
timestamp. Confirm at scale, then report:

```r
d <- data.table::fread(file = file.path(bra_totalbr_raw_dir, "totalbr_2026.csv"),
                       sep = ";", select = c("dh_eobt", "dh_eet"),
                       colClasses = "character")
d    <- d[nzchar(dh_eet) & !is.na(dh_eet)]
eobt <- as.POSIXct(d$dh_eobt, tz = "UTC")
tod  <- as.numeric(difftime(eobt, trunc(eobt, "days"), units = "mins"))
gap  <- as.numeric(difftime(as.POSIXct(d$dh_eet, tz = "UTC"), eobt, units = "mins"))
summary(gap - tod)     # all zero confirms the pattern
```

Run over the full 2026 file, that confirmed it and turned up a second fault:

```
     Min.   1st Qu.    Median      Mean   3rd Qu.      Max.
-29750400         0         0    -92576         0      1440
```

The three quartiles are exactly zero, so the identity holds for at least 75% of rows — a
rule, not an artefact of the sample. The minimum is roughly 56.6 years in minutes: the
distance back to **1970-01-01**. Those rows carry the **epoch as a null sentinel** instead
of an empty value, and the mean says they are about 0.31% of the file.

That is the most dangerous shape a null can take. An `NA` propagates and shows itself; a
1970 date enters an average silently and drags it — which is exactly what happened here,
where 0.3% of the rows pulled the mean from 0 to −92,576. Anything reading `dh_eet` must
reject epoch dates explicitly, not rely on `na.rm`. The duplicate check does this for
`dh_inicio`/`dh_fim` already: without it, every epoch-stamped row would sit at the same
instant and be flagged as a duplicate of every other.

Nothing here repairs anything, deliberately. De-duplicating a national movement table
changes every count computed from it, and which copy to keep is DECEA's call, not a choice
this pipeline should make silently.

### One cause we removed on our side first

Before measuring anyone else's duplication it was worth removing ours. The downloader
paginated with `order=dt_dia.asc`, and thousands of rows share a `dt_dia` — with ties, the
order between pages is whatever the database happens to produce per request, so a row can
come back on two consecutive pages (or on neither). The ordering now includes the unique
id, which makes the sort total and the pages a genuine partition of the window.

Re-downloading 2026 after the fix confirmed it: the same-`pk` count in the parts fell from
194 to 134 and the year gained exactly 60 rows. Sixty rows had been served twice, each in
place of one never served at all. What remains is the source's.

If your parts predate that fix, re-fetch a month with `force = TRUE` and compare.

## Traffic Profile

A first look at what the table actually contains, once a year is complete: the mix of
flight rules and flight types, and the busiest aerodromes. This is the sanity check that
the file describes Brazilian traffic and not a filtered subset of it.

```{r}
#| label: tbl-bra-totalbr-profile

if (length(files) == 0) {
  knitr::kable(tibble::tibble(note = "No source files yet — nothing to profile."))
} else {
  prof <- totalbr_count_by(c("li_regravoo", "li_tipovoo"),
                           bra_totalbr_raw_dir, totalbr_date_col)

  if (nrow(prof) == 0) {
    knitr::kable(tibble::tibble(
      note = "Neither source exposes li_regravoo / li_tipovoo as plain columns."
    ))
  } else {
    prof |>
      # the li_* columns are pipe-separated after the CSV round-trip; the first
      # value is the flight's own rule/type, further values are later changes
      dplyr::mutate(RULE = sub("\\|.*$", "", li_regravoo),
                    TYPE = sub("\\|.*$", "", li_tipovoo)) |>
      dplyr::group_by(YEAR) |>
      dplyr::summarise(
        MOVEMENTS = sum(MOVEMENTS),
        `IFR (%)` = round(100 * sum(MOVEMENTS[RULE == "I"]) / sum(MOVEMENTS), 1),
        `VFR (%)` = round(100 * sum(MOVEMENTS[RULE == "V"]) / sum(MOVEMENTS), 1),
        `Scheduled (%)`   = round(100 * sum(MOVEMENTS[TYPE == "S"]) / sum(MOVEMENTS), 1),
        `General av. (%)` = round(100 * sum(MOVEMENTS[TYPE == "G"]) / sum(MOVEMENTS), 1),
        .groups = "drop"
      ) |>
      dplyr::mutate(MOVEMENTS = format(MOVEMENTS, big.mark = ",")) |>
      knitr::kable()
  }
}
```

```{r}
#| label: tbl-bra-totalbr-study-airports
#| tbl-cap: "Movements at the twelve study airports (departures + arrivals)"

if (length(files) == 0) {
  knitr::kable(tibble::tibble(note = "No source files yet — nothing to summarise."))
} else {
  # one movement per END of the flight, so a domestic flight between two study
  # airports counts once at each — the usual airport movement convention. The two
  # ends are counted separately and stacked, never joined.
  dep <- totalbr_count_by("co_addep", bra_totalbr_raw_dir, totalbr_date_col) |>
    dplyr::rename(ICAO = co_addep)
  arr <- totalbr_count_by("co_addes", bra_totalbr_raw_dir, totalbr_date_col) |>
    dplyr::rename(ICAO = co_addes)

  movements <- dplyr::bind_rows(dep, arr)

  if (nrow(movements) == 0) {
    knitr::kable(tibble::tibble(
      note = "Neither source exposes co_addep / co_addes."
    ))
  } else {
    movements |>
      dplyr::filter(ICAO %in% bra_apts) |>
      dplyr::group_by(YEAR, ICAO) |>
      dplyr::summarise(MOVEMENTS = sum(MOVEMENTS), .groups = "drop") |>
      tidyr::pivot_wider(names_from = YEAR, values_from = MOVEMENTS,
                         values_fill = 0) |>
      dplyr::left_join(bra_apts_names, by = "ICAO") |>
      dplyr::relocate(NAME, .after = ICAO) |>
      dplyr::arrange(ICAO) |>
      knitr::kable(format.args = list(big.mark = ","))
  }
}
```

The same two counts also say how much of the table is *not* airport-to-airport traffic:

```{r}
#| label: tbl-bra-totalbr-no-aerodrome

if (length(files) == 0 || !exists("movements") || nrow(movements) == 0) {
  knitr::kable(tibble::tibble(note = "No source files yet — nothing to summarise."))
} else {
  dplyr::bind_rows(
    dplyr::mutate(dep, END = "departure"),
    dplyr::mutate(arr, END = "destination")
  ) |>
    dplyr::group_by(YEAR, END) |>
    dplyr::summarise(
      # AFIL = flight plan filed airborne (no departure aerodrome);
      # ZZZZ = the end is not an ICAO aerodrome. Both are valid values.
      `No aerodrome (%)` = round(
        100 * sum(MOVEMENTS[is.na(ICAO) | ICAO %in% c("AFIL", "ZZZZ")]) /
          sum(MOVEMENTS), 1),
      .groups = "drop"
    ) |>
    tidyr::pivot_wider(names_from = END, values_from = `No aerodrome (%)`) |>
    knitr::kable()
}
```

`AFIL` and `ZZZZ` are legitimate values, not errors: a flight plan filed airborne has no
departure aerodrome, and `ZZZZ` marks an end that is not an ICAO aerodrome. They matter
because any count of "flights between two aerodromes" that ignores them is wrong, and any
airport-level use of this table has to decide what to do with them.

## Open Points

1. **What a "movement" is in this table.** Each row is a flight, with `co_addep` and
   `co_addes` as its two ends, so an airport count has to decide whether to count the
   flight once or once per end. The table above counts per end; confirm that this matches
   how DECEA reports movements before comparing the numbers to a published figure.
2. ~~`dh_eet` is stored as a timestamp, not a duration~~ — **measured, twice.** The
   difference from `dh_eobt` is the EOBT's own time of day in the API (75%+ of rows,
   exactly), and zero in the archive. On top of that, ~0.3% of rows carry 1970-01-01 as a
   null sentinel. The EET is not recoverable from this column in either source, and any
   code touching it must exclude epoch dates rather than trust `na.rm`. ICEA need to
   confirm what the field is meant to hold.
3. **`dt_dia` versus `dh_inicio`.** They are equal on rows that carry a real time, but
   around 11% of rows have `dt_dia` at midnight while `dh_inicio` holds the actual movement
   time. Use `dh_inicio` for anything time-of-day; `dt_dia` is still safe for the month
   windows, because a midnight stamp still lands on the right day.
4. **`co_matricula` and `ds_rota` are largely null** in the sample. Confirm whether they
   are populated for later years before any analysis depends on them.
5. **Harmonisation to APDF is not defined yet.** Unlike taxi-time and ASMA, this table has
   no metric attached; what it gets harmonised into depends on what it is used for
   (traffic volumes, an IFR/VFR split, a national denominator). That step is deliberately
   left out of this document until the use is fixed, so nothing here bakes in a wrong
   assumption.
