# Brazil VRA (ANAC) Data Ingestion

```{r}
#| label: setup
#| include: false

source(here::here("_chapter-setup.R"))
source(here::here("VRA", "download_vra.R"))
source(here::here("VRA", "check_vra_duplicates.R"))
```

This note documents the ingestion of **VRA** — *Voo Regular Ativo* — ANAC's record of the
regular flights operated to, from and within Brazil, for the Brazil–Europe
operational-efficiency study.

VRA is a **different kind of source** from the others in this repository, and the difference
matters more than the schema does. `dstaxi`, `kpi08` and TOTALBR are ATC observations: what
the air navigation system saw happen. VRA is **filed by the airlines** — the scheduled time,
the realised time, and ANAC's verdict on the difference. When VRA and TOTALBR disagree about
a flight, that is two measurements differing, not automatically an error in either.

It is also the only source here that answers punctuality **directly**. `ds_situacao_voo`,
`ds_situacao_partida` and `ds_situacao_chegada` are already computed and carry ANAC's own
banding (`Pontual`, `Antecipado`, `Atraso 30-60`, `Atraso 60-120`), with
`ds_justificativa` for the reason. Nothing has to be derived to report delay — which also
means the definitions are ANAC's, not ours, and must be quoted as such.

## The endpoint

Base: `https://sas.anac.gov.br/sas/vra_api/vra`

```
.../vra?dt_referencia1=01122024&dt_referencia2=31122024
```

Four properties of this API are encoded in `VRA/download_vra.R` because none of them is
guessable, and each was found by probing rather than assumed:

| Property | What it means for the download |
| --- | --- |
| The window is `dt_referencia1` / `dt_referencia2`, both **DDMMYYYY**, both **inclusive** | Neither ISO nor `d/m/Y`. A wrong format returns an empty window, not an error. `vra_window()` formats it so no caller does it by hand. |
| The answer is **JSON inside JSON** | The body is a JSON *string* wrapping the array of records: one decode yields a character vector of length one that looks like corruption. `vra_parse_json()` unwraps it. |
| **The wide window is honoured** | December 2024 returns 85,919 rows over 31 distinct days, 66 MB — 32× a single day. So a month is one request, not thirty. |
| There is **no unique key** | No `id`, no hash. Nothing is de-duplicated on download. |

::: {.callout-note}
## Why the wide window was measured rather than assumed
TATIC accepts `datai`/`dataf` spanning a month and silently returns **only the first day** —
no error, no warning. Discovering that after a year had been "downloaded" is expensive, so
the check is built into the downloader: every month is compared against the days it should
hold, and a month window that comes back with one day says so and names the fix.
:::

## Downloading

```{r}
#| label: download-bra-vra-data
#| eval: false

source(here::here("VRA", "download_vra.R"))

# The period comes from _chapter-setup.R, and R holds the value it read when the
# session started. Editing the file does not change a running session -- source
# it again, and check, before a download that takes hours goes after the wrong
# years.
source(here::here("_chapter-setup.R"))
vra_data_years                        # <- confirm this is the period you meant

download_vra(vra_data_years)
download_vra(2026)                    # or name the year outright
```

One file per year in `data-raw/vra/`, with the month parts under `parts/`.

**Each month is written the moment it arrives**, before the next is asked for, so a run
stopped halfway keeps every month it had. Re-running skips the months already complete, so
interrupting costs nothing but the month in flight — and completeness is judged by the days
a part holds, not by a request having returned, so a month that came back short is fetched
again rather than trusted.

**Completeness is a statement about days present, never about a request having returned.** A
month fetched while it was still open is short by design and is fetched again once it
closes. A month that came back truncated is short by fault. Neither is accepted as complete,
and the two are deliberately not distinguished at this stage — both mean "fetch it again".

## Coverage — the days each month actually holds

```{r}
#| label: tbl-bra-vra-coverage
#| tbl-cap: "Days and rows per month, as downloaded"

cov <- vra_coverage(bra_vra_raw_dir)
if (is.null(cov)) {
  knitr::kable(tibble::tibble(
    note = "Nothing downloaded yet — run the chunk above."))
} else {
  knitr::kable(cov)
}
```

It reads the month **parts**, not the year files. The year file is written only after the
last month of a year has been fetched, so a run interrupted halfway — the normal way a long
download ends — leaves months on disk and no year file at all.

A `SHORT` month is either still open or truncated. Read it together with today's date before
concluding which.

## Is there a key?

VRA carries no id, so any de-duplication needs a key we construct — and a key that looks
plausible is exactly what costs the most when it is wrong. TOTALBR's `id` is the precedent:
it reads like an identifier, repeats on every flight of the same route, and de-duplicating on
it would have deleted almost the entire dataset.

The key is **operator + flight number + origin + destination + `dt_referencia` + scheduled
departure** — `VRA_KEY_COLS` in `VRA/check_vra_duplicates.R`. It took three passes over the
data to get there, and each component earned its place by a case that broke the version
before it:

**A flight number is not a route.** AAL `Z0930` on 14/01/2026 appears as SBGR → KFLL and
again as KFLL → KMIA — two legs of one number, two movements, two rows.

**`dt_partida_prevista` is often empty.** Those same rows carry `""`: they are `cd_di = 4`,
non-scheduled services, which have no planned departure. A blank component makes the key
`AAL|Z0930|`, so every Z0930 collides — including the same route on 14/01 and on 21/02. **A
key component that can be empty does not divide anything; it merges.** `dt_referencia` is
the day the API itself partitions on and is never blank, so it carries the date
unconditionally.

**A route flown twice in one day is two flights.** ACA0090 CYYZ → SBGR on 05/01/2026 is filed
once for 00:35 and once for 23:00 — same operator, number, route and day, two operations.
Only the scheduled time separates them, so it comes back into the key alongside
`dt_referencia` rather than instead of it: a blank scheduled time falls back on the day and
the route, and a shared day is still split by the scheduled time.

> **What this gives up.** A flight re-filed with a corrected schedule now reads as a separate
> flight rather than as a copy of the first. That distinction was never observable: VRA
> carries no submission timestamp, so "the later filing" cannot be identified, only guessed
> at. Sixteen of the twenty-three repeated groups found under the previous key differed on
> exactly this column and are now separate rows.

```{r}
#| label: tbl-bra-vra-key
#| tbl-cap: "How often the candidate key repeats"

prof <- vra_key_profile(quiet = TRUE)
if (is.null(prof)) {
  knitr::kable(tibble::tibble(note = "Nothing downloaded yet."))
} else {
  knitr::kable(prof$summary)
}
```

`SURPLUS` is what a naive de-duplication would delete; `IN_A_GROUP` is how many rows are
involved in a repeat at all — the two differ because a group of three copies contributes
three rows and two surplus.

### What the repeats ARE

A count says how much would be deleted. It does not say whether deleting is right. Two rows
under one key are one of three things, and only the last means the key is broken:

| The copies | What they are | What to do |
| --- | --- | --- |
| identical everywhere | the same filing received twice | either copy will do |
| differ on the realised times or the situation | the flight **re-filed** after a change | keep the later filing; fill blanks from the earlier |
| differ on the aerodromes | **two real flights** the key cannot tell apart | the key is wrong — widen it |

Which of the three it is can be read off directly: for each column, in how many repeated
groups does its value vary?

```{r}
#| label: tbl-bra-vra-key-columns
#| tbl-cap: "Where the copies differ, as a share of the repeated groups"

if (is.null(prof) || nrow(prof$columns) == 0) {
  knitr::kable(tibble::tibble(note = "No repeated key, or nothing downloaded."))
} else {
  knitr::kable(prof$columns)
}
```

A column varying in ~0% of groups is carried identically by every copy. A column varying in
most groups is what the copies exist to record — and is therefore what any collapse has to
choose between rather than discard.

Then look at the groups themselves rather than inferring from a percentage. `only_col`
restricts to the groups that differ on one column, which is how a suspicion raised above gets
confirmed:

```{r}
#| label: vra-key-examples
#| eval: false

vra_key_examples(n = 3)                                   # any repeated group
vra_key_examples(only_col = "dt_partida_real", n = 3)     # re-filed?
vra_key_examples(only_col = "sg_icao_destino", n = 3)     # two real flights?
```

**Nothing de-duplicates yet.** The rule follows from what the tables above show, and it is
written once that is known — collapsing before then would delete a filled field to keep an
empty one, which is exactly the trap TOTALBR's `pk` collapse exists to avoid.

## Scheduled against realised — the day-shift to watch

`dt_referencia` dates a flight **by its departure** and carries no time of day. A night
flight therefore lands on the following date, which is normal: AAL0904 leaves SBGL on
01/12 at 23:51 and reaches KMIA on 02/12 at 08:16.

What is **not** normal, and was found in the first five rows sampled, is a realised time
almost a full day away from its scheduled one:

| | AAL0905, KMIA → SBGL |
| --- | --- |
| `dt_partida_prevista` | 01/12/2024 23:55 |
| `dt_partida_real` | **01/12/2024 00:55** |
| `dt_chegada_prevista` | 02/12/2024 09:25 |
| `dt_chegada_real` | **01/12/2024 09:07** |
| `ds_situacao_partida` | `Antecipado` |

The realised times are coherent *with each other* — 00:55 to 09:07 is an eight-hour flight —
so the flight is real. What looks wrong is the pairing: the realised pair appears to belong
to the previous day's instance of the same service, and ANAC's own verdict reads it as
23 hours "early". Four of the five rows sampled were clean, so this is not systematic.

```{r}
#| label: tbl-bra-vra-dayshift
#| tbl-cap: "Departures whose realised time is more than 6 h from the schedule"

d <- vra_read(cols = c("sg_empresa_icao", "nr_voo", "dt_partida_prevista",
                       "dt_partida_real", "ds_situacao_partida"),
              raw_dir = bra_vra_raw_dir)

if (nrow(d) == 0) {
  knitr::kable(tibble::tibble(note = "Nothing downloaded yet."))
} else {
  p   <- as.POSIXct(d$dt_partida_prevista, format = "%d/%m/%Y %H:%M", tz = "UTC")
  r   <- as.POSIXct(d$dt_partida_real,     format = "%d/%m/%Y %H:%M", tz = "UTC")
  gap <- as.numeric(difftime(r, p, units = "hours"))

  knitr::kable(
    tibble::tibble(SITUACAO = d$ds_situacao_partida, GAP_H = gap) |>
      dplyr::filter(!is.na(GAP_H), abs(GAP_H) > 6) |>
      dplyr::count(SITUACAO, sort = TRUE) |>
      dplyr::mutate(PCT_OF_ALL = round(100 * n / nrow(d), 3)))
}
```

Run it before using `dt_partida_real` for anything. If these rows are a handful, exclude and
note them; if they are a percent of the data, the delay figures ANAC publishes rest on them
too, and that is a question for ANAC rather than a filter for us.

## Open Points

- **The candidate key is measured but not adopted.** Nothing de-duplicates on it until the
  repeated rows have been looked at: a leg re-filed is two rows that are not
  interchangeable, and deleting the wrong one loses a filled field.
- **The day-shift rows are counted but not handled.** Whether to exclude them, and what that
  does to a punctuality figure, is open.
- **`nr_assentos_ofertados` makes VRA a capacity source too** — seats offered per flight,
  which none of the other datasets carry. Not used yet; worth remembering it is here.
- **The relationship to TOTALBR is unmeasured.** Both describe the same flights; VRA covers
  only *regular* flights, so it is a subset, and how large a subset is a number worth having
  before either is used as a denominator.
- **The situation bands are ANAC's definitions**, not ours. Before reporting them beside a
  European figure, confirm the thresholds match what the comparison assumes.
