# Brazil VRA (ANAC) Data Ingestion

```{r}
#| label: setup
#| include: false

source(here::here("_chapter-setup.R"))
source(here::here("VRA", "download_vra.R"))
```

This note documents the ingestion of **VRA** — *Voo Regular Ativo* — ANAC's record of the
regular flights operated to, from and within Brazil, for the Brazil–Europe
operational-efficiency study.

VRA is a **different kind of source** from the others in this repository, and the difference
matters more than the schema does. `dstaxi`, `kpi08` and TOTALBR are ATC observations: what
the air navigation system saw happen. VRA is **filed by the airlines** — the scheduled time,
the realised time, and ANAC's verdict on the difference. When VRA and TOTALBR disagree about
a flight, that is two measurements differing, not automatically an error in either.

It is also the only source here that answers punctuality **directly**. `ds_situacao_voo`,
`ds_situacao_partida` and `ds_situacao_chegada` are already computed and carry ANAC's own
banding (`Pontual`, `Antecipado`, `Atraso 30-60`, `Atraso 60-120`), with
`ds_justificativa` for the reason. Nothing has to be derived to report delay — which also
means the definitions are ANAC's, not ours, and must be quoted as such.

## The endpoint

Base: `https://sas.anac.gov.br/sas/vra_api/vra`

```
.../vra?dt_referencia1=01122024&dt_referencia2=31122024
```

Four properties of this API are encoded in `VRA/download_vra.R` because none of them is
guessable, and each was found by probing rather than assumed:

| Property | What it means for the download |
| --- | --- |
| The window is `dt_referencia1` / `dt_referencia2`, both **DDMMYYYY**, both **inclusive** | Neither ISO nor `d/m/Y`. A wrong format returns an empty window, not an error. `vra_window()` formats it so no caller does it by hand. |
| The answer is **JSON inside JSON** | The body is a JSON *string* wrapping the array of records: one decode yields a character vector of length one that looks like corruption. `vra_parse_json()` unwraps it. |
| **The wide window is honoured** | December 2024 returns 85,919 rows over 31 distinct days, 66 MB — 32× a single day. So a month is one request, not thirty. |
| There is **no unique key** | No `id`, no hash. Nothing is de-duplicated on download. |

::: {.callout-note}
## Why the wide window was measured rather than assumed
TATIC accepts `datai`/`dataf` spanning a month and silently returns **only the first day** —
no error, no warning. Discovering that after a year had been "downloaded" is expensive, so
the check is built into the downloader: every month is compared against the days it should
hold, and a month window that comes back with one day says so and names the fix.
:::

## Downloading

```{r}
#| label: download-bra-vra-data
#| eval: false

source(here::here("VRA", "download_vra.R"))

# The period comes from _chapter-setup.R, and R holds the value it read when the
# session started. Editing the file does not change a running session -- source
# it again, and check, before a download that takes hours goes after the wrong
# years.
source(here::here("_chapter-setup.R"))
vra_data_years                        # <- confirm this is the period you meant

download_vra(vra_data_years)
download_vra(2026)                    # or name the year outright
```

One file per year in `data-raw/vra/`, with the month parts under `parts/`.

**Each month is written the moment it arrives**, before the next is asked for, so a run
stopped halfway keeps every month it had. Re-running skips the months already complete, so
interrupting costs nothing but the month in flight — and completeness is judged by the days
a part holds, not by a request having returned, so a month that came back short is fetched
again rather than trusted.

**Completeness is a statement about days present, never about a request having returned.** A
month fetched while it was still open is short by design and is fetched again once it
closes. A month that came back truncated is short by fault. Neither is accepted as complete,
and the two are deliberately not distinguished at this stage — both mean "fetch it again".

## Coverage — the days each month actually holds

```{r}
#| label: tbl-bra-vra-coverage
#| tbl-cap: "Days and rows per month, as downloaded"

cov <- vra_coverage(bra_vra_raw_dir)
if (is.null(cov)) {
  knitr::kable(tibble::tibble(
    note = "Nothing downloaded yet — run the chunk above."))
} else {
  knitr::kable(cov)
}
```

It reads the month **parts**, not the year files. The year file is written only after the
last month of a year has been fetched, so a run interrupted halfway — the normal way a long
download ends — leaves months on disk and no year file at all.

A `SHORT` month is either still open or truncated. Read it together with today's date before
concluding which.

## Is there a key?

VRA carries no id, so any de-duplication needs a key we construct — and a key that looks
plausible is exactly what costs the most when it is wrong. TOTALBR's `id` is the precedent:
it reads like an identifier, repeats on every flight of the same route, and de-duplicating on
it would have deleted almost the entire dataset.

The natural candidate here is **operator + flight number + scheduled departure**. This
measures it instead of trusting it:

```{r}
#| label: tbl-bra-vra-key
#| tbl-cap: "How often the candidate key repeats"

d <- vra_read(cols = c("sg_empresa_icao", "nr_voo", "dt_partida_prevista",
                       "sg_icao_origem", "sg_icao_destino"),
              raw_dir = bra_vra_raw_dir)

if (nrow(d) == 0) {
  knitr::kable(tibble::tibble(note = "Nothing downloaded yet."))
} else {
  k <- paste(d$sg_empresa_icao, d$nr_voo, d$dt_partida_prevista, sep = "|")
  knitr::kable(tibble::tibble(
    ROWS     = nrow(d),
    DISTINCT = dplyr::n_distinct(k),
    REPEATED = sum(duplicated(k)),
    PCT      = round(100 * sum(duplicated(k)) / nrow(d), 2)))
}
```

A repeat rate near zero means the key holds. A repeat rate of a few percent needs the
repeated rows looked at before anything is deleted: a flight can legitimately appear twice
when a leg is re-filed, and the two rows are then not interchangeable.

## Scheduled against realised — the day-shift to watch

`dt_referencia` dates a flight **by its departure** and carries no time of day. A night
flight therefore lands on the following date, which is normal: AAL0904 leaves SBGL on
01/12 at 23:51 and reaches KMIA on 02/12 at 08:16.

What is **not** normal, and was found in the first five rows sampled, is a realised time
almost a full day away from its scheduled one:

| | AAL0905, KMIA → SBGL |
| --- | --- |
| `dt_partida_prevista` | 01/12/2024 23:55 |
| `dt_partida_real` | **01/12/2024 00:55** |
| `dt_chegada_prevista` | 02/12/2024 09:25 |
| `dt_chegada_real` | **01/12/2024 09:07** |
| `ds_situacao_partida` | `Antecipado` |

The realised times are coherent *with each other* — 00:55 to 09:07 is an eight-hour flight —
so the flight is real. What looks wrong is the pairing: the realised pair appears to belong
to the previous day's instance of the same service, and ANAC's own verdict reads it as
23 hours "early". Four of the five rows sampled were clean, so this is not systematic.

```{r}
#| label: tbl-bra-vra-dayshift
#| tbl-cap: "Departures whose realised time is more than 6 h from the schedule"

d <- vra_read(cols = c("sg_empresa_icao", "nr_voo", "dt_partida_prevista",
                       "dt_partida_real", "ds_situacao_partida"),
              raw_dir = bra_vra_raw_dir)

if (nrow(d) == 0) {
  knitr::kable(tibble::tibble(note = "Nothing downloaded yet."))
} else {
  p   <- as.POSIXct(d$dt_partida_prevista, format = "%d/%m/%Y %H:%M", tz = "UTC")
  r   <- as.POSIXct(d$dt_partida_real,     format = "%d/%m/%Y %H:%M", tz = "UTC")
  gap <- as.numeric(difftime(r, p, units = "hours"))

  knitr::kable(
    tibble::tibble(SITUACAO = d$ds_situacao_partida, GAP_H = gap) |>
      dplyr::filter(!is.na(GAP_H), abs(GAP_H) > 6) |>
      dplyr::count(SITUACAO, sort = TRUE) |>
      dplyr::mutate(PCT_OF_ALL = round(100 * n / nrow(d), 3)))
}
```

Run it before using `dt_partida_real` for anything. If these rows are a handful, exclude and
note them; if they are a percent of the data, the delay figures ANAC publishes rest on them
too, and that is a question for ANAC rather than a filter for us.

## Open Points

- **The candidate key is measured but not adopted.** Nothing de-duplicates on it until the
  repeated rows have been looked at: a leg re-filed is two rows that are not
  interchangeable, and deleting the wrong one loses a filled field.
- **The day-shift rows are counted but not handled.** Whether to exclude them, and what that
  does to a punctuality figure, is open.
- **`nr_assentos_ofertados` makes VRA a capacity source too** — seats offered per flight,
  which none of the other datasets carry. Not used yet; worth remembering it is here.
- **The relationship to TOTALBR is unmeasured.** Both describe the same flights; VRA covers
  only *regular* flights, so it is a subset, and how large a subset is a number worth having
  before either is used as a denominator.
- **The situation bands are ANAC's definitions**, not ours. Before reporting them beside a
  European figure, confirm the thresholds match what the comparison assumes.
